---
title: 'Linear Regression with Categorical Variables'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Show all script by default
  message = FALSE,    # hide messages
  warning = FALSE,    # hide package warnings
  fig.width = 5,
  fig.height = 3.5,
  fig.align = "center",
  fig.show = "hold",
  results = "hold"
)
```

## The Promise From Last Chapter

The last chapter said the outcome has to be quantitative but the predictors are far less picky, and that we would start feeding regression categorical variables on purpose.

This is that chapter.

The question is genuinely strange the first time you meet it. Regression multiplies a predictor by a slope, and you cannot multiply "Texting" by 0.34. So how does a model built entirely out of arithmetic handle a variable whose values are words?

The answer is that we translate the words into numbers, in a way specific enough that the slopes still mean something afterward. Do that and something quietly excellent falls out: the *t*-test you learned in Part 1 turns out to have been a regression the whole time, wearing a disguise.

## Distracted Driving

Driving requires constant attention. When something unexpected happens, such as a car braking suddenly, you must react quickly. Psychologists studying attention have shown that distractions slow reaction time. One of the most dangerous distractions is texting while driving.

The data in this chapter are simulated. Do not reproduce the study by handing undergraduates phones, placing them in actual traffic, and chasing the car with a bat to improve ecological validity seen in city style road rage. Use a simulator and an ethics-approved protocol. The Institutional Review Board enjoys fewer surprises than the Probability Gods.

In this example we examine **braking reaction time** (seconds) under different driving conditions.

Reaction time is measured in seconds. Larger values indicate slower reactions (which is bad).

### Dummy Coding

When a predictor is categorical, regression cannot use the labels directly.  
Regression requires numbers, so the categories must be translated into numeric variables. 

This process is called **dummy coding**. *Who are you calling a dummy? The coding scheme not the coder.* It is the most common coding scheme, and the **default used by R** whenever it detects a factor.

### Dummy Coding Details

Dummy coding compares each group to a **reference group**.

The reference group serves as the **baseline** or **zero point** for the model.  
All other groups are interpreted relative to that baseline.

In other words, the model asks: *How much does each group deviate from the reference group?*


## Two-Level Example

Suppose we conduct a study comparing two driving conditions: **No distraction** vs **Texting while driving**

We collect data for: $n = 30$ drivers per group (*dataframe = dat2*)  

```{r example2, echo=FALSE}
set.seed(343)

n <- 30
sd_rt <- 0.15

NoDistraction <- rnorm(n, .70, sd_rt)
Texting <- rnorm(n, 1.00, sd_rt)

Condition <- c(rep("No Distraction", n),
               rep("Texting", n))

RT <- c(NoDistraction, Texting)

dat2 <- data.frame(Condition, RT)
```


We have two driving conditions (and predict):

| Condition | Reaction Time Prediction |
|-----------|---------------|
| No distraction | faster |
| Texting | slower |

To represent this in regression, we create a dummy variable.

| Variable | C1 |
|---------|---|
| No Distraction | 0 |
| Texting | 1 |

Here:

- **0 represents the reference group**
- **1 represents the comparison group**

So the dummy variable indicates whether a driver is texting. Note: We usually put the 1 on our experimental group. 

### How This Connects to Regression

The regression model becomes

$$
RT = b_0 + b_1C_1
$$

Interpretation:

- $b_0$ = the mean reaction time for the **reference group**
- $b_1$ = how much the **Texting** group differs from that reference group

### Predicted Values

For **No Distraction**:

$C_1 = 0$

$$
RT = b_0 + b_10
$$

Which simplifies to

$$
RT = b_0
$$

- the **intercept (b_0) equals the mean reaction time for No Distraction**

For **Texting**:

$C_1 = 1$

$$
RT = b_0 + b_11
$$

Which simplifies to
$$
RT = b_0 + b_1
$$

- the **slope ($b_1$) coefficient represents the difference between Texting and No Distraction**


So the regression coefficients directly reproduce the group means.

### Important

In R, you usually **do not need to create dummy variables yourself**.

When a predictor is stored as a **factor**, R automatically creates the dummy-coded variables internally when fitting the model.

The **first level of the factor becomes the reference group**.

For example:

```{r set-factor-levels, echo=TRUE}
dat2$Condition <- factor(dat2$Condition,
                         levels = c("No Distraction", "Texting"))
```

Because `"No Distraction"` appears first, it becomes the **reference group**.

When we get our regression output from R you see it as:

- the **intercept equals the mean reaction time for No Distraction**
- the **ConditionTexting coefficient represents the difference between Texting and No Distraction**


### Comparing the Groups with a *t*-test

We first test the hypothesis that texting increases reaction time.

First we load the packages needed for this example.

- **tidyverse**: used for data manipulation and summarizing the data  
- **apa**: used to produce APA-style statistical output

```{r load-packages, echo=TRUE}
library(apa)
library(tidyverse)
```

### Calculate Group Means and Standard Deviations

Before running a statistical test, it is always good practice to examine the descriptive statistics.

The code below calculates the **mean (M)** and **standard deviation (SD)** of reaction time for each condition.

```{r group-descriptives, echo=TRUE}
dat2 |>
  group_by(Condition) |>
  summarise(
    M = mean(RT),
    SD = sd(RT)
  )
```

This gives us a quick summary of how reaction time differs across the two driving conditions.


### Run the Independent Samples *t*-test

Next we test whether the difference between groups is statistically significant.

```{r run-ttest, echo=TRUE}
texting.ttest <- t.test(RT ~ Condition,
                        data = dat2,
                        var.equal = TRUE)
```

Reminder:

- `RT ~ Condition` tells R we are predicting **reaction time** from the **driving condition**
- `data = dat2` specifies which dataset to use
- `var.equal = TRUE` assumes equal variances across groups (the standard Student *t*-test)


### Print the Results in APA Format

Finally, we format the results in **APA style** using the `apa` package.

```{r ttest-apa, echo=TRUE}
apa(texting.ttest, format="rmarkdown") 
#Note: format can also be "text", "markdown", "rmarkdown", "html", "latex" or "docx"
#I have to use rmarkdown cause of how I created this chapter. You might use use "text"
```

### APA write up

```{r ttest-apa-values, echo=FALSE}
desc <- dat2 |>
  dplyr::group_by(Condition) |>
  dplyr::summarise(M = mean(RT), SD = sd(RT), n = dplyr::n())

sp <- sqrt(((desc$n[1]-1)*desc$SD[1]^2 + (desc$n[2]-1)*desc$SD[2]^2) / (sum(desc$n)-2))
d.texting <- (desc$M[1] - desc$M[2]) / sp

# APA drops the leading zero on statistics bounded by 1
apa.dec <- function(x, k = 2) sub("^(-?)0", "\\1", sprintf(paste0("%.", k, "f"), x))
```

Drivers who texted while driving had significantly slower reaction times ($M =$ `r round(desc$M[2], 2)`, $SD =$ `r round(desc$SD[2], 2)`) than drivers with no distraction ($M =$ `r round(desc$M[1], 2)`, $SD =$ `r round(desc$SD[1], 2)`), $t(`r texting.ttest$parameter`) =$ `r round(texting.ttest$statistic, 2)`, $p < .001$, $d =$ `r round(d.texting, 2)`, indicating that texting substantially increased reaction time.

**Small note**: the negative sign on $t$ simply reflects the order of subtraction (No Distraction minus Texting).

::: {.callout-warning title="A d of 2.6 Should Be Making You Suspicious"}
By now you have read the power chapter, so that effect size should raise an eyebrow. Real texting-while-driving effects are large, which is exactly why it is illegal, but they are not science-fiction large. Published estimates sit closer to $d$ of 0.5 to 1.

I built this simulation to make the pictures come out clean. Simulated data lies politely, and it will happily calibrate your intuitions to a world that does not exist.
:::

### The Same Analysis Using Regression

Now we run the equivalent regression model.

```{r two-level-model, echo=TRUE}
model2 <- lm(RT ~ Condition, data = dat2)
summary(model2)
```

### Why Are the *t*-values the Same?

Stop and compare two numbers before going any further, because this is the whole point of the section.

In the `ConditionTexting` row above: $t =$ `r round(summary(model2)$coefficients[2,3], 2)`, $p < .001$.

Back in the *t*-test: $t =$ `r round(texting.ttest$statistic, 2)`, $p < .001$.

Same magnitude, opposite sign. Not similar. *Identical.*

The sign flip is bookkeeping: `t.test()` subtracted in the order No Distraction minus Texting, and the regression reports Texting relative to No Distraction. Everything else matches because the two procedures are doing the same arithmetic:

- The regression slope **is** the difference between the group means.
- Its standard error **is** the pooled standard error of that difference.
- Dividing one by the other is precisely what the *t*-test did.

One condition on that equivalence, and it is worth naming. It holds because we ran the *t*-test with `var.equal = TRUE`. Ordinary regression assumes one common residual variance across groups, which is the same assumption the pooled Student *t*-test makes. Ask for Welch's version instead, which does not assume that, and the match breaks. This is the other side of the Welch discussion from the independent-samples chapter.

The regression equation is

$$
RT = b_0 + b_1X
$$

Interpretation:

- $b_0$ = mean reaction time for **No Distraction**
- $b_1$ = difference between **Texting** and **No Distraction**

Because the dummy variable for No Distraction is 0:

For No Distraction

$$
RT = b_0
$$

For Texting

$$
RT = b_0 + b_1
$$

So the intercept is simply the mean reaction time for the reference group.

```{r two-level-coefs, echo=TRUE}
coef(model2)
```

Which gives us:

| Coefficient | Value |
|--------------|------|
| Intercept | `r round(coef(model2)[1], 3)` |
| ConditionTexting | `r round(coef(model2)[2], 3)` |

Now we can reconstruct the means.

### No Distraction

$$
RT = b_0
$$

$$
RT = `r round(coef(model2)[1], 3)`
$$

So the predicted mean reaction time is **`r round(coef(model2)[1], 3)` seconds**.

### Texting

$$
RT = b_0 + b_1
$$

$$
RT = `r round(coef(model2)[1], 3)` + `r round(coef(model2)[2], 3)`
$$

$$
RT = `r round(coef(model2)[1] + coef(model2)[2], 3)`
$$

So the predicted mean reaction time is **`r round(coef(model2)[1] + coef(model2)[2], 3)` seconds** for texting while driving.

These values match the observed group means from the data.

**Regression coefficients are simply another way of expressing the group means and their differences.**


### Extracting the Means with `emmeans`

We can also extract the model-estimated means so we do not have to reconstruct them manually from the regression equation.

```{r emmeans-two-level, echo=TRUE}
library(emmeans)

emmeans(model2, ~ Condition)
```

### What the Code Does

`library(emmeans)` loads the **emmeans** package. This package is designed to extract and compare means from statistical models such as regression.

`emmeans(model2, ~ Condition)` tells R:

- use the regression model **model2**
- estimate the means for each level of **Condition**

The `~ Condition` part means:  
*compute the means for the variable Condition.*


### What "Estimated Marginal Means" Means

The name **emmeans** stands for **Estimated Marginal Means**.

Estimated marginal means are the **means predicted by the statistical model**.

Why *marginal*? Before computers did all this instantly, people wrote the means in the margins of tables using actual pencils. They were Number 2 pencils, naturally---the same pencils you never remember to bring to my exams. iPads are new. Most of the old people who write R packages or statistics books were born in the 1900s and are older than the internet.

In this simple example, they correspond directly to the group means:

| Condition | Predicted Mean |
|-----------|---------------|
| No Distraction | $b_0$ |
| Texting | $b_0 + b_1$ |

So `emmeans()` is simply doing the same math we just performed manually:

- extracting the intercept
- adding the appropriate regression coefficients
- reporting the predicted group means


### Why This Is Useful

For simple models like this one, calculating the means by hand is straightforward.

However, as models become more complex (for example, with multiple predictors or interactions), manually reconstructing the means becomes tedious.

`emmeans` provides a convenient way to:

- extract the model-based means
- compare groups
- obtain confidence intervals
- perform additional tests

All while keeping the results tied directly to the regression model.


### Testing the Difference with `emmeans`

We can also test the difference between groups directly from the regression model using `emmeans`.

```{r pairs-two-level, echo=TRUE}
pairs(emmeans(model2, ~ Condition))
```

### What the Code Does

The function `emmeans(model2, ~ Condition)` first computes the **estimated marginal means** for each level of the factor **Condition** based on the regression model.

The function `pairs()` then compares those means to one another.

In this case, there are only two conditions, so the function tests the difference between:

- **Texting**
- **No Distraction**


### How This Connects to the Regression Model

Recall that the regression equation is

$$
RT = b_0 + b_1X
$$

where

- $b_0$ = mean reaction time for **No Distraction**
- $b_1$ = difference between **Texting** and **No Distraction**

The comparison performed by `pairs()` is therefore testing whether:

$$
b_1 = 0
$$

If $b_1$ is significantly different from zero, the two groups have different mean reaction times.


### Comparison to *t*-test

Even though the output may look similar to a *t*-test, the comparison is now being computed **within the regression framework**.

The means and the group comparison are both derived directly from the fitted regression model.


### APA Regression Write up

Drivers who texted while driving had significantly slower reaction times than drivers with no distraction. Relative to this reference group (no distractions), the Texting condition increased reaction time by `r round(coef(model2)[2], 2)` seconds, $b =$ `r round(coef(model2)[2], 2)`, $SE =$ `r round(summary(model2)$coefficients[2,2], 2)`, $t(`r df.residual(model2)`) =$ `r round(summary(model2)$coefficients[2,3], 2)`, $p < .001$. The model explained a substantial proportion of variance in reaction time, $R^2 =$ `r apa.dec(summary(model2)$r.squared)`, $F(1, `r df.residual(model2)`) =$ `r round(summary(model2)$fstatistic[1], 2)`, $p < .001$, indicating that differences in driving condition accounted for about `r round(summary(model2)$r.squared * 100)`% of the variability in reaction times across drivers.


### Plotting the Model-Estimated Means

We can also visualize the model-estimated means returned by `emmeans`.

First we store the estimated means in an object.

```{r store-means, echo=TRUE}
means <- emmeans(model2, ~ Condition)

means
```

To make plotting easier, we convert the results into a data frame.

```{r means-to-df, echo=TRUE}
means_df <- as.data.frame(means)

means_df
```

The data frame contains:

- `emmean` = the estimated mean
- `lower.CL` = lower confidence interval
- `upper.CL` = upper confidence interval

Now we can quickly plot these values using **ggplot**.

```{r two-level-means-plot, echo=TRUE}
#| fig-cap: "Model-estimated mean reaction time in each condition, with 95% confidence intervals. The bars come straight out of the regression."
#| fig-alt: "Bar chart with two bars, No Distraction at about 0.66 seconds and Texting at about 1.01 seconds, each with a narrow error bar. The Texting bar is clearly taller and the intervals do not overlap."
library(ggplot2)

ggplot(means_df, aes(x = Condition, y = emmean)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .1) +
  ylab("Reaction Time (seconds)") +
  xlab("Driving Condition")+theme_bw()
```

This plot shows the **model-estimated means and their 95% confidence intervals**.

Because the means come from `emmeans`, the figure directly represents the regression model we fit earlier.


## Three-Level Example

Now suppose we repeat the study but add a third condition:

**Listening to a statistics lecture while driving to class.**


```{r simulate-three-level, echo=FALSE}
set.seed(543)

Lecture <- rnorm(n, .85, sd_rt)

Condition <- c(rep("No Distraction", n),
               rep("Texting", n),
               rep("Lecture", n))

RT <- c(NoDistraction, Texting, Lecture)

dat3 <- data.frame(Condition, RT)
```


### Factoring the Three-Level Predictor

Again we convert the predictor to a factor.

```{r factor-three-level, echo=TRUE}
dat3$Condition <- factor(dat3$Condition,
                         levels = c("No Distraction",
                                    "Texting",
                                    "Lecture"))

levels(dat3$Condition)
```

The first level becomes the reference group.

So again:

```
NoDistraction = reference (0)
```

R now creates **two dummy variables** internally.

Conceptually this looks like:

| Condition | Texting | Lecture |
|-----------|--------|--------|
| No distraction | 0 | 0 |
| Texting | 1 | 0 |
| Lecture | 0 | 1 |


### Multiple *t*-tests Problem

If we used *t*-tests, we would need three comparisons:

1. No distraction vs texting  
2. No distraction vs lecture  
3. Lecture vs texting  

Running multiple tests increases the probability of a **Type I error**.

If $\alpha = .05$, then the probability of at least one false positive ($1-(1-\alpha)^j$) [$j$ = # of *t*-tests]: 

$$
1 - (1 - .05)^3
$$

```{r type-one-inflation, echo=TRUE}
1 - (1 - .05)^3
```

This is about **14%**, much larger than the intended 5%.

Regression helps here, but it is worth being precise about how, because the honest version is less magical than it sounds. Fitting one model does not delete the multiple-testing problem. What it does is organize every comparison under a single model and hand you a **gatekeeper**: one overall test of whether the group means differ at all, which you check *before* going looking for which pairs differ. That gate is the omnibus $F$, and we are about to meet it.


### Regression with Three Groups

Regression allows us to test all comparisons within a single model.

```{r three-level-model, echo=TRUE}
model3 <- lm(RT ~ Condition, data = dat3)
summary(model3)
```

### The Gate: the Omnibus F

Look at the very last line of that output, the one everybody scrolls past:

`F-statistic: `r round(summary(model3)$fstatistic[1], 2)` on `r summary(model3)$fstatistic[2]` and `r summary(model3)$fstatistic[3]` DF`

That is the gatekeeper. It tests one null hypothesis: **all three group means are equal**. Not "texting differs from no distraction," not any particular pair, just whether there is any difference anywhere in the set.

Ours is $F(`r summary(model3)$fstatistic[2]`, `r summary(model3)$fstatistic[3]`) =$ `r round(summary(model3)$fstatistic[1], 2)`, $p < .001$, so the gate is open. Something in there differs, and we have earned the right to go find out what.

Had that $F$ come back non-significant, the honest move is to stop. The means are not distinguishable, and hunting through pairs anyway is precisely the 14% problem we just calculated, now with extra steps.

The model is

$$
RT = b_0 + b_1(\text{Texting}) + b_2(\text{Lecture})
$$

Interpretation:

- $b_0$ = mean reaction time for **No Distraction**
- $b_1$ = difference between **Texting** and **No Distraction**
- $b_2$ = difference between **Lecture** and **No Distraction**


### Reconstructing the Group Means from the Regression Model

Recall that the regression model is

$$
RT = b_0 + b_1(\text{Texting}) + b_2(\text{Lecture})
$$

From the model output we obtained the following coefficients:

| Coefficient | Value |
|-------------|------|
| $b_0$ (Intercept) | `r round(coef(model3)[1], 3)` |
| $b_1$ (Texting) | `r round(coef(model3)[2], 3)` |
| $b_2$ (Lecture) | `r round(coef(model3)[3], 3)` |

Again, the intercept represents the **mean reaction time for the reference group**, which is **No Distraction**.

```{r three-level-coefs, echo=TRUE}
coef(model3)
```


### No Distraction

For the reference group, both dummy variables equal 0.

$$
RT = b_0
$$

Substituting the coefficient:

$$
RT = `r round(coef(model3)[1], 3)`
$$

So the predicted mean reaction time for **No Distraction** is **`r round(coef(model3)[1], 3)` seconds**.


### Texting

For the texting group:

Texting = 1  
Lecture = 0

$$
RT = b_0 + b_11 + b_20 
$$

Substituting the coefficients:

$$
RT = `r round(coef(model3)[1], 3)` + `r round(coef(model3)[2], 3)`
$$

$$
RT = `r round(coef(model3)[1] + coef(model3)[2], 3)`
$$

So the predicted mean reaction time for **Texting** is **`r round(coef(model3)[1] + coef(model3)[2], 3)` seconds**.


### Lecture

For the lecture group:

Texting = 0  
Lecture = 1

$$
RT = b_0 + b_10 + b_21 
$$

Substituting the coefficients:

$$
RT = `r round(coef(model3)[1], 3)` + `r round(coef(model3)[3], 3)`
$$

$$
RT = `r round(coef(model3)[1] + coef(model3)[3], 3)`
$$

So the predicted mean reaction time for **Lecture** is **`r round(coef(model3)[1] + coef(model3)[3], 3)` seconds**.


### Prediction summarized

The regression coefficients allow us to **reconstruct the group means directly**.

| Condition | Model Equation | Predicted Mean |
|-----------|---------------|---------------|
| No Distraction | $b_0$ | `r round(coef(model3)[1], 3)` |
| Texting | $b_0 + b_1$ | `r round(coef(model3)[1] + coef(model3)[2], 3)` |
| Lecture | $b_0 + b_2$ | `r round(coef(model3)[1] + coef(model3)[3], 3)` |

**Regression is simply expressing the group means and their differences using an equation.**

### Extracting the Model-Estimated Means

Just like before, `emmeans` will do all of that reconstruction for us.

```{r three-level-emmeans, echo=TRUE}
means3 <- emmeans(model3, ~ Condition)
means3
```

These values represent the **means predicted by the regression model**, and they match the numbers we just assembled by hand.

### Changing the Reference Group

The reference group is a choice, not a fact about the world. `factor(levels = )` sets it when you build the variable. To switch it afterwards, use `relevel()`:

```{r relevel-demo, echo=TRUE}
dat3$Condition2 <- relevel(dat3$Condition, ref = "Texting")

model3b <- lm(RT ~ Condition2, data = dat3)
coef(model3b)
```

Every coefficient is now expressed relative to Texting. The intercept is Texting's mean, `r round(coef(model3b)[1], 3)`, and the other two rows are how much *slower or faster* the remaining conditions are compared to it. No Distraction now carries a negative coefficient, `r round(coef(model3b)[2], 3)`, because being undistracted is faster than texting.

Here is the part worth watching, though. Nothing about the model actually moved:

```{r relevel-unchanged, echo=TRUE}
# Same variance explained, same omnibus test, same fitted means
c(R2.original = summary(model3)$r.squared,
  R2.releveled = summary(model3b)$r.squared)

c(F.original = summary(model3)$fstatistic[1],
  F.releveled = summary(model3b)$fstatistic[1])
```

Identical. Releveling changed which comparisons the coefficients describe. It did not change the fit, the group means, the residuals, or the drivers.


### APA Write-Up

```{r three-level-apa-values, echo=FALSE}
s3   <- summary(model3)
em3  <- as.data.frame(emmeans(model3, ~ Condition))
ref  <- em3[em3$Condition == "No Distraction", ]
```

A regression model was used to predict reaction time from driving condition. The overall model was significant, $R^2 =$ `r apa.dec(s3$r.squared)`, $F(`r s3$fstatistic[2]`, `r s3$fstatistic[3]`) =$ `r round(s3$fstatistic[1], 2)`, $p < .001$, indicating that driving condition accounted for about `r round(s3$r.squared * 100)`% of the variability in reaction times.

Because **No Distraction** was entered as the first factor level, it served as the **reference group**, and the intercept represented the mean reaction time for that condition (*M* = `r round(ref$emmean, 2)`, *SE* = `r round(ref$SE, 2)`, 95% CI [`r round(ref$lower.CL, 2)`, `r round(ref$upper.CL, 2)`]).

Relative to this reference group, texting significantly increased reaction time, $b =$ `r round(coef(model3)[2], 2)`, $SE =$ `r round(s3$coefficients[2,2], 2)`, $t(`r df.residual(model3)`) =$ `r round(s3$coefficients[2,3], 2)`, $p < .001$. Listening to a statistics lecture also increased reaction time relative to the no-distraction condition, $b =$ `r round(coef(model3)[3], 2)`, $SE =$ `r round(s3$coefficients[3,2], 2)`, $t(`r df.residual(model3)`) =$ `r round(s3$coefficients[3,3], 2)`, $p < .001$.


### Testing Differences Between Groups

We can also test the differences between all pairs of conditions.

```{r pairwise-comparisons, echo=TRUE}
pairs(means3, adjust="none")
```

This compares the estimated means for:

- No Distraction vs Texting
- No Distraction vs Lecture
- Lecture vs Texting

All comparisons are derived directly from the regression model. This contains the comparison we did not test earlier (texting vs. lecture). You can add that to your APA report as well.

### About That `adjust="none"`

You should be suspicious of that argument, because we spent a whole section calculating that three unadjusted tests inflate the error rate to 14%. So why are we allowed to run three unadjusted tests?

Because we checked the gate first. The omnibus $F$ was significant, which is what earns the right to look at individual pairs without adjusting each one. This is **Fisher's protected LSD**, and the word doing the work is *protected*: the $F$ is the protection, and it was bought before any pair was examined.

The protection has a hard expiry date, and it is worth memorizing, because it is a favorite exam question:

::: {.callout-important title="The Protection Expires at Three Groups"}
With exactly **three** groups, a significant omnibus $F$ followed by unadjusted pairwise tests keeps the familywise error rate at $\alpha$. This is a genuine mathematical result, not a convention.

With **four or more** groups it stops working, and the error rate creeps back up. At that point you need a real correction (Tukey, Holm, Bonferroni), which is what the [Multiple Comparisons](Ch_20_Multiple_Comparisons.qmd) chapter is for.

So `adjust = "none"` here is a defensible choice with three groups and a significant $F$. Pasted into a four-group analysis, it is just the 14% problem again.
:::

### APA for follow up test

```{r followup-values, echo=FALSE}
pw <- as.data.frame(pairs(means3, adjust = "none"))
tl <- pw[pw$contrast == "Texting - Lecture", ]
```

In addition, pairwise comparisons of the estimated marginal means showed that texting produced significantly slower reaction times than listening to a statistics lecture, $b =$ `r round(tl$estimate, 2)`, $SE =$ `r round(tl$SE, 2)`, $t(`r tl$df`) =$ `r round(tl$t.ratio, 2)`, $p < .001$. Thus, although both distractions impaired reaction time relative to the no-distraction condition, texting was associated with the greatest slowing of reactions.


### Plotting the Model-Estimated Means

We can also visualize the results using a simple plot.

First we convert the estimated means to a data frame.

```{r three-level-means-df, echo=TRUE}
means3_df <- as.data.frame(means3)
means3_df
```

Now we plot the estimated means and their 95% confidence intervals.

```{r three-level-means-plot, echo=TRUE}
#| fig-cap: "All three conditions, model-estimated means with 95% confidence intervals. Lecture sits between the other two."
#| fig-alt: "Bar chart with three bars: No Distraction at about 0.66 seconds, Texting at about 1.01, and Lecture at about 0.86, each with a narrow error bar. Texting is tallest and No Distraction shortest."
library(ggplot2)

ggplot(means3_df, aes(x = Condition, y = emmean)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .1) +
  ylab("Reaction Time (seconds)") +
  xlab("Driving Condition")+theme_bw()
```

This figure shows the **model-estimated mean reaction time for each driving condition** along with the 95% confidence intervals.

## The Short Story

- R represents a categorical predictor with numeric contrast variables.
- Under ordinary treatment/dummy coding, the intercept is the reference-group mean.
- Each category coefficient is that group's difference from the reference group.
- With two groups, the regression test is equivalent to an independent-samples *t*-test under the matching variance assumptions.
- With three or more groups, the overall model asks whether the group means are all equal; follow-ups ask which contrasts differ.
- Changing the reference group changes the coefficient labels, not the fitted group means or the underlying data.
- Set factor levels deliberately. Otherwise R may select a reference group alphabetically, because the alphabet has apparently been granted authority over your scientific question.

::: {.callout-note title="Coding Changes the Coefficients, Not the People"}
Changing the reference group changes which comparison appears in each coefficient. It does not change the fitted group means, residuals, $R^2$, or the participants—who remain blissfully unaware that R renamed the argument.
:::
