---
title: "Mixed Designs: From Two Measurements to Change Over Time"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5.5,
  fig.height = 3.6,
  fig.align = "center",
  results = "hold"
)
```

## One Within-Subjects Predictor with Two Levels

A paired *t*-test compares two measurements from the same people. A mixed model can do the same job. The advantage is that it does not stop working when the design grows.

The [paired-samples *t*-test chapter](Ch_09_Paired_t.qmd) gave 48 students one lecture with chocolate and one without, then measured how long their eyes stayed on the screen. Here is that study again, with the same seed and the same students.

```{r simulate-chocolate-data}
set.seed(343)

n <- 48
participant_effect <- rnorm(n, mean = 0, sd = 1.20)

paired_wide <- data.frame(
  participant  = factor(seq_len(n)),
  no_chocolate = 5.00 + participant_effect + rnorm(n, 0, 0.80),
  chocolate    = 5.75 + participant_effect + rnorm(n, 0, 0.80)
)

paired_wide$difference <-
  paired_wide$chocolate - paired_wide$no_chocolate

paired_long <- tidyr::pivot_longer(
  paired_wide,
  cols      = c(no_chocolate, chocolate),
  names_to  = "condition",
  values_to = "fixation"
)

paired_long$condition <- factor(
  paired_long$condition,
  levels = c("no_chocolate", "chocolate"),
  labels = c("No chocolate", "Chocolate")
)
```

First, analyze the two measurements with the test you already know:

```{r paired-t-model}
paired_result <- t.test(
  paired_wide$chocolate,
  paired_wide$no_chocolate,
  paired = TRUE
)

paired_result
```

Now give the mixed model all 96 rows and tell it which rows came from the same student:

```{r paired-mixed-model}
library(lme4)
library(lmerTest)

paired_model <- lmer(
  fixation ~ condition + (1 | participant),
  data = paired_long
)

summary(paired_model)
```

```{r paired-model-comparison}
paired_fixed <- summary(paired_model)$coefficients

paired_comparison <- data.frame(
  method = c("Paired t-test", "Mixed model"),
  t = c(
    unname(paired_result$statistic),
    paired_fixed["conditionChocolate", "t value"]
  ),
  df = c(
    unname(paired_result$parameter),
    paired_fixed["conditionChocolate", "df"]
  ),
  p = c(
    paired_result$p.value,
    paired_fixed["conditionChocolate", "Pr(>|t|)"]
  ),
  estimate = c(
    unname(paired_result$estimate),
    paired_fixed["conditionChocolate", "Estimate"]
  )
)

print(paired_comparison, digits = 6, row.names = FALSE)
```

Those are not two answers that happen to agree. They are one answer, printed twice.

The fixed effect for `conditionChocolate` is the average chocolate effect. The random intercept, `(1 | participant)`, gives each student their own baseline level of staring. That is what **each person is their own control** looks like in a model formula.

With two measurements, the paired *t*-test is simpler. Keep using it. The mixed model matters because most repeated-measures studies refuse to remain this polite.

## The Mixed 2-by-2: Week 0 vs. Week 6 by Treatment Group

Now make one small change to the 2-by-2 from the [categorical-by-categorical interaction chapter](Ch_17_Categorical_Interaction.qmd). That chapter had two categorical predictors and four cells. So does this design.

Suppose researchers compare cognitive behavioral therapy (**Therapy**) with a control condition. Depression symptoms are measured at Week 0 and Week 6.

- **Occasion** is within subjects. Each person appears at Week 0 and Week 6.
- **Treatment** is between subjects. Each person belongs to Control or Therapy, never both.
- **Symptoms** is the quantitative outcome.

Here are the same four cells you saw before:

|  | **Control** | **Therapy** |
|---|---|---|
| **Week 0** | $\mu_{11}$: Control at baseline | $\mu_{12}$: Therapy at baseline |
| **Week 6** | $\mu_{21}$: Control at Week 6 | $\mu_{22}$: Therapy at Week 6 |

The interaction asks whether the Week 0 to Week 6 change is the same in both groups:

$$B_3 = (\mu_{22}-\mu_{12})-(\mu_{21}-\mu_{11})$$

That is the same **difference of differences** from the prior chapter. First find the change in Therapy. Then find the change in Control. Finally, compare those two changes.

### What Stays the Same, and the One Thing That Changes

| Feature | Earlier categorical 2-by-2 | This mixed 2-by-2 |
|---|---|---|
| Number of categorical predictors | Two | Two |
| Number of cells | Four | Four |
| Interaction | Difference of differences | Difference of differences |
| Who contributes to the cells | Different people in every cell | The same person contributes to both occasions |
| Model | `lm(Y ~ A * B)` | `lmer(Y ~ A * B + (1 | Person))` |

The fixed part of the formula barely changes. We add `(1 | Person)` because the two rows from one person are related. Pretending they came from strangers would count one human twice and let the model become confident in ways it has not earned.

### The Data

We will simulate four visits, but use only Week 0 and Week 6 at first. Weeks 2 and 4 return in the Advanced Topics section, after the 2-by-2 is secure.

```{r simulate-treatment-data}
set.seed(346)

n_person <- 96
times <- c(0, 2, 4, 6)

People <- data.frame(
  Person = factor(seq_len(n_person)),
  Treatment = factor(
    rep(c("Control", "Therapy"), each = n_person / 2),
    levels = c("Control", "Therapy")
  ),
  person_intercept = rnorm(n_person, 0, 6),
  person_slope = rnorm(n_person, 0, 0.80)
)

LongData <- expand.grid(
  Person = People$Person,
  Time = times
)

person_row <- match(LongData$Person, People$Person)
LongData$Treatment <- People$Treatment[person_row]

LongData$Symptoms <-
  55 -
  0.35 * LongData$Time -
  0.90 * LongData$Time * (LongData$Treatment == "Therapy") +
  People$person_intercept[person_row] +
  People$person_slope[person_row] * LongData$Time +
  rnorm(nrow(LongData), 0, 4)

EndpointData <- LongData[LongData$Time %in% c(0, 6), ]
EndpointData$Occasion <- factor(
  EndpointData$Time,
  levels = c(0, 6),
  labels = c("Week 0", "Week 6")
)
```

### Look at the Four Cells First

```{r endpoint-cell-means}
aggregate(
  Symptoms ~ Occasion + Treatment,
  data = EndpointData,
  FUN = function(x) c(M = mean(x), SD = sd(x))
)
```

```{r endpoint-plot, echo=FALSE}
#| fig-cap: "The mixed 2-by-2 before the model. Faint lines connect each person's Week 0 and Week 6 scores. Heavy lines connect the group means."
#| fig-alt: "A line graph of depression symptoms at Week 0 and Week 6 for Control and Therapy. Ninety-six faint lines connect each participant's two scores. The two heavy group-mean lines begin near the same level at Week 0. The Control mean falls slightly by Week 6, while the Therapy mean falls much more."

library(ggplot2)

EndpointMeans <- aggregate(
  Symptoms ~ Occasion + Treatment,
  data = EndpointData,
  FUN = mean
)

ggplot(
  EndpointData,
  aes(x = Occasion, y = Symptoms, group = Person, color = Treatment)
) +
  geom_line(alpha = 0.18, linewidth = 0.45) +
  geom_line(
    data = EndpointMeans,
    aes(group = Treatment),
    linewidth = 1.4
  ) +
  geom_point(
    data = EndpointMeans,
    aes(group = Treatment),
    size = 3
  ) +
  scale_color_manual(values = c("Control" = "#d97757", "Therapy" = "#287c8e")) +
  labs(
    x = "Occasion",
    y = "Depression Symptoms",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

The group means start close together. By Week 6, Therapy has dropped farther than Control. The lines are not parallel, which is still the visual signature of an interaction. Here, it means **unequal changes**.

### Fit the Same Interaction with One Extra Term

The earlier 2-by-2 used:

```{r earlier-2x2-syntax, eval=FALSE}
lm(Outcome ~ PredictorA * PredictorB, data = Data)
```

This one uses:

```{r endpoint-model}
EndpointModel <- lmer(
  Symptoms ~ Occasion * Treatment + (1 | Person),
  data = EndpointData
)

summary(EndpointModel)
```

The interaction did not change jobs. It is still the difference of differences. The random intercept is the small change that tells the model Week 0 and Week 6 came from the same people.

### Read the Four Coefficients as Four Cells

Because Week 0 and Control are the reference levels:

| Coefficient | Meaning |
|---|---|
| $B_0$ | Mean symptoms for Control at Week 0 |
| $B_1$ | Week 6 minus Week 0 for Control |
| $B_2$ | Therapy minus Control at Week 0 |
| $B_3$ | How much more or less Therapy changed than Control |

```{r endpoint-coefficients, include=FALSE}
endpoint_fixed <- summary(EndpointModel)$coefficients
endpoint_b3 <- endpoint_fixed["OccasionWeek 6:TreatmentTherapy", ]
```

The interaction estimate is `r round(endpoint_b3["Estimate"], 2)`. Therapy changed by about `r abs(round(endpoint_b3["Estimate"], 2))` points more than Control over the same six weeks. That is the treatment question.

### The Interaction Is Also a *t*-Test on Change Scores

With exactly two occasions, we can subtract Week 0 from Week 6 for every person and compare those change scores between the two treatment groups. This is an independent-samples *t*-test on change.

```{r change-score-test}
ChangeData <- reshape(
  EndpointData[, c("Person", "Treatment", "Time", "Symptoms")],
  idvar = c("Person", "Treatment"),
  timevar = "Time",
  direction = "wide"
)

ChangeData$Change <- ChangeData$Symptoms.6 - ChangeData$Symptoms.0
ChangeData$TreatmentForTest <- relevel(ChangeData$Treatment, ref = "Therapy")

change_result <- t.test(
  Change ~ TreatmentForTest,
  data = ChangeData,
  var.equal = TRUE
)

change_comparison <- data.frame(
  method = c("t-test on change scores", "Mixed-model interaction"),
  t = c(
    unname(change_result$statistic),
    endpoint_b3["t value"]
  ),
  df = c(
    unname(change_result$parameter),
    endpoint_b3["df"]
  ),
  p = c(
    change_result$p.value,
    endpoint_b3["Pr(>|t|)"]
  ),
  estimate = c(
    mean(ChangeData$Change[ChangeData$Treatment == "Therapy"]) -
      mean(ChangeData$Change[ChangeData$Treatment == "Control"]),
    endpoint_b3["Estimate"]
  )
)

print(change_comparison, digits = 6, row.names = FALSE)
```

Again, one answer printed twice. In a complete, balanced 2-by-2 like this one, the mixed-model interaction is the same comparison as the *t*-test on change scores. The mixed model wins when the design grows beyond two occasions or people miss visits.

### Follow Up the Interaction

The interaction tells us the two changes differ. To describe the pattern, estimate Week 6 minus Week 0 separately for each treatment group:

```{r endpoint-simple-effects}
library(emmeans)

EndpointTrends <- emmeans(EndpointModel, ~ Occasion | Treatment)
contrast(
  EndpointTrends,
  method = list("Week 6 - Week 0" = c(-1, 1)),
  adjust = "none"
)
```

These are **simple effects**. They tell us how each group changed. They do not replace the interaction test, which asks whether the changes differ from each other.

### Reporting the Mixed 2-by-2

> Depression symptoms were analyzed with a linear mixed-effects model containing fixed effects of Occasion, Treatment, and their interaction, plus a participant random intercept. The Occasion-by-Treatment interaction was statistically significant, $b =$ `r round(endpoint_b3["Estimate"], 2)`, $SE =$ `r round(endpoint_b3["Std. Error"], 2)`, $t(`r round(endpoint_b3["df"], 1)`) =$ `r round(endpoint_b3["t value"], 2)`, $p$ `r ifelse(endpoint_b3["Pr(>|t|)"] < .001, "< .001", paste("=", round(endpoint_b3["Pr(>|t|)"], 3)))`, indicating that symptoms decreased more from Week 0 to Week 6 in Therapy than in Control.

The report names the repeated unit, the fixed effects, and the interaction. The model formula can look compact. The design still has to be described in words.

# Advanced Topics

The mixed 2-by-2 is enough for two occasions. Everything below extends that same design. The first extension is the important one: put Weeks 2 and 4 back and treat time as continuous.

## Two Groups by Continuous Time

The 2-by-2 asked whether the Week 0 to Week 6 change differed between groups. It ignored the middle of the study. With four occasions, time can become a quantitative predictor and the interaction becomes a difference in **slopes**.

```{r continuous-time-model}
GrowthModel <- lmer(
  Symptoms ~ Time * Treatment + (1 + Time | Person),
  data = LongData
)

summary(GrowthModel)
```

The fixed effects answer four questions:

- The intercept is expected symptoms for Control at Week 0.
- `Time` is expected weekly change for Control.
- `TreatmentTherapy` is the Therapy minus Control difference at Week 0.
- `Time:TreatmentTherapy` is how much the Therapy slope differs from the Control slope.

The random time slope allows people to change at different rates. The fixed interaction describes the average difference between the treatment groups. The random slope admits that actual humans decline to move in formation.

```{r continuous-time-plot, echo=FALSE}
#| fig-cap: "All four occasions. Faint lines show individual trajectories; heavy lines show the group means. The 2-by-2 used only the first and last points."
#| fig-alt: "A line graph of depression symptoms at Weeks 0, 2, 4 and 6. Faint participant lines vary in starting point and rate of change. The heavy Control mean line declines gently across time. The heavy Therapy mean line declines more steeply, so the gap between the groups grows across the four occasions."

LongMeans <- aggregate(
  Symptoms ~ Time + Treatment,
  data = LongData,
  FUN = mean
)

ggplot(
  LongData,
  aes(x = Time, y = Symptoms, group = Person, color = Treatment)
) +
  geom_line(alpha = 0.15, linewidth = 0.4) +
  geom_line(
    data = LongMeans,
    aes(group = Treatment),
    linewidth = 1.4
  ) +
  geom_point(
    data = LongMeans,
    aes(group = Treatment),
    size = 2.7
  ) +
  scale_x_continuous(breaks = times) +
  scale_color_manual(values = c("Control" = "#d97757", "Therapy" = "#287c8e")) +
  labs(
    x = "Week",
    y = "Depression Symptoms",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

The interaction has changed from a difference between two changes to a difference between two slopes. The research question has not changed: **did the groups change differently?**

### Center Time Where the Intercept Matters

With Time = 0 at baseline, the intercept is baseline symptoms and the Treatment coefficient is the baseline group difference. If time is centered at Week 6, the intercept becomes expected Week 6 symptoms and the Treatment coefficient becomes the group difference at Week 6.

Predictions remain the same. The coefficient questions change.

```{r center-time-at-week-six}
LongData$TimeAtWeek6 <- LongData$Time - 6

Week6Model <- lmer(
  Symptoms ~ TimeAtWeek6 * Treatment + (1 + TimeAtWeek6 | Person),
  data = LongData
)

fixef(Week6Model)
```

::: {.callout-note title="Centering Time Is a Scientific Decision"}
Choose zero to represent a meaningful occasion: baseline, treatment completion, or another theoretically important point. Centering at the mean is convenient but not automatically the most useful choice.
:::

### Time as Continuous or Categorical

Treat time as continuous when a straight line, or another specified trajectory, is a meaningful summary. Treat it as categorical when change may be irregular and each occasion deserves its own estimate.

```{r categorical-time-model}
LongData$TimeFactor <- factor(
  LongData$Time,
  levels = times,
  labels = c("Week 0", "Week 2", "Week 4", "Week 6")
)

OccasionModel <- lmer(
  Symptoms ~ TimeFactor * Treatment + (1 | Person),
  data = LongData
)

summary(OccasionModel)
```

The continuous model tells a compact growth story. The categorical model tells an occasion-by-occasion story and spends more degrees of freedom doing it. Neither is automatically correct.

### Missing Repeated Observations

Likelihood-based mixed models can retain people with incomplete outcomes under the model's missing-data assumptions. A classical repeated-measures ANOVA may discard an entire participant after one missed visit.

```{r missing-occasions}
set.seed(347)

LongDataMissing <- LongData
miss <- LongDataMissing$Time > 0 & runif(nrow(LongDataMissing)) < 0.10
LongDataMissing$Symptoms[miss] <- NA

MissingModel <- lmer(
  Symptoms ~ Time * Treatment + (1 + Time | Person),
  data = LongDataMissing,
  na.action = na.omit
)

c(
  rows_available = sum(!is.na(LongDataMissing$Symptoms)),
  people_retained = length(unique(LongDataMissing$Person[!is.na(LongDataMissing$Symptoms)]))
)
```

This is not permission to ignore dropout. Report missingness, consider why observations are missing, and conduct sensitivity analyses when missingness may depend on unobserved outcomes.

## Other Sampling Structures: Nested and Crossed

Repeated observations are nested in people, but mixed models are not limited to repeated measures.

- **Nested:** Every lower-level unit belongs to one higher-level unit. Students are nested in classrooms.
- **Crossed:** Units encounter multiple levels of another factor. Participants respond to many stimuli, and each stimulus is seen by many participants.
- **Longitudinal:** Repeated observations are ordered in time within people or other units.

::: {.callout-warning title="The Spreadsheet Does Not Reveal the Design"}
One row per observation does not make rows independent. You must know which person, item, classroom, therapist, or time series produced each row. The design lives in the data-generating process, not the rectangular shape of the file.
:::

### Crossed Participants and Items

In many psychology experiments, participants respond to stimuli sampled from a broader population: words, faces, scenarios, voices, or images of objects that researchers insist are emotionally neutral.

A crossed model can include participant and item effects:

$$
Y_{si}=\beta_0+\beta_1X_{si}+
u_{0s}+u_{1s}X_{si}+v_{0i}+v_{1i}X_{si}+e_{si}.
$$

- $u$ terms vary across subjects.
- $v$ terms vary across items.

```{r crossed-data}
set.seed(345)
n_subject <- 48
n_item <- 24

CrossedData <- expand.grid(
  Subject = factor(seq_len(n_subject)),
  Item = factor(seq_len(n_item))
)

CrossedData$ConditionNum <- ifelse(
  (as.integer(CrossedData$Subject) + as.integer(CrossedData$Item)) %% 2 == 0,
  -0.5, 0.5
)

SubjectRE <- MASS::mvrnorm(
  n_subject, c(0, 0), matrix(c(35, 2, 2, 2), 2, 2)
)
ItemRE <- MASS::mvrnorm(
  n_item, c(0, 0), matrix(c(18, 1, 1, 1.2), 2, 2)
)

s <- as.integer(CrossedData$Subject)
i <- as.integer(CrossedData$Item)

CrossedData$Attention <-
  70 + 5 * CrossedData$ConditionNum +
  SubjectRE[s, 1] + SubjectRE[s, 2] * CrossedData$ConditionNum +
  ItemRE[i, 1] + ItemRE[i, 2] * CrossedData$ConditionNum +
  rnorm(nrow(CrossedData), 0, 5)
```

```{r crossed-model}
CrossedModel <- lmer(
  Attention ~ ConditionNum +
    (1 + ConditionNum | Subject) +
    (1 + ConditionNum | Item),
  data = CrossedData
)

broom.mixed::tidy(CrossedModel, effects = "fixed", conf.int = TRUE)
```

Ignoring item variation treats the particular stimuli as fixed and perfectly representative. Forty participants responding to the same twelve items do not provide forty independent samples of the item population.

::: {.callout-important title="Participants and Items Can Both Be Random Factors"}
If conclusions should generalize to new participants and new stimuli, model both sampling sources when the design supports it. “But the software output was shorter without items” is not a sampling argument.
:::

### Residual Correlation Across Time

Random intercepts and slopes may not absorb every temporal dependency. Nearby observations can remain more similar than distant ones. Some mixed-model engines can represent structures such as autoregressive correlation:

```{r ar1-syntax, eval=FALSE}
AR1Model <- glmmTMB::glmmTMB(
  Symptoms ~ Time * Treatment +
    (1 + Time | Person) +
    ar1(TimeFactor + 0 | Person),
  data = LongData,
  family = gaussian()
)
```

Structured covariance models require correctly ordered occasions and enough information. Do not add `ar1()` merely because the formula looks graduate-level.

## The Short Story

- A paired *t*-test is a mixed model with one two-level within-subjects predictor and a participant random intercept. Same estimate, same *t*, same df, same *p*.
- A mixed 2-by-2 still has four cells and a difference-of-differences interaction. The small change is that one factor repeats within people, so the model must account for the person.
- With exactly two occasions, the interaction is also an independent-samples *t*-test on change scores.
- With three or more occasions, time can be categorical or continuous. Continuous time turns the interaction into a difference between group slopes.
- Build random effects from the sampling structure. Count independent people and items, not spreadsheet rows.
