---
title: "Advanced Mixed Models II: Random Slopes, Centering, and Cross-Level Interactions"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

## When One Slope Does Not Fit Every Cluster

A random-intercept model allows classrooms to start at different exam-score levels but forces the study-hours slope to be identical everywhere. That may be unreasonable. Study time could help more in classrooms with organized instruction and less in classrooms where the final exam covers material communicated through rumors.

## Random-Slope Equations

Level 1:

$$
Y_{ij}=\beta_{0j}+\beta_{1j}X_{ij}+e_{ij}.
$$

Level 2:

$$
\beta_{0j}=\gamma_{00}+u_{0j},
$$

$$
\beta_{1j}=\gamma_{10}+u_{1j}.
$$

Combined:

$$
Y_{ij}=\gamma_{00}+\gamma_{10}X_{ij}+u_{0j}+u_{1j}X_{ij}+e_{ij}.
$$

The random intercept and slope have covariance matrix:

$$
\mathbf G=
\begin{bmatrix}
\tau_{00} & \tau_{01}\\
\tau_{01} & \tau_{11}
\end{bmatrix}.
$$

$\tau_{01}$ describes whether clusters with higher intercepts tend to have steeper or flatter slopes. Centering choices determine which within- and between-cluster questions those coefficients answer [@enders2007].

::: {.callout-note title="A Random Slope Requires Replication Within Clusters"}
To estimate how a slope varies across people or classrooms, the predictor must vary within those units. You cannot estimate a person's random treatment slope if each person experienced only one treatment level.
:::

## Simulating Classroom Slopes

```{r random-slope-data}
set.seed(344)
n_class <- 36
n_student <- 18

ClassInfo <- data.frame(
  Classroom = factor(seq_len(n_class)),
  TeacherSupport = rnorm(n_class)
)

RandomEffects <- MASS::mvrnorm(
  n_class,
  mu = c(0, 0),
  Sigma = matrix(c(30, 2.8, 2.8, 1.2), 2, 2)
)

SlopeData <- do.call(rbind, lapply(seq_len(n_class), function(j) {
  Study <- rnorm(n_student, mean = runif(1, 4.5, 7.5), sd = 1.5)
  data.frame(
    Classroom = factor(j, levels = seq_len(n_class)),
    TeacherSupport = ClassInfo$TeacherSupport[j],
    StudyHours = Study,
    u0 = RandomEffects[j, 1],
    u1 = RandomEffects[j, 2]
  )
}))

SlopeData$StudyMean <- ave(SlopeData$StudyHours, SlopeData$Classroom)
SlopeData$StudyWithin <- SlopeData$StudyHours - SlopeData$StudyMean
SlopeData$StudyMean.C <- SlopeData$StudyMean - mean(SlopeData$StudyMean)

SlopeData$Exam <-
  68 +
  3.2 * SlopeData$StudyWithin +
  2.0 * SlopeData$StudyMean.C +
  2.5 * SlopeData$TeacherSupport +
  1.3 * SlopeData$StudyWithin * SlopeData$TeacherSupport +
  SlopeData$u0 +
  SlopeData$u1 * SlopeData$StudyWithin +
  rnorm(nrow(SlopeData), 0, 6)
```

## Within- and Between-Cluster Effects

The raw predictor contains two relationships:

1. **Within classrooms:** Do students who study more than their classmates score higher?
2. **Between classrooms:** Do classrooms with higher average study time score higher?

Decompose the predictor:

$$
X_{ij}=(X_{ij}-\bar X_j)+\bar X_j.
$$

- $X_{ij}-\bar X_j$ is the cluster-centered within effect.
- $\bar X_j$ is the cluster mean and carries between-cluster information.

::: {.callout-warning title="Group-Mean Centering Changes the Question"}
Grand-mean centering changes the zero point. Group-mean centering separates within- and between-cluster variation. Those operations are not interchangeable merely because both subtract a mean.
:::

## Fit the Random-Slope Model

```{r random-slope-model}
library(glmmTMB)

RandomInterceptModel <- glmmTMB(
  Exam ~ StudyWithin + StudyMean.C + TeacherSupport +
    StudyWithin:TeacherSupport +
    (1 | Classroom),
  data = SlopeData,
  family = gaussian(),
  REML = FALSE
)

RandomSlopeModel <- glmmTMB(
  Exam ~ StudyWithin + StudyMean.C + TeacherSupport +
    StudyWithin:TeacherSupport +
    (1 + StudyWithin | Classroom),
  data = SlopeData,
  family = gaussian(),
  REML = FALSE
)

summary(RandomSlopeModel)
```

The fixed `StudyWithin` coefficient is the expected within-classroom study slope when teacher support is zero. The interaction says that slope changes as teacher support changes. The random-slope variance says classrooms still differ around the modeled slope.

## Compare Random Structures Carefully

```{r compare-random-structures}
anova(RandomInterceptModel, RandomSlopeModel)
```

Likelihood-ratio tests for variance components require caution because zero variance lies on a boundary. Use the comparison as evidence alongside estimated variances, convergence, design, and theoretical need.

Do not automatically choose the largest possible random-effects structure or delete every small variance. A defensible model represents important sampling and repeated-measures structure while remaining estimable by the data.

::: {.callout-important title="A Zero Variance Is Information"}
If a random-slope variance approaches zero, the data may not support meaningful slope differences, the design may contain too few clusters, or the model may be overparameterized. Simplification should follow investigation, not a panicked sequence of deleted parentheses.
:::

## Plot the Cross-Level Interaction

```{r cross-level-plot, fig.width=7, fig.height=4.8}
support_values <- c(Low = -1, Average = 0, High = 1)
study_grid <- seq(-3, 3, length.out = 100)

plot(
  SlopeData$StudyWithin, SlopeData$Exam,
  pch = 19, col = rgb(.2, .2, .2, .12),
  xlab = "Study hours relative to classroom mean",
  ylab = "Predicted exam score",
  main = "A level-2 variable changes a level-1 slope"
)

line_colors <- c("#D55E00", "#0072B2", "#009E73")
for (i in seq_along(support_values)) {
  newdata <- data.frame(
    StudyWithin = study_grid,
    StudyMean.C = 0,
    TeacherSupport = support_values[i],
    Classroom = SlopeData$Classroom[1]
  )
  lines(
    study_grid,
    predict(RandomSlopeModel, newdata = newdata, re.form = NA),
    col = line_colors[i], lwd = 3
  )
}
legend("topleft", legend = paste(names(support_values), "support"),
       col = line_colors, lwd = 3, bty = "n")
```

## Three Levels

Students may be nested in classrooms nested in schools. A random-intercept model becomes:

$$
Y_{ijk}=\gamma_{000}+v_{00k}+u_{0jk}+e_{ijk},
$$

where $k$ indexes schools, $j$ classrooms, and $i$ students.

In `glmmTMB`, nested intercepts can be written:

```{r three-level-syntax, eval=FALSE}
ThreeLevelModel <- glmmTMB(
  Score ~ Predictor + (1 | School / Classroom),
  data = ThreeLevelData,
  family = gaussian()
)
```

This expands to school intercepts plus classroom-within-school intercepts. Additional random slopes require enough clusters and variation at the level where the slope is estimated.

## Effect Size

```{r random-slope-effect-size}
performance::r2_nakagawa(RandomSlopeModel)
performance::icc(RandomSlopeModel)
```

Marginal $R^2$ summarizes fixed-effect variation; conditional $R^2$ includes random effects. The ICC in a random-slope model can depend on predictor values because cluster similarity changes with $X$.

## The Short Story

Random slopes allow predictor effects to vary across clusters. Separate within- and between-cluster information before interpreting level-1 predictors. Cross-level interactions explain some slope variation; random-slope variance describes what remains.
