---
title: "From the Paired *t*-Test to the One-Way Within-Subjects ANOVA"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 4.75,
  fig.height = 2.75
)
```

## Types of Design
- Repeated: The same person is measured twice.
- Matched: "Doppelgangers we pretend are clones."
    - A. Natural matches, such as twins or siblings
    - B. Participants matched on other characteristics, such as IQ, age, gender, or other controls

### Correlated Measures
- When participants repeat trials or complete multiple conditions, their scores tend to be **correlated**.
    - We want to control for individual variability.
    - In the formula below, we remove shared variance from the standard error using Pearson's r between conditions.

$$ t = \frac{M_1-M_2}{\sqrt{\frac{S_1^2 + S_2^2}{n}-\frac{2rS_1S_2}{n} }}$$

#### Between-Subjects Assumption

Below is a covariance matrix for a between-subjects design with two levels, as in an independent-samples *t*-test. We assume HOV, meaning that the population variances are equal: $\sigma_1^2 = \sigma_2^2 = \sigma^2$. The cells show the correlations between groups. Note that $\rho$ is the population correlation.

$$
\mathbf{Cov} = \sigma^2\left[\begin{array}
          {rrr}
          1 & 0  \\
          0 & 1  \\
          \end{array}\right]
$$

- Cell 1 = Group 1 with Group 1 ($\rho$ = 1): The cell data are correlated with themselves.
- Cell 2 = Group 1 with Group 2 ($\rho$ = 0): The groups are unrelated, so the correlation should be around zero.
- Cell 3 = Group 2 with Group 1 ($\rho$ = 0): Same as Cell 2 because the matrix is symmetric.
- Cell 4 = Group 2 with Group 2 ($\rho$ = 1): The cell data are correlated with themselves.

#### Within-Subjects Assumption 

Below is a covariance matrix for a within-subjects design with two levels, as in a paired-samples *t*-test. We still assume HOV, meaning that the population variances are equal: $\sigma_1^2 = \sigma_2^2 = \sigma^2$. The cells show the correlations between conditions.

$$
\mathbf{Cov} = \sigma^2\left[\begin{array}
          {rrr}
          1 & \rho  \\
          \rho & 1  \\
          \end{array}\right]
$$ 
    
- Cell 1 = Group 1 with Group 1 ($\rho$ = 1): The cell data are correlated with themselves.
- Cell 2 = Group 1 with Group 2 ($\rho$ > |0|): Each person's scores are correlated across conditions and should NOT be zero.
- Cell 3 = Group 2 with Group 1 ($\rho$ > |0|): Same as Cell 2 because the matrix is symmetric.
- Cell 4 = Group 2 with Group 2 ($\rho$ = 1): The cell data are correlated with themselves.


## One Repeated Factor: Two-Level Designs

> Does eating chocolate during a lecture increase fixation on the lecture? We measure fixation as the mean time a student's eyes remain on the screen, in seconds per glance. We collect multiple trials per condition for each person and compare the means with a paired-samples *t*-test. The DV is mean fixation time. The IV is a two-level repeated factor: no chocolate in Lecture 1 versus one fun-size Snickers in Lecture 2. We simulate $n = 5$ students measured during both lectures, with means of $M_1 = 5$ and $M_2 = 10$ and $\sigma^2 = 4$. The responses across conditions are correlated at $\rho = .4$.

```{r, echo=FALSE}
set.seed(7580)
library(tidyr)
library(dplyr)
library(MASS)
n = 5 ; p = .4 ; M1=5; M2=10; S2=4
# Generate Cov matrix
SigmaMatrix<-matrix(c(1,p, 
                      p,1), nrow=2, ncol=2)*S2

DataSim0 = mvrnorm(n=n, mu=c(M1, M2), Sigma=SigmaMatrix, empirical=FALSE)
colnames(DataSim0) <- c("X1","X2")
DataSim1<-gather(data.frame(DataSim0),key = "Chocolate", value = "Fixation")
DataSim1$SS<-as.factor(rep(seq(1:n),2))
DataSim1$Chocolate<-factor(DataSim1$Chocolate, 
       levels=c("X1","X2"),
       labels=c("No Snickers", "Fun Size Snickers"))
```

### Plot of Simulation
- Spaghetti plot per person
- Each line represents a person's score from Lecture 1 (No Chocolate) to Lecture 2 (Chocolate).
- As you can see, most do better, while some do worse.

```{r, echo=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "")) 
Rm.t.test.plot <-ggplot(data = DataSim1, aes(x = Chocolate, y=Fixation))+
  geom_point(aes(colour = SS))+
  geom_smooth(method = "lm", se = FALSE, aes(group=SS,colour = SS))+
  ylab("Fixation (in seconds)")+xlab("Chocolate")+
  theme(legend.position = "none")
Rm.t.test.plot
```

```{r, echo=FALSE}
Rval=round(cor(DataSim1$Fixation[1:5],DataSim1$Fixation[6:10]),2)
```


The correlation matrix:  

Corr Matrix |None     | Snickers
------------|---------|---------
None        |   1     | `r Rval`
Snickers    | `r Rval`| 1

The value of r is reasonably close to the $\rho = .4$ that we simulated.

Variance per cell:  

```{r, echo=FALSE}
Vars<-
DataSim1 %>%
  group_by(Chocolate)%>%
  summarise(Varience=var(Fixation))
knitr::kable(Vars)
```

The variances are not very homogeneous, but we have an $N$ of only 5.

### Conduct a Paired *t*-Test in R
We use the `t.test()` function with the argument `paired = TRUE`. First, we reshape the data so that each row contains one participant's scores in both conditions. This makes the pairing explicit. The $df$ equals the number of participants minus 1.

```{r}
DataSim1_Wide <- DataSim1 %>%
  dplyr::select(SS, Chocolate, Fixation) %>%
  pivot_wider(names_from = Chocolate, values_from = Fixation)

T1 <- with(
  DataSim1_Wide,
  t.test(`Fun Size Snickers`, `No Snickers`, paired = TRUE)
)
T1
```


#### APA-Style Report
The `apa` package can automatically convert our *t*-test result into APA format.

```{r,results="asis"}
library(apa)
cat(apa(T1, format = "rmarkdown", print = FALSE))
```


### Logic of the Paired *t*-Test
We remove **individual differences** because we care about how people **change** from one condition to another. Although the raw data look like the spaghetti plot above, the analysis removes these individual differences by calculating a difference score: *Fun-Size Snickers Condition* - *No Snickers Condition*.

Now that we have difference scores, we will plot the dots:

```{r, echo=FALSE}
DataSim2<-data.frame(diff=(DataSim0[,2]-DataSim0[,1]))

theme_set(theme_bw(base_size = 12, base_family = "")) 
Rm.t.test.plot.2 <-ggplot(data = DataSim2, aes(x="Snickers - None" , y=diff))+
  geom_point()+
  ylab("Fixation Difference Score")+xlab("")+
  theme(legend.position = "none")
Rm.t.test.plot.2
```

#### One-Sample *t*-Test on Difference Scores
Because we have one score per person, the difference between conditions, we can run a one-sample *t*-test to determine whether the difference scores differ from zero.

$$t = \frac{M - \mu}{\sqrt{S^2/n}}$$ 

```{r}
T2<-t.test(x=DataSim2$diff, y = NULL, mu=0)
T2
```

Notice that the results are identical to those from the paired-samples *t*-test.

#### Simplified Paired *t*-Test Formula
To calculate this paired *t*-test by hand, calculate the difference scores ($D$), then find their mean and variance, or $S$. The result is identical to that from the more complex formula above, which removes individual differences using the Pearson correlation.

$$t = \frac{M_D - \mu_D}{\sqrt{S_D^2/n}}$$ 

### Effect Size
There is considerable debate about calculating Cohen's $d$ for paired designs. The simple formula looks like Cohen's $d$ for a one-sample *t*-test, except that it uses difference scores:

$$d = \frac{M_D}{S_D}$$

## One Repeated Factor: Designs with Three or More Levels
> Does eating chocolate during a lecture increase fixation on the lecture? We measure fixation as the mean time a student's eyes remain on the screen, in seconds per glance. We collect multiple trials per condition for each person and compare the means with a repeated-measures ANOVA. The DV is mean fixation time. The IV is a three-level repeated factor: no chocolate in Lecture 1, one fun-size Snickers in Lecture 2, and one king-size Snickers in Lecture 3. We simulate $n = 5$ students measured during all three lectures, with means of $M_1 = 5$, $M_2 = 10$, and $M_3 = 3$, assuming HOV with $\sigma^2 = 4$. Responses across conditions are correlated at $\rho = .4$.


```{r, echo=FALSE}
set.seed(42)
n = 5 ; p = .4 ; M1=5; M2=10;M3=3; S2=4
# Generate Cov matrix
SigmaMatrix<-matrix(c(1,p,p, 
                      p,1,p,
                      p,p,1), nrow=3, ncol=3)*S2

DataSim3 = mvrnorm(n=n, mu=c(M1, M2,M3), Sigma=SigmaMatrix, empirical=FALSE)
colnames(DataSim3) <- c("X1","X2","X3")
DataSim4<-gather(data.frame(DataSim3),key = "Chocolate", value = "Fixation")
DataSim4$SS<-as.factor(rep(seq(1:n),3))
DataSim4$Chocolate<-factor(DataSim4$Chocolate, 
       levels=c("X1","X2","X3"),
       labels=c("No Snickers", "Fun Size Snickers","King Size Snickers"))
DataSim4$Fixation<-round(DataSim4$Fixation,0)

```


### Plot of Simulation
- Spaghetti plot per person
- Each line represents a person's score across conditions

```{r, echo=FALSE}
theme_set(theme_bw(base_size = 12, base_family = "")) 
Rm.t.test.plot.2 <-ggplot(data = DataSim4, aes(x = Chocolate, y=Fixation, group=SS))+
  geom_point(aes(colour = SS))+
  geom_line(aes(colour = SS))+
  #geom_smooth(method = "lm", se = FALSE, aes(group=SS,colour = SS))+
  ylab("Fixation (in Seconds)")+xlab("Chocolate")+
  theme(legend.position = "none")
Rm.t.test.plot.2
```


### Repeated-Measures ANOVA Logic
In a repeated-measures ANOVA, we identify variance due to individual differences. We estimate it from the row sums, which are the sums of each participant's scores.

#### One-way ANOVA Table With the formulas (conceptually simplified)


Source | SS     | DF     | MS     | F
-------|--------|--------|--------|----
Between| $SS_B=nSS_{treatment}$ | $K-1$ | $MS_{B}=\frac{SS_B}{df_B}$ | $\frac{MS_B}{MS_W}$
Within | $SS_W=\displaystyle \sum SS_{within}$| $N-K$ | $MS_{W}=\frac{SS_W}{df_W}$ | 
Total  | $SS_T=SS_{scores}$   | $N-1$ 

Note: $SS_B + SS_W = SS_T$ and $df_B + df_W = df_T$.

#### One-way RM ANOVA Table With the formulas (conceptually simplified)

Source     | SS                    | DF      | MS       | F
-----------|-----------------------|---------|----------|---
$RM_1$ | $SS_{RM}=nSS_{all\,cell\,means}$ | $c-1$| $MS_{RM}=\frac{SS_{RM}}{df_{RM}}$ | $\frac{MS_{RM}}{MS_E}$
Within  | $SS_{w}=\displaystyle \sum SS_{within}$     |   
$\,Subject\,[Sub]$  | $SS_{sub}=\frac{1}{c} SS_{subjects:\, rows\,sums}$     | $n-1$     | $MS_{Sub}=\frac{SS_{sub}}{df_{sub}}$ | 
$\,Error [Sub\,x\,RM_1]$  | $SS_E=SS_W-SS_{sub}$     | $(n-1)(c-1)$     | $MS_E=\frac{SS_E}{df_E}$ | 
Total   | $SS_T=SS_{scores}$      | $N-1$ 

Note: $SS_{sub} + SS_E = SS_W$, $SS_B + SS_W = SS_T$, and $df_B + df_W = df_T$.

##### Explained vs. Unexplained Terms

**Explained:** 

$SS_{RM}$ = variance caused by the treatment.
$SS_{sub}$ = individual differences. We do not know why people differ, but we can parse this variance, so we have *explained it*.

**Unexplained:** 

$SS_{E}$ = variance within participants that cannot be explained. This is not individual-difference variance.


### Assumptions
As with the paired *t*-test, we assume HOV ($\sigma^2_1 = \sigma^2_2 =\sigma^2_3$). More importantly, we assume that the correlations among the three conditions are equal ($\rho_1 = \rho_2 = \rho_3$). We call this *compound symmetry*:

$$
\mathbf{Cov} = \sigma^2\left[\begin{array}
          {rrrr}
          1 & \rho & \rho  \\
          \rho & 1 & \rho  \\
          \rho & \rho & 1  \\
          \end{array}\right]
$$ 

When HOV and compound symmetry are met, we satisfy the assumption of **homogeneity of covariance**.

#### Sphericity
We evaluate this assumption with Mauchly's test of sphericity, which tests the variances of the difference scores among all conditions.

$H_0: \sigma^2_{2-1}=\sigma^2_{3-1}=\sigma^2_{3-2}$

$H_1: \sigma^2_{2-1}\neq\sigma^2_{3-1}\neq\sigma^2_{3-2}$

When sphericity is violated, we reduce the df to lower the risk of Type I error. We will not cover the calculation in detail. The basic idea is that a value called $\epsilon$ is calculated from the degree of violation. The greater the violation, the larger the df correction. The old-school approach corrected only after rejecting the null in Mauchly's test. The modern approach always applies a sphericity correction, even when Mauchly's test is nonsignificant. R follows the second approach by default.

There are two methods for calculating $\epsilon$. The most common and slightly more conservative is the Greenhouse-Geisser (GG) correction. A more powerful but less common approach is the Huynh-Feldt (HF) correction. R defaults to GG, as do psychologists, but you can request HF.

### Effect Sizes

#### What SPSS and Your Book Suggest
Your book suggests an effect size that is very similar to $\eta^2_p$.

Remember in two-way ANOVA: 

$$\eta_p^2 = \frac{SS_A}{SS_A+SS_W}$$

Because our error term is no longer $SS_W$, we now use $SS_E$.

$$\eta_p^2 = \frac{SS_{RM}}{SS_{RM}+SS_E}$$

#### R and Generalized Effect Size
Olejnik and Algina (2003) instead propose generalized eta-squared, which can be compared across within- and between-subjects designs and is what `afex` has been reporting.

$$\eta_g^2 = \frac{SS_{RM}}{SS_{RM} + SS_{sub}+ SS_E} = \frac{SS_{RM}}{SS_T} $$

You can also calculate an unbiased generalized effect size by hand or use the Excel sheets.

$$\omega_g^2 = \frac{df_{RM}(MS_{RM}-MS_E)}{MS_{sub}+ SS_T} $$

See Bakeman (2005) for additional issues concerning effect sizes in repeated-measures and mixed designs.

### R Calculations
#### R Defaults
- `afex` calculates the repeated-measures ANOVA and automatically corrects for sphericity violations.
- `Error(Subject ID/RM factor)`: This tells `afex` that participants vary as a function of the repeated-measures condition.
- It reports $\eta_g^2$ assuming a manipulated treatment.

```{r}
library(afex)
Model.1<-aov_car(Fixation~ Chocolate + Error(SS/Chocolate), 
        data = DataSim4)
Model.1
```

#### SPSS-Like Report
Use the following code to see results formatted like SPSS output:

```{r}
aov_car(Fixation~ Chocolate + Error(SS/Chocolate), 
        data = DataSim4, return="univariate")
```

This reports Mauchly's test, which is nonsignificant, along with the uncorrected df and p-value. It also shows the p-values for the GG and HF corrections, if applied.

#### Manually Select DF Correction & $\eta_p^2$

You can request `HF`, `GG`, or `none`. You can also request $\eta_p^2$ with `pes`.

```{r}
aov_car(Fixation~ Chocolate + Error(SS/Chocolate), 
        data = DataSim4,
        anova_table=list(correction = "HF", es='pes'))
```

## Challenges of Repeated Design
- Practice effects
- Carryover effects
- Order effects
- Fatigue

### Solutions
- Counterbalancing
- Time between trials/distractor tasks
- Using a matched design

## References
Bakeman, R. (2005). Recommended effect size statistics for repeated measures designs. *Behavior Research Methods*, 37(3), 379-384.

Olejnik, S., & Algina, J. (2003). Generalized eta and omega squared statistics: Measures of effect size for some common research designs. *Psychological Methods*, 8(4), 434-447. doi:10.1037/1082-989X.8.4.434
