---
title: "Mixed Regression: Between People and Within People"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5.5,
  fig.height = 3.6,
  fig.align = "center",
  fig.pos = "H",
  results = "hold"
)
```

## Your Paired *t*-Test Was a Mixed Model All Along

The [paired-samples *t*-test chapter](Chapter_Paired_t_test.qmd) gave 48 students one lecture with chocolate and one without, then measured how long their eyes stayed on the screen. Here is that study again, same seed and same 48 students, so this chapter runs from the top without sending you back a chapter to fetch the data.

```{r simulate-chocolate-data}
set.seed(343)

n <- 48

# Each student has a personal tendency to stare that follows them into BOTH
# conditions. That shared tendency is the entire reason pairing works.
participant_effect <- rnorm(n, mean = 0, sd = 1.20)

paired_wide <- data.frame(
  participant  = factor(seq_len(n)),
  no_chocolate = 5.00 + participant_effect + rnorm(n, 0, 0.80),
  chocolate    = 5.75 + participant_effect + rnorm(n, 0, 0.80)
)

paired_wide$difference <-
  paired_wide$chocolate - paired_wide$no_chocolate

# 48 rows become 96: one row per measurement instead of one row per person.
paired_long <- tidyr::pivot_longer(
  paired_wide,
  cols      = c(no_chocolate, chocolate),
  names_to  = "condition",
  values_to = "fixation"
)

paired_long$condition <- factor(
  paired_long$condition,
  levels = c("no_chocolate", "chocolate"),
  labels = c("No chocolate", "Chocolate")
)
```

There are two ways to analyze this. The first is the one you already know. Collapse each student down to a single difference score and ask whether those differences average to zero:

```{r chocolate-paired-t}
paired_result <- t.test(
  paired_wide$chocolate,
  paired_wide$no_chocolate,
  paired = TRUE
)

paired_result
```

The second way never computes a difference at all. It hands R all 96 measurements in long format, plus one piece of information the *t*-test was only ever told indirectly, through the row order: which measurements came from the same human.

```{r chocolate-lmer}
library(lme4)      # fits the mixed model
library(lmerTest)  # adds degrees of freedom and p-values to the fixed effects

chocolate_model <- lmer(
  fixation ~ condition + (1 | participant),
  data = paired_long
)

summary(chocolate_model)
```

Find the `conditionChocolate` row in that output. Then look back at the *t*-test.

```{r chocolate-identity}
choc_fixed <- summary(chocolate_model)$coefficients

comparison <- data.frame(
  method   = c("paired t-test", "lmer"),
  t        = c(unname(paired_result$statistic),
               choc_fixed["conditionChocolate", "t value"]),
  df       = c(unname(paired_result$parameter),
               choc_fixed["conditionChocolate", "df"]),
  p        = c(paired_result$p.value,
               choc_fixed["conditionChocolate", "Pr(>|t|)"]),
  estimate = c(unname(paired_result$estimate),
               choc_fixed["conditionChocolate", "Estimate"])
)

print(comparison, digits = 6, row.names = FALSE)
```

Those are not two answers that happen to agree. They are one answer, printed twice.

The degrees of freedom are worth a second look, because the two rows got to 47 by completely different routes. The *t*-test knows it has 48 differences and subtracts one. `lmer` has no idea what a difference score is; `lmerTest` works out how much independent information the 96 rows actually contain, using an approximation called **Satterthwaite's method**, and arrives at 47 anyway.

### What `(1 | participant)` Actually Did

That term is the whole difference between the two calls. It tells R that rows sharing a participant number came from the same person, and that people arrive at a study with their own resting level of staring. Some students watch the screen. Some students watch the door.

It handles this without estimating 48 separate student coefficients. It estimates the *spread* of the students, a single number, and treats each person's own level as a draw from it:

```{r chocolate-variance}
print(VarCorr(chocolate_model), comp = c("Variance", "Std.Dev."))
```

The `participant` line is how much students differ from each other. The `Residual` line is how much one student bounces around their own level from one measurement to the next. Those are the two piles the 96 numbers get sorted into, and we can ask what fraction of the variance sits in the first pile:

```{r chocolate-icc}
choc_var <- as.data.frame(VarCorr(chocolate_model))
icc <- choc_var$vcov[1] / sum(choc_var$vcov)

icc
```

So roughly `r round(icc * 100)` percent of the variation in fixation time is students being different students, before chocolate enters the discussion. The paired *t*-test dealt with that variation by subtracting it away and never mentioning it again. The mixed model estimates it and prints it.

That proportion is also the model's answer to a question the paired *t*-test chapter asked directly: how much do a person's two measurements agree? That chapter measured the agreement head-on, as a plain correlation of `r round(cor(paired_wide$chocolate, paired_wide$no_chocolate), 2)`, and called it the $r$ in the $-2rS_1S_2$ term that everything depended on. The two numbers are close without being identical, because they are two different estimates of the same thing: one computed directly from the pairs, one implied by the variances the model fit. The point is that the quantity has stopped being a term in a formula and become a parameter with a variance attached.

"Each person is their own control" is a sentence from a methods lecture. `(1 | participant)` is that sentence written as an equation.

### The Fixed Effect Is the Chocolate Effect

The **fixed effects** are the part of the model that answers the research question, and there are only two:

- `(Intercept)`, `r round(fixef(chocolate_model)[1], 2)` seconds, is average fixation time with no chocolate.
- `conditionChocolate`, `r round(fixef(chocolate_model)[2], 2)` seconds, is what chocolate adds on top of that.

The second one is not merely close to the *t*-test's mean difference. It is the mean difference, `r round(mean(paired_wide$difference), 2)` seconds, reached from the opposite direction: the *t*-test computed 48 differences and averaged them, while the model never computed a single difference and got there anyway.

Fixed effects answer the research question. Random effects are how the model earns the right to answer it using 96 rows that came from 48 people.

### So What Was the Point of All That?

Nothing yet, honestly. With two measurements and no other predictors, the paired *t*-test is shorter to type and gives the same answer, so keep using it. The mixed model has not earned anything.

It earns its keep the moment the design acquires something a difference score cannot hold:

- **A between-person predictor.** Treatment group, age, baseline severity. Subtracting within a person throws the person away, and group is a property of the person.
- **A missing row.** A student who skipped a session has no difference score at all, so the *t*-test deletes the entire student.
- **Three or more occasions.** There stops being such a thing as *the* difference.
- **People who change at different *rates*,** not merely by different amounts. A difference score has nowhere to put a slope.

The rest of this chapter does the first two. The last two need a bigger model than this chapter builds, and they get their own chapter later in the book.

::: {.callout-note title="Where This Chapter Goes, and Where It Stops"}
Two stops, and you just finished the first. Next is the **mixed 2-by-2**: one within-subjects factor, one between-subjects factor, four cells, one interaction. That is the design most repeated-measures studies in psychology actually run, and it is where this chapter stops.

A mixed design combines a **between-subjects** variable with a **within-subjects** variable, which summons fixed effects, an interaction, repeated observations, and random effects all at once. It is not impossible. It is merely statistics arriving with all its relatives and expecting you to provide chairs.
:::

## The Mixed 2-by-2: Baseline versus Week 6, by Group

Suppose researchers compare cognitive behavioral therapy (**CBT**) with a control condition for reducing depression symptoms. Participants are randomly assigned to CBT or control and complete a depression inventory twice: at baseline and again after six weeks.

- **Occasion** is within subjects: every person appears at baseline and at Week 6.
- **Group** is between subjects: each person is in CBT *or* control, never both.
- **Depression score** is the quantitative outcome.

You have seen this shape before. The [categorical-by-categorical interaction chapter](Chapter_CatXCat_Interaction.qmd) had two categorical predictors and four cells. So does this:

|  | **Control** | **CBT** |
|---|---|---|
| **Baseline** | $\mu_{11}$: Control at baseline | $\mu_{12}$: CBT at baseline |
| **Week 6** | $\mu_{21}$: Control at Week 6 | $\mu_{22}$: CBT at Week 6 |

The research question is not whether the groups differ, and not whether scores change. It is whether they change **differently**:

$$B_3 = (\mu_{22}-\mu_{12})-(\mu_{21}-\mu_{11})$$

Find the change in CBT. Find the change in Control. Compare the two changes. That is the same **difference of differences** the earlier chapter spent itself on, and it has not changed jobs here.

### What Stays the Same, and the One Thing That Changes

| Feature | The earlier categorical 2-by-2 | This mixed 2-by-2 |
|---|---|---|
| Categorical predictors | Two | Two |
| Cells | Four | Four |
| The interaction | Difference of differences | Difference of differences |
| Who fills the cells | Different people in every cell | The same person fills two of them |
| The model | `lm(Y ~ A * B)` | the same, plus a participant random intercept |

The fixed part of the formula does not change at all. We add the random intercept because the two rows from one person are related, and pretending they came from strangers would count one human twice and let the model become confident in ways it has not earned.

### The Data

This is simulated data. Real treatment studies require clinical expertise, ethical oversight, careful monitoring, informed consent, and substantially more planning than "Alex found `rnorm()`."

```{r simulate-mixed-data}
set.seed(3436)

n_per_group <- 50

people <- data.frame(
  participant = factor(seq_len(2 * n_per_group)),
  group = factor(
    rep(c("Control", "CBT"), each = n_per_group),
    levels = c("Control", "CBT")
  ),
  # each person's own baseline level, and their own rate of change
  person_intercept = rnorm(2 * n_per_group, 0, 5.0),
  person_slope     = rnorm(2 * n_per_group, 0, 0.45)
)

# every person gets a baseline row and a Week 6 row
mixed_data <- merge(people, data.frame(week = c(0, 6)), by = NULL)
mixed_data <- mixed_data[order(mixed_data$participant,
                               mixed_data$week), ]

mixed_data$depression <-
  35 +
  mixed_data$person_intercept +
  (-0.20 * mixed_data$week) +                     # control drifts down slowly
  ifelse(mixed_data$group == "CBT",
         -1.20 * mixed_data$week, 0) +            # CBT drops faster: the effect
  mixed_data$person_slope * mixed_data$week +
  rnorm(nrow(mixed_data), 0, 2.0)

# the same variable twice: week as a number, occasion as a readable label
mixed_data$occasion <- factor(
  mixed_data$week,
  levels = c(0, 6),
  labels = c("Baseline", "Week 6")
)
```

The data are **long**: one row per person per occasion, 200 rows from 100 people. The participant identifier repeats because the person repeats. If your participant column disappears, stop. Do not proceed directly to the ritual sacrifice.

Notice that the simulation gave every person their own `person_slope`, so the invented humans genuinely do improve at different rates. Hold onto that. It matters at the end of the chapter, where it turns out this model cannot see it.

### Look at the Four Cells First

```{r cell-means}
aggregate(
  depression ~ occasion + group,
  data = mixed_data,
  FUN = function(x) c(M = mean(x), SD = sd(x))
)
```

```{r mixed-design-plot}
#| fig-cap: "The four cells before any model. Each faint line is one participant's baseline and Week 6 score. The heavy lines connect the group means. Non-parallel lines are the visual signature of an interaction, and here that means unequal change."
#| fig-alt: "A line plot with Occasion on the x axis, Baseline and Week 6, and depression score on the y axis running from about 10 to about 53. One hundred faint lines each connect one participant's two scores, colored salmon for Control and teal for CBT, and they slope in both directions. Two heavy lines with round markers connect the group means. Both groups start near 35. The Control mean falls slightly to about 34, while the CBT mean falls steeply to about 27, so the two heavy lines are clearly not parallel."
library(ggplot2)

group_means <- aggregate(
  depression ~ occasion + group,
  data = mixed_data,
  FUN = mean
)

ggplot(
  mixed_data,
  aes(x = occasion, y = depression, group = participant, color = group)
) +
  geom_line(alpha = 0.15) +
  geom_line(
    data = group_means,
    aes(group = group), linewidth = 1.4
  ) +
  geom_point(
    data = group_means,
    aes(group = group), size = 2.8
  ) +
  labs(
    x = NULL,
    y = "Depression score",
    color = "Group",
    title = "Group averages sitting on top of actual humans"
  ) +
  theme_minimal(base_size = 11)
```

The two groups start at nearly the same place, which is what randomization is supposed to buy. By Week 6 the CBT mean has dropped much further. But people do not all begin at the same score or change by the same amount. Some improve a lot, some barely, and some appear to have received therapy from a motivational poster in a dentist's office.

Never decide whether an effect is statistically significant by staring at whether confidence intervals overlap. CI overlap is not the direct test of a difference, and interaction tests are about **differences between differences**. Fit the model that asks the question.

## Why Ordinary Regression Is Not Enough

We could hand those four cells to `lm()` and it would answer without complaint:

```{r ordinary-model-warning}
ordinary_model <- lm(depression ~ occasion * group, data = mixed_data)
```

That model treats all 200 rows as independent. They are not. Two rows belong to each participant, and two scores from the same person know each other, which is the identity theft the paired *t*-test chapter warned about.

This structure is called **nesting**:

$$
\text{occasions nested within participants}
$$

The levels are:

- Level 1: repeated occasions ($j$);
- Level 2: participants ($i$).

Group belongs to Level 2 because it does not change within a person. Occasion belongs to Level 1 because it does.

We will come back to `ordinary_model` once the mixed model is fit, because the comparison is more interesting than the warning.

## The Fixed Effects

The fixed part is ordinary regression on the four cells:

$$
Y_{ij}=b_0+b_1(\text{Week 6}_{ij})+b_2(\text{CBT}_i)
+b_3(\text{Week 6}_{ij}\times\text{CBT}_i)+\text{random part}+e_{ij}
$$

Because Baseline and Control are the reference levels, each coefficient is one step around the 2-by-2:

| Coefficient | What it estimates |
|---|---|
| $b_0$ | Mean depression for Control at baseline |
| $b_1$ | Week 6 minus baseline, within Control |
| $b_2$ | CBT minus Control, at baseline |
| $b_3$ | How much *more* CBT changed than Control |

The CBT change is $b_1+b_3$. The interaction $b_3$ is the treatment question, and if it is negative then CBT scores fell further than Control scores.

Notice why coding matters. A "main effect of Group" is not a universal average group difference here. It is the group difference *at baseline*, before anybody has been treated, which in a randomized trial we hope is nothing at all. Choosing baseline as the reference is what makes that coefficient answer a question worth asking instead of describing the difference during Week Zero Hundred and Twelve, a period known only to the Probability Gods.

## The Random Intercept

The repeated observations require us to say how people differ. Each participant gets their own starting level:

$$
Y_{ij}=b_0+b_1\text{Week 6}_{ij}+b_2\text{CBT}_i
+b_3(\text{Week 6}_{ij}\times\text{CBT}_i)+u_{0i}+e_{ij}
$$

That $u_{0i}$ is the same `(1 | participant)` that reproduced the paired *t*-test at the top of the chapter, doing the same job on a larger design: give participants different intercepts, by estimating the variance of those intercepts rather than 100 unrelated participant coefficients.

```{r mixed-model}
mixed_model <- lmer(
  depression ~ occasion * group + (1 | participant),
  data = mixed_data
)

summary(mixed_model)
```

::: {.callout-note title="REML: The Deliberately Oversimplified Version"}
By default, `lmer()` uses **REML** (*restricted maximum likelihood*) to estimate the model. I am going to hand-wave aggressively here because understanding exactly how REML works requires considerably more statistical machinery than we need.

You already know **OLS**, which chooses regression coefficients to make the residual errors as small as possible. Mixed-effects models have an additional problem: they also have to estimate **variances**, such as how much participant intercepts vary from person to person.

REML is a method designed to get better estimates of those variance components. Very loosely, it tries to estimate the variance **after accounting for the fact that we already used the data to estimate the fixed effects**. That correction matters especially when samples are not enormous.

For now, the practical rule is simple:

**When we care about estimating the random-effects structure, we generally use REML. When we later compare models that differ in their fixed effects, we will switch REML off.**

That is intentionally an oversimplification. Future Us will deal with the details when Future Us actually needs them.
:::

### Why There Is No Random Slope in That Formula

The obvious next thought is to let people change at different *rates*, not merely start at different levels, by writing `(1 + week | participant)`. Try it on this data and `lmer` refuses:

```
number of observations (=200) <= number of random effects (=200) for
term (1 + week | participant); the random-effects parameters and the
residual variance are probably unidentifiable
```

This is not a bug and it is not fixable by trying a different optimizer. With two measurements per person, a line through each person's two points is *exactly determined*: two points, one line, nothing left over. There is no information left to tell "this person changed unusually fast" apart from "this measurement was unusually noisy," so the model cannot estimate both.

The individual differences in change are real. We put them in the simulation ourselves. They are simply invisible to a two-occasion design, and they get absorbed into the residual variance, where they inflate it. Three occasions is the price of seeing them, and that is a later chapter.


## Read the Model Without Panic

```{r fixed-effects-table}
fixed_effects <- summary(mixed_model)$coefficients
round(fixed_effects, 3)
```

Read the rows in order, and check each one against the table of four coefficients above:

1. `(Intercept)`, `r round(fixed_effects["(Intercept)", "Estimate"], 2)`, is the Control group's baseline mean.
2. `occasionWeek 6`, `r round(fixed_effects["occasionWeek 6", "Estimate"], 2)`, is how much Control changed over six weeks.
3. `groupCBT`, `r round(fixed_effects["groupCBT", "Estimate"], 2)`, is the baseline difference between groups, and it is not significant, $p =$ `r round(fixed_effects["groupCBT", "Pr(>|t|)"], 3)`. Randomization did its job.
4. `occasionWeek 6:groupCBT`, `r round(fixed_effects["occasionWeek 6:groupCBT", "Estimate"], 2)`, is the interaction.

So CBT scores fell about `r abs(round(fixed_effects["occasionWeek 6:groupCBT", "Estimate"], 2))` points further than Control scores over the same six weeks. That is the treatment effect. It does not mean CBT lowered every participant's score by that amount. Fixed effects describe an average pattern while the random effects document the inconvenient existence of individuals.

<!-- ---Alex--- The paragraph below is new, not in any review item. The old 2x4 version of this
section said "suppose the interaction estimate is about -1.2", which was the value baked into
the rnorm() call rather than the value the model returns. While fixing that I noticed nothing in
the chapter mentioned that the control group also improved, which seemed like a wasted teaching
moment in a chapter about interactions. Cut it if it distracts. -->

Notice that Control changed too, and significantly, $p$ `r ifelse(fixed_effects["occasionWeek 6", "Pr(>|t|)"] < .001, "< .001", paste("=", round(fixed_effects["occasionWeek 6", "Pr(>|t|)"], 3)))`. People in the control condition got better over six weeks. If we had run a one-group study, measured twice, and found that, we would have congratulated ourselves and possibly written a press release. This is the entire reason the control group exists, and the reason the interaction, not the change, is the research question.

::: {.callout-warning title="The p-Values Are Approximations, and This Is Not Your Engine's Fault"}
Every column in that table except the estimate itself rests on an approximation. There is no exact sampling distribution waiting for a model like this one, so the numbers in the `df` column were not counted, they were **estimated**, using Satterthwaite's method. They are free to come out fractional, and in models with more occasions they usually do.

<!-- ---Alex--- This callout is where the chapter's old Wald-z paragraph went (CH18-U01 asked
for exactly that shrink), and it is the only surviving mention of glmmTMB in the file. The dead
pointer to "the later inference chapter" went with it, which closes CH18-P01. -->

This is a property of mixed models, not of `lmer`. Change engines and the approximation changes costume rather than disappearing: `glmmTMB`, which you will meet in graduate school, fits the same model and reports Wald *z* statistics with no degrees of freedom at all. Same model, different plumbing, and the same honest uncertainty underneath.

So say which method you used, look at the confidence intervals, and do not treat a *p*-value from a mixed model as a measurement taken with calipers. "R produced a number and I felt hopeful" is not an inferential method.
:::

## How Much Did the Random Intercept Actually Buy You?

The paired *t*-test chapter ran its analysis wrong on purpose to show what pairing was worth. Do the same thing here. `ordinary_model` is still sitting there, fit by `lm()`, believing it met 200 unrelated people.

```{r ordinary-vs-mixed}
lm_fixed <- summary(ordinary_model)$coefficients

data.frame(
  term        = rownames(fixed_effects),
  estimate_lm = round(lm_fixed[, "Estimate"], 3),
  estimate_mixed = round(fixed_effects[, "Estimate"], 3),
  SE_lm       = round(lm_fixed[, "Std. Error"], 3),
  SE_mixed    = round(fixed_effects[, "Std. Error"], 3),
  row.names   = NULL
)
```

The estimates are **identical**, to every decimal place shown. Ignoring the pairing did not bias a single coefficient. That is worth saying plainly, because it is the part people get wrong: this is not a story about wrong answers.

The standard errors are a different matter. The interaction's standard error is `r round(lm_fixed["occasionWeek 6:groupCBT", "Std. Error"], 2)` when we pretend the rows are strangers and `r round(fixed_effects["occasionWeek 6:groupCBT", "Std. Error"], 2)` when we admit they are not, roughly `r round(lm_fixed["occasionWeek 6:groupCBT", "Std. Error"] / fixed_effects["occasionWeek 6:groupCBT", "Std. Error"], 1)` times larger. And look at what that does to the Control change, `occasionWeek 6`:

```{r lm-vs-mixed-p}
c(
  p_lm    = round(lm_fixed["occasionWeek 6", "Pr(>|t|)"], 3),
  p_mixed = round(fixed_effects["occasionWeek 6", "Pr(>|t|)"], 5)
)
```

Same estimate. One analysis finds nothing, the other finds a clear effect. The difference is entirely whether the model was told that the two rows belong to one human.

Now look at the row the random intercept did *not* help: `groupCBT`, where the standard error is the same in both models, out to nine decimal places. That is not a coincidence and it is worth understanding, because it tells you what the random intercept is actually for. `groupCBT` compares two different sets of people at one occasion. There is no repeated measurement inside that comparison, so there is nothing to subtract out, and it has to pay full price for person-to-person variation either way. The random intercept helps *within*-person comparisons. It has nothing to offer a between-person one.

Where did the extra error come from, then? From the people. `r round(as.data.frame(VarCorr(mixed_model))$vcov[1] / sum(as.data.frame(VarCorr(mixed_model))$vcov) * 100)` percent of the variance here is participants differing from one another, and `lm()` has no mechanism for setting that aside, so it lands in the error term and makes every within-person comparison look less trustworthy than it is. The random intercept is how you get it back.

## The Interaction Is Also a *t*-Test on Change Scores

With exactly two occasions we can go back to difference scores after all. Subtract baseline from Week 6 for every person, then compare those change scores between the two groups with an ordinary independent-samples *t*-test, which is a test from Part 1 of this book.

```{r change-score-test}
change_data <- tidyr::pivot_wider(
  mixed_data[, c("participant", "group", "week", "depression")],
  names_from   = week,
  values_from  = depression,
  names_prefix = "week"
)

change_data$change <- change_data$week6 - change_data$week0

# ref = "CBT" so the contrast runs CBT minus Control, matching the model
change_result <- t.test(
  change ~ relevel(group, ref = "CBT"),
  data = change_data,
  var.equal = TRUE
)

change_result
```

```{r change-score-identity}
b3 <- fixed_effects["occasionWeek 6:groupCBT", ]

change_comparison <- data.frame(
  method = c("t-test on change scores", "mixed-model interaction"),
  t      = c(unname(change_result$statistic), b3["t value"]),
  df     = c(unname(change_result$parameter), b3["df"]),
  p      = c(change_result$p.value,           b3["Pr(>|t|)"]),
  estimate = c(
    mean(change_data$change[change_data$group == "CBT"]) -
      mean(change_data$change[change_data$group == "Control"]),
    b3["Estimate"]
  )
)

print(change_comparison, digits = 6, row.names = FALSE)
```

One answer, printed twice, for the second time in this chapter.

So this chapter has now caught the mixed model impersonating two different *t*-tests. A paired *t*-test is a mixed model with a random intercept. A mixed 2-by-2 interaction is an independent-samples *t*-test on change scores. Neither is a coincidence, and neither is a reason to abandon the *t*-tests: in a complete, balanced 2-by-2 like this one they are easier to run and they are right.

The mixed model is what keeps working when the design stops being complete and balanced, which is the next thing that happens to it.

## Test the Interaction Directly

Coefficient tests are not the only way to ask about the interaction. We can also compare two models: one that lets the groups change differently, and one that does not.

When comparing **fixed effects**, refit both with maximum likelihood. This is the switch the REML callout promised, and this is where we throw it, in both models.

```{r compare-fixed-effects}
reduced_ml <- lmer(
  depression ~ occasion + group + (1 | participant),
  data = mixed_data,
  REML = FALSE
)

full_ml <- lmer(
  depression ~ occasion * group + (1 | participant),
  data = mixed_data,
  REML = FALSE
)

anova(reduced_ml, full_ml)
```

This likelihood-ratio test asks whether allowing the groups to change by different amounts improves model fit. It is not the only defensible test; coefficient tests and other approximations are also used. The important point is to match the test to the question and identify the method, not to collect p values like cursed trading cards.

::: {.callout-note title="`anova()` Would Have Caught Us, but Only `anova()`"}
Hand two REML fits to `anova()` and it refits them with maximum likelihood itself, printing `refitting model(s) with ML (instead of REML)` on its way past. So the `REML = FALSE` above is belt-and-braces: it makes the decision visible in your code instead of leaving it to a message you might scroll past.

The protection stops there, though. Ask for `AIC()` on two REML fits and nothing refits and nothing warns: you get a gap of 52.33 where the honest ML answer is 51.78. Subtract two REML log-likelihoods by hand and you get 54.33 where the correct likelihood-ratio statistic is 53.78. Neither is a disaster here, and both are wrong.

`anova()` is being the adult in the room. Do not assume the rest of the room joined in.
:::

After model comparison, report estimates from the final model fit with REML when appropriate.

## Follow Up the Interaction

The interaction says the two changes differ. It does not say what either change was. For that, estimate Week 6 minus baseline separately within each group:

```{r simple-effects}
library(emmeans)

cell_means <- emmeans(mixed_model, ~ occasion | group)

contrast(
  cell_means,
  method = list("Week 6 - Baseline" = c(-1, 1)),
  adjust = "none"
)
```

These are **simple effects**: how each group changed on its own. Control improved a little, CBT improved a lot, and the interaction is the statement that those two amounts are not the same. Simple effects describe the pattern. They do not replace the interaction test, and reporting only the simple effects is how people accidentally claim an interaction they never tested.

We can also ask for the four cell means the model implies:

```{r estimated-means}
cell_means
```

Do not automatically test every possible pair because the software makes it easy. That is how a sensible research question becomes 28 tests and a dead [salmon](Chapter_Multiple_Comparisons.qmd) [@bennett2010]. Choose comparisons that answer the theory, and adjust for multiplicity when you conduct a family of tests.

## Reporting the Result

A compact results paragraph could read:

> Depression scores were analyzed with a linear mixed-effects model containing fixed effects of Occasion, Group, and their interaction, plus a participant random intercept. The model was fit by restricted maximum likelihood using `lmer()` from the lme4 package, and *p*-values were obtained from lmerTest using Satterthwaite's approximation to the denominator degrees of freedom. The groups did not differ at baseline, $b =$ `r round(fixed_effects["groupCBT", "Estimate"], 2)`, $p =$ `r round(fixed_effects["groupCBT", "Pr(>|t|)"], 3)`. The Occasion by Group interaction was negative, $b =$ `r round(b3["Estimate"], 2)`, $SE =$ `r round(b3["Std. Error"], 2)`, $t(`r round(b3["df"], 1)`) =$ `r round(b3["t value"], 2)`, $p$ `r ifelse(b3["Pr(>|t|)"] < .001, "< .001", paste("=", round(b3["Pr(>|t|)"], 3)))`, indicating that depression declined more from baseline to Week 6 in the CBT condition than in the control condition.

A real report should also provide confidence intervals, the cell means, the exact random-effects structure, software, missing-data handling, and theoretically planned follow-ups. The paragraph above is the skeleton. Please give it organs before submitting it to Reviewer 2.

# Advanced Topics {.unnumbered}

That is the required chapter. What follows is the part that matters once the data are real rather than simulated: people miss appointments, models need checking, and two occasions eventually stop being enough. The Short Story at the end covers the whole chapter, both halves.

## Missing Repeated Observations

Mixed models can use participants who are missing some occasions; they do not require deleting every person with one missing row. This is a major advantage over older complete-case repeated-measures procedures, and with two occasions it is the difference between losing a person and losing half of one.

```{r missing-occasions}
set.seed(347)

mixed_missing <- mixed_data
# 10% of the Week 6 visits simply do not happen
gone <- mixed_missing$week == 6 & runif(nrow(mixed_missing)) < 0.10
mixed_missing$depression[gone] <- NA

missing_model <- lmer(
  depression ~ occasion * group + (1 | participant),
  data = mixed_missing
)

c(
  rows_used = sum(!is.na(mixed_missing$depression)),
  people_used = length(unique(
    mixed_missing$participant[!is.na(mixed_missing$depression)]
  ))
)
```

Every one of the 100 people is still in the model, contributing whatever they turned up for. A change-score analysis would have deleted all `r sum(gone)` of the people who missed the second visit, because a difference needs both halves, and with it the perfectly good baseline score each of them did provide.

However, "the model ran" does not mean missing data ceased to matter. Standard likelihood-based mixed-model inference commonly relies on a **missing at random** assumption conditional on variables in the model. If people with worsening symptoms leave the study for reasons the model does not capture, the missingness can bias results. Investigate dropout, include relevant predictors when justified, and conduct sensitivity analyses in serious work.

## Check the Model

At minimum, examine:

- whether the cell means are represented sensibly;
- residual-versus-fitted patterns;
- severe residual outliers;
- the distribution and influence of clusters;
- convergence and singular-fit warnings;
- whether variance changes across occasions or groups;
- whether enough participants, not merely enough rows, support the model.

```{r mixed-diagnostics, fig.width=6, fig.height=3.4}
#| fig-cap: "Residuals against fitted values, with a straight-line summary fit separately within each occasion. A well-behaved residual plot has no relationship at all, which is roughly what Baseline shows. Week 6 slopes upward, and that is not noise. It is the individual differences in rate of change that a two-occasion model is not allowed to estimate, sitting in the residuals instead."
#| fig-alt: "A scatterplot of residuals against fitted values, fitted values from about 10 to 51 and residuals from about -4.3 to +4.3, with points colored by occasion and a straight trend line drawn within each. Both clouds are wide and centered on zero, and a dashed red horizontal line marks zero. The Baseline trend line is nearly horizontal. The Week 6 trend line rises clearly from left to right."
diagnostic_data <- data.frame(
  fitted   = fitted(mixed_model),
  residual = resid(mixed_model),
  occasion = mixed_data$occasion
)

ggplot(diagnostic_data, aes(x = fitted, y = residual, color = occasion)) +
  geom_point(alpha = 0.35) +
  geom_hline(yintercept = 0, linetype = 2, color = "firebrick") +
  # a straight line, not a loess: the question here is whether there is a
  # linear trend, and a wiggly smoother invites you to over-read its ends
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Fitted value", y = "Residual", color = NULL) +
  theme_minimal(base_size = 11)
```

The two occasions are not telling the same story, so measure them separately rather than squinting at the lines:

```{r residual-trend-by-occasion}
round(
  c(
    Baseline = cor(diagnostic_data$residual[diagnostic_data$occasion == "Baseline"],
                   diagnostic_data$fitted[diagnostic_data$occasion == "Baseline"]),
    `Week 6` = cor(diagnostic_data$residual[diagnostic_data$occasion == "Week 6"],
                   diagnostic_data$fitted[diagnostic_data$occasion == "Week 6"])
  ),
  3
)
```

At baseline there is essentially no relationship between how high the model expects a score to be and how wrong it is, which is what a residual plot is supposed to look like. At Week 6 there is a clear one: the people the model expects to score high are also the people it under-predicts.

We know exactly what that is, because we put it in the simulation ourselves. Every person got their own `person_slope`, so these invented humans genuinely improve at different rates. At baseline that does nothing, because each person's rate of change has had zero weeks to act. By Week 6 it has had six, and the model has no random slope to keep it in, so it goes into the residual and drags it around. The residuals correlate `r round(cor(resid(mixed_model), mixed_data$person_slope * mixed_data$week), 2)` with each person's rate of change multiplied by week, which is the culprit holding a sign.

This is not something a better two-occasion model can fix, and the earlier callout is the reason why. It is the cost of the design, made visible. Report it, keep the model, and notice that the residual plot just told you what a third measurement occasion would buy.

Traditional repeated-measures ANOVA emphasizes **sphericity**, and with exactly two occasions that assumption is not merely satisfied, it is vacuous: there is only one difference, so there are no differences whose variances could disagree. Sphericity becomes a real problem at three occasions and up. A mixed model does not require that assumption at any number of occasions, but it does require a defensible covariance and random-effects structure. We did not escape assumptions. We traded a rigid one for a model that must be specified thoughtfully.

## When Two Occasions Are Not Enough

Everything in this chapter rested on each person appearing exactly twice. That is what made the difference score possible, the interaction a difference of differences, and the random slope unavailable.

Add a third occasion and all of it changes:

- There is no longer *the* difference, so change has to be described by a slope or by one estimate per occasion.
- The interaction becomes a difference between two **slopes** rather than between two changes.
- A random slope becomes estimable, so people can finally be allowed to change at different rates. They are humans, not synchronized microwave clocks.
- Sphericity starts to matter, and so does whether nearby occasions resemble each other more than distant ones.

That is a chapter of its own, and it is coming later in the book. Note what it is *not*: a reason to reach for complexity here. A random slope you cannot estimate is not a more sophisticated analysis. R will accept many bad ideas with the serene indifference of a vending machine, and this is one of the few times it declines.

## The Short Story

- A paired *t*-test is a mixed model with one two-level within-subjects predictor and a random intercept. Same estimate, same *t*, same df, same *p*.
- A mixed design contains at least one between-subjects predictor and one within-subjects predictor.
- Repeated occasions are nested within people, so ordinary independent-error regression is incomplete.
- Ignoring that nesting does not bias the coefficients. It ruins the standard errors, which is enough to change the answer.
- A mixed 2-by-2 has four cells and a difference-of-differences interaction, exactly like the all-between 2-by-2. The one new thing is the participant random intercept.
- With Baseline as the reference, the Group coefficient is the baseline group difference, which randomization should have made boring.
- The Occasion by Group coefficient is the difference between the two groups' changes.
- With exactly two occasions, that interaction is also an independent-samples *t*-test on change scores.
- Random intercepts let people start at different levels. Random *slopes* let them change at different rates, and they need at least three occasions to exist.
- Compare fixed-effects structures using models fit with maximum likelihood, not REML.
- Mixed models can retain partially observed participants, but missingness can still bias the answer.
- Never count rows and forget to count people. Two hundred observations from 100 people are not 200 independent humans, no matter how urgently the grant deadline approaches.

The model is "mixed" because it combines fixed and random effects, not because we placed several statistical procedures into a blender and hoped the lid stayed on.
