---
title: 'One Sample t-Test'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 4.5,
  fig.height = 3.5
)
```


## Experimental Logic

In psychology, we test claims by making comparisons.

Sometimes we compare a group to a known value: *Do more frequent train rides increase CTA liking above the neutral baseline of 3?* → one-sample t-test.

Sometimes we compare two groups: *Does CTA liking differ between people in quiet vs. regular subway cars?* → independent samples t-test.

Either way, the logic is the same: **are the differences we observe larger than we'd expect from random sampling noise alone?**

### Logic behind testing: The Blind Men and the Elephant

A group of blind men were traveling together when they encountered a strange object blocking their path. Each man touched a different part of the object and described what he thought it was, based entirely on his limited experience.

One man, feeling the elephant’s trunk, exclaimed, *“This must be a thick branch of a tree!”*
Another, grasping the elephant’s leg, disagreed: *“No, it’s clearly a pillar!”*
A third, holding the elephant’s ear, insisted, *“You’re both wrong—it’s obviously a fan!”*

Each man was convinced he was right, believing his own perspective represented the full truth. Their disagreement quickly escalated into an argument, with each trying to prove the others wrong. Eventually, the elephant, thoroughly unimpressed, grew irritated by their foolishness and tossed them all into a nearby river for disturbing its peace.

— **A paraphrased ancient Indian parable**

This story is a useful reminder when trying to understand the world, and especially when building theories from data. Our perspectives are almost always limited, shaped by where we stand, what we’ve experienced, and what we want to believe. Just as each blind man grasped only one part of the elephant, our samples and hypotheses capture only small pieces of a much larger reality.

Statistical testing exists precisely because of this limitation: *it gives us a disciplined way to ask whether our small, noisy glimpses are consistent with the bigger picture, or whether we might just be arguing over trunks, legs, and ears.*

### Experimental Example

-   *Step 1*: Begin with a theory.
-   *Step 2*: Develop a hypothesis derived logically from your theory.
-   *Step 3*: Design a method to test your hypothesis, including
    measurements, sample recruitment, and analytical tools.
-   *Step 4*: Collect data and compare it with your prediction.

When Step 4 fails, meaning the data don’t match the prediction, we face a classic scientific dilemma. It is often difficult to determine why things went wrong:

- Was the theory flawed?
- Was the hypothesis poorly specified?
- Or were the methods inadequate?

This problem is well known in psychology and has been discussed extensively [@meehl1978].

<!-- ALEX: these three headings ("Theory:", "Hypothesis:", "Important Note:") end in colons
     and read like slide titles, and the whole stretch from here down to "Logic of Hypothesis
     Test" is the driest run in the chapter. It is also where the cookie study gets
     introduced, which should be the fun part. I did not add jokes because CH06-B01 was
     scoped to structure only, but this is the stretch that most wants your voice. -->

### Theory:

Reinforcing good behavior in children makes them happy because it signals that they have pleased the adults around them. However, rewarding children with something they perceive as a forbidden treat, such as cookies, may have unintended consequences. Children know that cookies are usually off-limits and therefore associate them with rule-breaking. Receiving such a reward may overexcite them, pushing their self-control beyond its limits and leading to hyperactive behavior as they struggle to regulate their impulses.

### Hypothesis:

Based on this theory, we hypothesize that giving children cookies will cause them to lose control of their impulses and become hyperactive.

To test this hypothesis, we design a simple study in which children are rewarded with chocolate chip cookies and their behavior is observed.

### Important Note:

Regardless of the study’s outcome, it would not rule out other plausible explanations. For example, one alternative account is that sugar in cookies is rapidly metabolized, producing a temporary spike in blood glucose that leads to increased activity levels.

This is the core difficulty of experimental inference: a single result rarely tells us which explanation is correct. It only tells us which explanations are still plausible.

### Logic of Hypothesis Test

If X is true, then Y should also be true.

-   We assume X is true.
-   Therefore, we *expect* to observe Y.

However:

- If we observe Y, this does not necessarily confirm that X is true, because Y could result from other factors.

- If we do not observe Y, especially under rigorous and well-controlled conditions, this suggests that our theory may be flawed or incomplete.

Example:

-   Theory: Reinforcing good behavior in children with treats leads to increased hyperactivity because children perceive the treat as a forbidden reward, triggering an excited and poorly controlled response

-   Hypothesis: Giving children chocolate chip cookies as a reward will result in increased hyperactivity compared to children who do not receive such rewards.

-   Testing the Hypothesis:

    -   If children who receive cookies exhibit increased hyperactivity (Y is observed), this outcome is consistent with the theory but does not conclusively prove it, since alternative explanations (e.g., sugar content) could also account for the behavior.

        -   If children do not exhibit increased hyperactivity (Y is not observed), this challenges the theory, suggesting it may be incorrect or incomplete. At that point, we must re-examine the theory, the methods, and consider alternative explanations.

In hypothesis testing, **falsification** plays a central role. Supporting evidence increases our confidence in a theory, but evidence that contradicts our predictions undermines it. Scientific progress depends on continuously testing and attempting to falsify our theories, refining them through rigorous scrutiny and new evidence.

**4 steps for hypothesis testing**:

-   State the hypothesis
-   Set the Criteria Decision
-   Collect Data and Compute Sample Statistics
-   Make a Decision

### The Null Hypothesis ($H_0$)

- The null hypothesis ($H_0$) asserts that in the general population, there is no change, no difference, or no relationship.
    - In an experiment, $H_0$ predicts that the independent variable (e.g., treatment) will have no effect on the dependent variable for the population.
- $H_0$ is the hypothesis that is subject to a final decision: to either *reject* or *fail to reject*.

### The Scientific or Alternative Hypothesis ($H_1$)

- The alternative hypothesis ($H_1$) suggests that there is a change or a relationship in the general population.
    - In an experiment, $H_1$ predicts that the independent variable (e.g., treatment) will influence the dependent variable for the population.

### Directional Hypothesis

- **$H_0$**: Chocolate chip cookies do not make children hyper.
- **$H_1$**: Chocolate chip cookies do make children hyper.

Written symbolically, where $\mu$ is the population mean hyperactivity score and 50 is the value we expect without cookies:

- $H_0: \mu \le 50$
- $H_1: \mu > 50$

### Non-directional Hypothesis

- **$H_0$**: Chocolate chip cookies do not change children's level of hyperactivity.
- **$H_1$**: Chocolate chip cookies do change children's level of hyperactivity.

Written symbolically:

- $H_0: \mu = 50$
- $H_1: \mu \neq 50$

<!-- ALEX: the symbolic lines are new (CH06-U01). They are deliberately plain, since the
     exam writes them this way and this is the one spot where the book should look like the
     exam. If you want a joke, the space just under the non-directional pair is where it
     goes, not inside the notation. -->

*Note: The default in psychology is to choose the non-directional test, so we will only talk about that version for the rest of below*

### Set the Decision Criteria

- The decision criteria depend on the type of test, the number of tests conducted, the sample size you have, and whether the hypothesis is directional or non-directional.
- A key decision is determining how far sample means need to be from the population mean to reject the null hypothesis ($H_0$).
- The challenge is identifying what constitutes abnormal behavior in the population.
  - Should we consider deviations in both tails, or just one tail (either lower or higher)?
  - Where does the tail begin, and the body of the distribution end?
- To address this, we set a cut-off point, known as the Alpha level ($\alpha$).
  - Common Alpha levels are: $\alpha = .05$, $\alpha = .01$, $\alpha = .005$, $\alpha = .001$, $\alpha = .0001$.
- The Alpha level represents the threshold for what we consider "abnormal," meaning the area under the curve beyond this cut-off.

### Accounting for Sample Size: The Transition to the $t$-Distribution

We also have to account for the sample size of our study, since in small samples, we know we will not get the best estimate of the population numbers (as we saw above). So, we apply a correction by moving from the standard Normal ($z$) distribution to the Student's $t$-distribution.

::: {.callout-note title="Student's t-Test Was Created Because of Beer"}
The word *Student* does not refer to an undergraduate who invented a test between classes. It was the pen name of William Sealy Gosset, a chemist and brewer at Guinness. Guinness used small samples to study ingredients and production, and company policy restricted employees from publishing their work. Gosset therefore published *The Probable Error of a Mean* as "Student" in 1908 [@student1908; @lehmann1999student].

In other words, one of the most famous tools in statistics exists because a brewer wanted to make better beer without pretending a tiny sample was a giant one. Statistics has rarely produced a more convincing origin story.
:::

### Why the $t$-Distribution?

- Estimation of Variance: we will never know the population standard deviation ($\sigma$), so we must estimate it using the sample standard deviation ($S$).
- The Penalty for Small Samples: Small samples are more prone to extreme fluctuations. Because our estimate of the population variance is less reliable when $n$ is small, the $t$-distribution is "heavier" in the tails than the normal distribution.
- Degrees of Freedom ($df$): The shape of the $t$-distribution is determined by the degrees of freedom, calculated as $df = n - 1$.
  - As $n$ (and thus $df$) increases, the $t$-distribution approaches the shape of the standard normal distribution.
  - As $n$ decreases, the tails become "fatter," requiring a larger critical value to reach the same alpha level ($\alpha$).

### Adjusting the Critical Region

Because the $t$-distribution has more area in its tails, the "cut-off" points for our Alpha level move further away from the mean. This means:

- It is harder to reject the null hypothesis with a small sample.
- We are being mathematically conservative to account for the increased risk of error when we don't have a large amount of data.

The $t$-distribution acts as a "safety buffer." It ensures that we only label a result as "statistically significant" if the effect is strong enough to overcome the inherent noise of a small sample size.

### Determining the Cut-Off Points for Two-Tailed Tests

For a two-tailed test with Alpha = 0.05, the cut-offs depend on sample size. For our study with 36 kids that would look like this: 

```{r t-critical-regions}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The t-distribution for df = 35. The shaded red areas represent the critical regions where we reject the null hypothesis."
#| fig-alt: "A symmetric bell-shaped t-distribution curve centered on zero. The two tails beyond about negative and positive two are shaded red, and dashed vertical lines mark those cut-offs."

library(ggplot2)

# Define parameters
n <- 36
df <- n - 1
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df)

# Generate Plot
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  # Draw the t-distribution curve
  stat_function(fun = dt, args = list(df = df), color = "black", linewidth = 1) +
  # Shade the lower tail
  stat_function(fun = dt, args = list(df = df), 
                xlim = c(-4, -t_crit), geom = "area", fill = "red", alpha = 0.5) +
  # Shade the upper tail
  stat_function(fun = dt, args = list(df = df), 
                xlim = c(t_crit, 4), geom = "area", fill = "red", alpha = 0.5) +
  # Add critical value lines
  geom_vline(xintercept = c(-t_crit, t_crit), linetype = "dashed", color = "darkred") +
  labs(title = paste0("t-Distribution Rejection Regions (df = ", df, ")"),
       subtitle = "Alpha = .05, Two-Tailed",
       x = "t-statistic",
       y = "Density") +
  theme_bw()

# Save values for inline code
t.cut.lower <- -t_crit
t.cut.upper <- t_crit
```


```{r}
#| echo: false
#| message: false
# Define parameters
n <- 36
df <- n - 1
alpha <- 0.05

# Calculate t-cutoffs
t_lower <- qt(alpha / 2, df)
t_upper <- qt(1 - (alpha / 2), df)
```

Our cut-off ($t_{critical}$) values for $df = 35$ would be:

- $t_{lower}$ < `r round(t_lower, 3)` 
- $t_{upper}$ > `r round(t_upper, 3)`

### The Impact of Sample Size on Critical Regions

If we were to increase our sample size from $n = 36$ to $n = 100$, our degrees of freedom would increase to $df = 99$.

As $n$ increases, our confidence in our estimate of the population standard deviation also increases. Mathematically, the $t$-distribution begins to pull its "fat tails" inward, looking more and more like the standard Normal ($z$) distribution.

```{r t-dist-sample-size}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Two t-distributions overlaid. The larger sample has thinner tails, so its cut-offs sit closer to the middle."
#| fig-alt: "Two overlapping bell-shaped curves centered on zero. The red curve for 36 people has slightly fatter tails than the blue curve for 100 people, and dotted vertical lines show the blue cut-offs sitting just inside the red ones."

# Define parameters for n = 100
n_large <- 100
df_large <- n_large - 1
t_crit_large <- qt(1 - alpha/2, df_large)

# Generate comparison plot
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dt, args = list(df = 35), aes(color = "n = 36 (df = 35)"), linewidth = 1) +
  stat_function(fun = dt, args = list(df = 99), aes(color = "n = 100 (df = 99)"), linewidth = 1) +
  geom_vline(xintercept = c(-t_upper, t_upper), linetype = "dotted", color = "red") +
  geom_vline(xintercept = c(-t_crit_large, t_crit_large), linetype = "dotted", color = "blue") +
  scale_color_manual(values = c("n = 36 (df = 35)" = "red", "n = 100 (df = 99)" = "blue")) +
  labs(title = "Distribution Shift: n = 36 vs. n = 100",
       subtitle = "Note how the boundaries move inward as sample size increases",
       x = "t-statistic", y = "Density", color = "Sample Size") +
  theme_bw()
```

New Cut-Off ($t_{critical}$) Points ($n = 100$):

- $t_{lower}$ < `r round(-t_crit_large, 3)`
- $t_{upper}$ > `r round(t_crit_large, 3)`

Why the boundaries are smaller: with 100 children instead of 36, our sample mean is a much more reliable "anchor" for the population mean. We don't need as much of a "safety buffer."

Here is how the boundaries shift with different sample sizes:

```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
library(kableExtra)

# Create the data
comparison_data <- data.frame(
  Sample_Size = c(10, 36, 100, "Infinity (z)"),
  DF = c(9, 35, 99, "N/A"),
  Critical_Value = c(
    round(qt(0.975, 9), 3),
    round(qt(0.975, 35), 3),
    round(qt(0.975, 99), 3),
    1.960
  )
)

# Render the table
kable(comparison_data, 
      col.names = c("Sample Size (n)", "Degrees of Freedom (df)", "Absolute Critical Value"),
      caption = "How Sample Size Changes the Threshold for Significance") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Collect Data and Compute Sample Statistics

- Get a sample of children, $n = 36$, and feed them a bunch of cookies! We collected that data and in R stored their hyperactivity scores in an object called `kids.1`
- **Dependent Variable (DV):** We will measure the children's hyperactivity, which we expect the cookies to raise.

Here is a plot of kids.1:

```{r kids-data}
#| echo: false
set.seed(123)
N <- 1e6
Population <- rnorm(N, mean = 50, sd = 5)
n <- 36

set.seed(42)
kids.sample <- sample(Population, n, replace = TRUE)

# every kid gets a 2-point cookie boost (the true effect)
kids.1 <- kids.sample + 2
```

```{r kids-histogram}
#| echo: false
#| fig-cap: "Hyperactivity scores for our 36 cookie-fed children."
#| fig-alt: "Histogram of 36 hyperactivity scores, roughly mound-shaped and centered in the low fifties, with values spanning from the mid forties to the low sixties."
hist(kids.1,
     main = "Histogram",
     xlab = "Hyperactivity Score", ylab = "Frequency")
```

### Sample Results

The children who received cookies showed a mean of $M = `r round(mean(kids.1), 3)`$ with a standard deviation of $SD = `r round(sd(kids.1), 3)`$. Based on prior theory and past research, we  hypothesize that the population mean is $\mu = 50$

We can then ask whether the data we observed are consistent with that claim, or whether the difference between the sample mean and the hypothesized population mean is larger than we would expect from random sampling error alone.

### Compute Sample Statistics

The test statistic is computed using the following formula:

$$
t_{test} = \frac{M - \mu}{S_M}
$$

where the denominator is the standard error of the mean:

$$
S_M = \frac{s}{\sqrt{n}}
$$

We can do the math by hand, or use the t-test function in R. 
t-test by hand: 

```{r}
M = mean(kids.1)
s = sd(kids.1)
n = 36
Mu = 50 # given to you from the population
t.kids <- (M - Mu) / (s/sqrt(n))
t.kids
```

one sample t-test in R:

```{r}
Kids.t.test <- t.test(x = kids.1,
                      alternative = "two.sided",
                      mu = 50,
                      conf.level = 0.95)  # .95 is the default; shown here so you can see it
```


### Test Statistic Result

We get a $t_{test} =$ `r round(Kids.t.test$statistic, 3)`. Now, the question is: **does this value fall in the tail of the population distribution**?

Let's check our critical value.

Our $t_{critical}$ values for $df = 35$ would be:

- $t_{lower}$ < `r round(t_lower, 3)` 
- $t_{upper}$ > `r round(t_upper, 3)`

This t-test value is greater than our upper value. 

Further from the R t-test function, we can see our exact p-value which is well below $p < .05$.

::: {.callout-important title="The p-Value, Actually Defined (That IOU from Chapter 1)"}
Back in Chapter 1 I promised you a real definition of statistical significance once you had some probability theory. Here it is, at the first moment you actually need it.

**The p-value is the probability of getting a test statistic at least this far from the null value, *if* the null hypothesis and the model's assumptions were true.**

Look back at the shaded red tails in the figure above. Your observed $t$ lands somewhere on that curve, and the p-value is the area past it in both directions. Small p means your data sit far out in a region the null model rarely produces.

Note what the definition does not say. It is a statement about the data given an assumption, not a statement about the assumption given the data. It is not the probability that the null hypothesis is true, not the probability that you are wrong, and not the probability that the result will replicate.
:::

We can also see the 95 percent confidence interval from our R t-test function, and that range does NOT include $\mu = 50$, which is another way of seeing that our result is bigger than chance alone would produce.

### Make a Decision

There are two possible outcomes:

- **Reject the null hypothesis:** This decision is reached when the data fall within the critical region.
    - This indicates the sample mean is farther from the hypothesized value than random sampling error can comfortably explain. Whether the *cookies* did it is a design question, not a statistics question. See the Important Note above.
    - **We reject the null** because it is generally easier to demonstrate that something is false than to prove it true. (For the other way of weighing evidence, see the Bayes section of the [Probability chapter](Chapter_Probability.qmd).)

- **Fail to reject the null hypothesis:** If the data do not provide strong evidence (i.e., they do not fall within the critical region), then the treatment is *assumed* to have no effect.

### Decision Outcome

**YES, our $t_{test}$ > $t_{critical}$, so we REJECT the null hypothesis.**

- We report: $t_{test}$ = `r round(Kids.t.test$statistic, 3)`, $p < .001$ (since we know the exact p-value).
- In modern APA format, it is recommended to report the exact p-value.

### One-sample t-test in APA format

<!-- ALEX: this used to be library(apa) with #apa(Kids.t.test) commented out, and the
     "what is the d thing" line pointed at output nobody could see (CH06-P01). I found out
     why you commented it: the apa package refuses one-sample tests when mu is not 0
     ("cohens_d does currently not support t-tests with mu != 0"). So option (a) in the
     review was never going to work without subtracting 50 first, which would wreck the
     mu = 50 teaching point. I took option (b): built the line by hand, dropped the
     dependency, kept your hook. -->

Written the way it would appear in a paper, our result looks like this:

> $t(35) =$ `r round(unname(Kids.t.test$statistic), 2)`, $p$ `r ifelse(Kids.t.test$p.value < .001, "< .001", paste("=", round(Kids.t.test$p.value, 3)))`, 95% CI [`r round(Kids.t.test$conf.int[1], 2)`, `r round(Kids.t.test$conf.int[2], 2)`], $d =$ `r format(round((mean(kids.1) - 50) / sd(kids.1), 2), nsmall = 2)`

Wait, what is the $d$ thing at the end? The problem with significance testing is it does not tell you how "big" the effect is. It just tells you the result is probably not due to chance.

## Effect Sizes

I need a way to see how "big" an effect might be. Cohen borrowed a useful idea at the time, signal detection theory, developed for radar operators in World War II who had to decide whether that blip was a plane or noise. The idea was that you want to know how strong a signal needed to be for you to detect it over the noise. The noise for us is the control group and the signal the effect of your treatment.

```{r signal-detection}
#| echo: false
#| fig-width: 6.25
#| fig-height: 3.0
#| fig-cap: "Signal detection. The distance between the noise curve and the signal-plus-noise curve, measured in standard deviations, is d."
#| fig-alt: "Two overlapping bell curves. The left curve is labeled N for noise and the right curve is labeled N plus S for signal plus noise. A double-headed blue arrow between their peaks is labeled d."
# modifed from http://rpsychologist.com/creating-a-typical-textbook-illustration-of-statistical-power-using-either-ggplot-or-base-graphics
library(ggplot2)
library(grid) # for arrow()

m1 <- 0    # mu H0
sd1 <- 1.5 # sigma H0
m2 <- 3.5  # mu HA
sd2 <- 1.5 # sigma HA

z_crit <- qnorm(1 - (0.05/2), m1, sd1)

# set range
min1 <- m1 - sd1*4
max1 <- m1 + sd1*4
min2 <- m2 - sd2*4
max2 <- m2 + sd2*4

x <- seq(min(min1, min2), max(max1, max2), .01)

# distributions
df1 <- data.frame(
  x = x,
  y = dnorm(x, m1, sd1),
  id = "h0"
)

df2 <- data.frame(
  x = x,
  y = dnorm(x, m2, sd2),
  id = "h1"
)

poly <- rbind(df1, df2)

ggplot(poly, aes(x, y, group = id)) +
  geom_polygon(alpha = 0.8, show.legend = FALSE) +
  geom_line(data = df1,
            aes(x, y, color = "H0"),
            linewidth = 1.5,
            show.legend = FALSE) +
  geom_line(data = df2,
            aes(x, y, color = "HA"),
            linewidth = 1.5,
            show.legend = FALSE) +
  scale_color_manual(
    values = c("HA" = "#981e0b", "H0" = "black")
  ) +
  annotate("segment",
           x = m1 + .4, y = 0.265,
           xend = m2 - .4, yend = 0.265,
           color = "blue",
           arrow = arrow(length = unit(0.3, "cm")),
           linewidth = 1) +
  annotate("segment",
           x = m2 - .4, y = 0.265,
           xend = m1 + .4, yend = 0.265,
           color = "blue",
           arrow = arrow(length = unit(0.3, "cm")),
           linewidth = 1) +
  annotate("text", label = "d", x = (m1 + m2)/2, y = 0.25, size = 8) +
  annotate("text", label = "N", x = m1, y = 0.28, size = 8) +
  annotate("text", label = "N + S", x = m2, y = 0.28, size = 8) +
  ggtitle("Signal Detection") +
  theme(
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "#ffffff"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(size = 22)
  )
```

$$ d = \frac{\mu_{(Noise+Signal)}-\mu_{(Noise)}}{\sigma_{(Noise)}} $$
Note this implicit assumption: $$\sigma_{(Noise)} = \sigma_{(Noise+Signal)}$$

### Cohen's $d$

Cohen surmised that we could apply the same logic to experiments to determine how big the effect would be (in other words how easy it is to distinguish the signal from the noise). 

$$d = \frac{M_{H1}-\mu_{H0}}{S_{H1}} $$

Notice the denominator changed. The signal detection version divides by $\sigma_{(Noise)}$; here we divide by the sample's own SD. We do that because it is the only SD we have, and under the equal-spread assumption above it is also our best guess at the noise.

So Cohen's $d$ is in standard deviation units:

Size   | $d$
------ | --------
Small  | .2
Medium | .5
Large  | .8
Huge   | 1.2
              
### Small Visualized



```{r effect-small}
#| echo: false
#| fig-width: 4.25
#| fig-height: 2.0
#| fig-cap: "A small effect, d = .2. The two curves are nearly on top of each other."
#| fig-alt: "Two bell curves that overlap almost completely, with their peaks barely separated."
# modified from http://rpsychologist.com/creating-a-typical-textbook-illustration-of-statistical-power-using-either-ggplot-or-base-graphics
library(ggplot2)
library(grid) # for arrow()

plot_signal <- function(
  m2,
  title,
  m1 = 0,
  sd1 = 1,
  sd2 = 1,
  alpha = 0.8
) {

  # set range
  min1 <- m1 - sd1*4
  max1 <- m1 + sd1*4
  min2 <- m2 - sd2*4
  max2 <- m2 + sd2*4

  x <- seq(min(min1, min2), max(max1, max2), 0.01)

  df1 <- data.frame(
    x = x,
    y = dnorm(x, m1, sd1),
    id = "h0"
  )

  df2 <- data.frame(
    x = x,
    y = dnorm(x, m2, sd2),
    id = "h1"
  )

  poly <- rbind(df1, df2)

  ggplot(poly, aes(x, y, group = id)) +
    geom_polygon(alpha = alpha, show.legend = FALSE) +
    geom_line(
      data = df1,
      aes(x, y, color = "H0"),
      linewidth = 1.5,
      show.legend = FALSE
    ) +
    geom_line(
      data = df2,
      aes(x, y, color = "HA"),
      linewidth = 1.5,
      show.legend = FALSE
    ) +
    scale_color_manual(values = c("HA" = "#981e0b", "H0" = "black")) +
    ggtitle(title) +
    theme(
      panel.grid = element_blank(),
      panel.background = element_blank(),
      plot.background = element_rect(fill = "#ffffff"),
      panel.border = element_blank(),
      axis.line = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      plot.title = element_text(size = 22)
    )
}
plot_signal(0.2, "Small Effect")
```

### Medium Visualized

```{r effect-medium}
#| echo: false
#| fig-width: 4.25
#| fig-height: 2.0
#| fig-cap: "A medium effect, d = .5."
#| fig-alt: "Two bell curves with clearly separated peaks but still a large shared area in the middle."
plot_signal(0.5, "Medium Effect")
```

### Large Visualized

```{r effect-large}
#| echo: false
#| fig-width: 4.25
#| fig-height: 2.0
#| fig-cap: "A large effect, d = .8."
#| fig-alt: "Two bell curves whose peaks are well apart, though the curves still overlap substantially through the middle."
plot_signal(0.8, "Large Effect")
```

### Huge Visualized

```{r effect-huge}
#| echo: false
#| fig-width: 4.25
#| fig-height: 2.0
#| fig-cap: "A huge effect, d = 1.2. Even here the curves share a lot of ground."
#| fig-alt: "Two bell curves separated by more than one standard deviation, yet a visible region in the middle still belongs to both."
plot_signal(1.2, "Huge Effect")
```

You notice that even in a huge effect there is a lot of overlap between the curves. 

##  APA Style Report

Children who received cookies showed higher hyperactivity scores ($M =$ `r round(mean(kids.1), 2)`, $SD =$ `r round(sd(kids.1), 2)`) than the hypothesized population value ($\mu = 50$), $t(35) =$ `r round(unname(Kids.t.test$statistic), 2)`, $p$ `r ifelse(Kids.t.test$p.value < .001, "< .001", paste("=", round(Kids.t.test$p.value, 3)))`, 95% CI [`r round(Kids.t.test$conf.int[1], 2)`, `r round(Kids.t.test$conf.int[2], 2)`], $d =$ `r format(round((mean(kids.1) - 50) / sd(kids.1), 2), nsmall = 2)`. The standardized difference describes magnitude; whether the effect is practically meaningful requires context and better evidence than our imaginary cookie operation.

## Uncertainty and Error in Hypothesis Testing

- Hypothesis testing is an inferential process because we do **not** know the actual population data, and our samples could be in error.
- When we infer, we can make at least two types of errors (though there are others).

### Type I Error

- Also called a "false alarm" or "alpha error."
- This occurs when a researcher rejects a null hypothesis that is actually **true**. In other words, we reject the null even though the treatment does **nothing**.
- **Example:** Testing zinc or ColdEase on colds using self-report as a measure of improvement. Colds get better on their own, and people who paid for zinc lozenges want to believe they helped. So the study can happily "detect" an improvement that is really just the cold ending on schedule plus some hopeful self-report. The null was true and we rejected it anyway.
- Before observing the data, the $\alpha$ level sets the long-run Type I error rate for the testing procedure when the null hypothesis and its assumptions hold.
- Choosing $\alpha=.05$ does not mean there is a 5% probability that this particular rejected null hypothesis is true. The p-value cannot read the hidden state of the universe.

### Type II Error

- Also called a "miss" or "beta error."
- A Type II error occurs when a researcher fails to reject the null hypothesis that is actually **false**.
- In other words, a Type II error occurs when the treatment has an effect, but it seems like the treatment has had no effect because the hypothesis test fails to detect it.
- This can occur when the treatment has an effect but the study has too little power to detect it. That happens when the effect is modest, the measurement is noisy, or the sample is small.
- **Example:** Taking kava tea to reduce stress. Kava may genuinely calm people down a little. But give it to ten stressed undergraduates, measure them with a mood scale they fill out between classes, and the effect is too quiet to hear over the noise. The null was false and we failed to reject it.

::: {.callout-note title="The Courtroom Version, Which Your TA Will Use"}
A defendant is presumed innocent, which is the null hypothesis. The jury either convicts or fails to convict.

- **Type I error:** convicting an innocent person. You rejected a true null.
- **Type II error:** acquitting a guilty person. You failed to reject a false null.

Notice that "not guilty" is not the same verdict as "innocent." The jury is saying the evidence was not strong enough, which is exactly what "fail to reject" means. This is also why raising the standard of proof trades one error for the other. Make conviction harder and fewer innocent people go to prison, while more guilty ones walk.

You met the extreme version of this in the [Probability chapter](Chapter_Probability.qmd), where we fired the doctors and let a 95% accurate test convict people Judge Dredd style. It got 84% of them wrong.
:::

| Your decision | Treatment does not work ($H_0$=TRUE) | Treatment does work ($H_0$=FALSE) |
|-----------------|-----------------------------|---------------------------|
| Reject $H_0$  | Type 1                               | Correct Decision                  |
| Fail to reject $H_0$  | Correct Decision                     | Type II                           |

### Does our Alpha Set the Error/Miss Level?

<!-- ALEX: this section is one bullet answering its own heading with "it depends," which is
     the last thing a reader wants the night before Exam 1. Either it earns two sentences
     (alpha fixes the Type I rate by decree; beta is whatever the design, effect size, and
     n leave you with, which is why you cannot just set it) or it folds into the Power
     chapter pointer. Your call. -->

- The answer is: it depends. We will revisit this issue when we discuss *statistical power*.

### The Short Story

- A one-sample t-test compares a sample mean with a hypothesized population value.
- The t statistic is the observed difference divided by its standard error.
- A p-value measures compatibility with the null model; it is not the probability that the null is true.
- “Fail to reject” does not mean “prove no effect.” Sometimes the study simply arrived carrying a tiny sample and no flashlight.
- Report the estimate, confidence interval, p-value, and effect size. The asterisk cannot do all the talking.

::: {.callout-important title="The t Statistic Is Signal Divided by Noise"}
The numerator is how far the sample mean sits from the null value. The denominator is the standard error of that mean. A large t statistic can come from a large difference, a precise estimate, or both, so report the effect and its uncertainty, not merely the ceremonial p-value.
:::

::: {.callout-note title="Which t-Test Comes Next?"}
Use the one-sample t-test when one group is being compared with one number. If you want to ask whether **Group A differs from Group B**, continue to the [independent-samples t-test chapter](Chapter_Independent_t_test.qmd). For example, you might compare the running speed of one group of undergraduates chased by a goose with a completely different group chased by two geese. The ethics board will compare your application with a paper shredder.

If you measure the **same people twice**, continue to the [paired-samples t-test chapter](Chapter_Paired_t_test.qmd). For example, you might measure the same students' stress before and after telling them that the group presentation they forgot about has been moved to today. Same humans, two measurements, one sudden desire to transfer universities.
:::
