---
title: "Advanced: Bootstrapping, the Jackknife, and Resampling"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

## Before the Bootstrap: Why We Needed Resampling

Many familiar statistics come with a theoretical sampling distribution. If the assumptions behave, we can use that distribution to calculate a standard error, confidence interval, or test statistic. Unfortunately, statistics such as medians, indirect effects, complicated effect sizes, and estimates from strange models may not arrive with a pleasant little formula attached.

The resampling solution is wonderfully rude: make the computer imitate repeated sampling. We use the data we observed to create many slightly different samples, calculate the statistic in every sample, and watch how much it moves. The movement gives us information about uncertainty.

This idea sounds obvious now because your laptop can perform thousands of resamples before you finish complaining about the assignment. It was less obvious when a computer occupied a room, demanded punch cards, and had less memory than a suspicious toaster.

::: {.callout-warning title="Advanced Does Not Mean Assumption-Free"}
Resampling replaces some mathematical work with computation. It does not remove the need to understand the design, the estimator, or the assumptions. If you resample the wrong unit 10,000 times, you obtain a very stable estimate of the wrong uncertainty.
:::

## A Very Short History of People Deleting and Reusing Data

Resampling did not appear from a statistical volcano fully formed.

- **The jackknife:** Quenouille developed leave-one-out calculations to study bias. Tukey extended the method to variance estimation and gave it the name *jackknife*: a general-purpose statistical tool that could be carried around and used when needed.
- **The bootstrap:** Efron's 1979 paper was titled *Bootstrap Methods: Another Look at the Jackknife*. The name referred to pulling oneself up by one's bootstraps: using the sample itself to learn about repeated samples when the population is unavailable.
- **The computer changes everything:** The jackknife was designed when computation was expensive. The bootstrap became practical because computers could repeatedly resample and recalculate statistics. By 2003, Efron described the bootstrap as a method with roots in jackknife work on bias and variance, but with a much wider reach [@efron2003].

The history matters because the jackknife and bootstrap are not two unrelated buttons in R. They answer the same basic question in different ways: **how sensitive is our estimate to the particular observations we happened to collect?**

## The Jackknife: Remove One Person at a Time

Suppose our sample has $n$ observations and the statistic from all observations is $\widehat\theta$. The jackknife removes observation $i$, calculates the statistic again, and calls the result $\widehat\theta_{(i)}$. After doing this for every observation, we have $n$ leave-one-out estimates.

The average leave-one-out estimate is

$$
\overline{\theta}_{(.)}=\frac{1}{n}\sum_{i=1}^{n}\widehat\theta_{(i)}.
$$

The jackknife estimate of bias is

$$
\widehat{Bias}_{jack}=(n-1)
\left(\overline{\theta}_{(.)}-\widehat\theta\right),
$$

and the jackknife standard error is

$$
SE_{jack}=\sqrt{
\frac{n-1}{n}
\sum_{i=1}^{n}
\left(\widehat\theta_{(i)}-\overline{\theta}_{(.)}\right)^2
}.
$$

That standard error asks how much the estimate changes when each person temporarily vanishes. This is a statistical disappearance, not a proposal for managing annoying group members.

### A Jackknife Example by Hand

Suppose 12 students report how many minutes they delayed before opening an assignment. One student reports 190 minutes because procrastination became a full-time position.

```{r jackknife-data}
Delay.Small <- c(12, 18, 21, 23, 27, 31, 34, 38, 42, 49, 65, 190)

OriginalMean <- mean(Delay.Small)
JackMeans <- sapply(seq_along(Delay.Small), function(i) {
  mean(Delay.Small[-i])
})

JackMean <- mean(JackMeans)
JackBias <- (length(Delay.Small) - 1) * (JackMean - OriginalMean)
JackSE <- sqrt(
  (length(Delay.Small) - 1) / length(Delay.Small) *
    sum((JackMeans - JackMean)^2)
)

c(
  OriginalMean = OriginalMean,
  JackknifeBias = JackBias,
  JackknifeSE = JackSE
)
```

```{r jackknife-plot, fig.width=7, fig.height=4.4}
plot(
  seq_along(JackMeans), JackMeans,
  pch = 19, col = "#0072B2",
  xlab = "Observation removed",
  ylab = "Mean after removing that observation",
  main = "The jackknife exposes influential observations"
)
abline(h = OriginalMean, lwd = 3, col = "#D55E00")
text(12, JackMeans[12], labels = "190-minute student removed", pos = 3, cex = .8)
```

The last leave-one-out estimate moves far more than the others. The jackknife therefore does more than generate a standard error. It also shows which observations are carrying the estimate around like exhausted undergraduate research assistants.

### Pseudovalues and Bias Correction

The jackknife pseudovalue for observation $i$ is

$$
PV_i=n\widehat\theta-(n-1)\widehat\theta_{(i)}.
$$

The mean of the pseudovalues is the jackknife bias-corrected estimate:

$$
\widehat\theta_{jack}=n\widehat\theta-(n-1)\overline\theta_{(.)}.
$$

Pseudovalues transform the leave-one-out results so they behave more like observation-level contributions to the estimate. You will encounter them in older methods and in some specialized modern procedures. You do not need to place them under your pillow and hope they become intuitive overnight.

### Where the Jackknife Struggles

The jackknife works especially well for smooth statistics such as means, regression coefficients, and correlations. It can behave badly for statistics that change abruptly when a case is removed. The sample median is the famous example [@singhxie2008].

```{r jackknife-median}
OriginalMedian <- median(Delay.Small)
JackMedians <- sapply(seq_along(Delay.Small), function(i) {
  median(Delay.Small[-i])
})

c(
  OriginalMedian = OriginalMedian,
  UniqueJackknifeMedians = length(unique(JackMedians)),
  JackknifeSE = sqrt(
    (length(Delay.Small) - 1) / length(Delay.Small) *
      sum((JackMedians - mean(JackMedians))^2)
  )
)
```

Removing one observation may leave the median unchanged or make it jump between only a few values. That choppy behavior can make the leave-one-out approximation poor. The bootstrap usually gives the median more room to wander because each resample can omit several observations and repeat several others.

## The Bootstrap: Reuse the Sample with Replacement

Bootstrapping asks what would happen if we repeatedly sampled from a population resembling the one we observed. Because we do not possess the population, the **empirical distribution** of the sample stands in for it. Each observed case receives probability $1/n$, and we draw a new sample of size $n$ **with replacement** [@efron1993; @efron2003].

With replacement means that after selecting a participant, we return that participant before the next draw. Some participants appear several times; others disappear entirely. This is statistically acceptable and ethically preferable to cloning the undergraduates.

### The Plug-In Principle

Let $F$ represent the unknown population distribution and let $\widehat F$ represent the empirical distribution created from our sample. If the population quantity is

$$
\theta=t(F),
$$

we estimate it by plugging in the empirical distribution:

$$
\widehat\theta=t(\widehat F).
$$

The bootstrap then draws $F_1^*,F_2^*,\ldots,F_B^*$ from $\widehat F$ and calculates

$$
\widehat\theta_b^*=t(F_b^*)
$$

for each bootstrap sample $b$. This is the **plug-in principle**: replace the unknown population with the best estimate we have of it, then study what repeated sampling from that estimate does [@efron2003].

::: {.callout-note title="What the Bootstrap Treats as the Population"}
The observed sample temporarily plays the role of the population. If the sample is biased, excludes important groups, or was collected with a design that we ignore, bootstrapping reproduces that problem thousands of times with impressive efficiency.
:::

### Why About 63.2% of Cases Appear

In one bootstrap sample, the probability that a particular case is never selected is

$$
\left(1-\frac{1}{n}\right)^n.
$$

As $n$ grows, this approaches $e^{-1}\approx .368$. Therefore, a typical bootstrap sample contains about $1-.368=.632$, or 63.2%, of the original cases at least once. The remaining positions are duplicates. Nothing is missing from R; repetition is the entire point.

## Bootstrapping a Skewed Median

Suppose students report how many minutes they procrastinated before opening an assignment. The distribution is strongly right-skewed because several students began procrastinating during the previous semester.

```{r bootstrap-data}
set.seed(343)
Delay <- round(rlnorm(55, meanlog = log(35), sdlog = .8))
median(Delay)
mean(Delay)
```

We can show one bootstrap sample before asking R to perform the process thousands of times.

```{r one-bootstrap-sample}
set.seed(344)
OneBootstrapSample <- sample(
  Delay,
  size = length(Delay),
  replace = TRUE
)

c(
  OriginalMedian = median(Delay),
  BootstrapMedian = median(OneBootstrapSample),
  UniqueCasesRepresented = length(unique(OneBootstrapSample))
)
```

Now repeat the resampling process $B=5{,}000$ times.

```{r bootstrap-median, fig.width=7, fig.height=4.5}
set.seed(345)
B <- 5000
BootMedian <- replicate(
  B,
  median(sample(Delay, size = length(Delay), replace = TRUE))
)

hist(
  BootMedian, breaks = 30,
  col = "#9ECAE1", border = "white",
  xlab = "Bootstrapped median delay",
  main = "Five thousand alternative samples from the sample"
)
abline(v = median(Delay), col = "#D55E00", lwd = 3)
```

The bootstrap standard error is the standard deviation of the bootstrap estimates:

$$
SE_{boot}=\sqrt{
\frac{1}{B-1}
\sum_{b=1}^{B}
(\widehat\theta_b^*-\overline\theta^*)^2
}.
$$

The bootstrap estimate of bias is

$$
\widehat{Bias}_{boot}=\overline\theta^*-\widehat\theta.
$$

```{r bootstrap-summary}
BootSE <- sd(BootMedian)
BootBias <- mean(BootMedian) - median(Delay)

c(
  ObservedMedian = median(Delay),
  BootstrapBias = BootBias,
  BootstrapSE = BootSE
)
```

## Bootstrap Confidence Intervals: More Than One Flavor

There is no single thing called *the bootstrap confidence interval*. Several methods use the bootstrap distribution differently, and they do not have identical accuracy.

### Normal Interval

The normal interval uses the bootstrap standard error in a familiar symmetric interval:

$$
\widehat\theta\pm z_{1-\alpha/2}SE_{boot}.
$$

It is simple, but symmetry may be a bad approximation when the bootstrap distribution is skewed.

### Percentile Interval

The percentile interval uses the bootstrap quantiles directly:

$$
[q^*_{\alpha/2},q^*_{1-\alpha/2}].
$$

For a 95% interval, take the 2.5th and 97.5th percentiles. This is intuitive, but intuition and good coverage are not always roommates.

### Basic Interval

The basic interval reflects the bootstrap errors around the original estimate:

$$
[2\widehat\theta-q^*_{.975},
  2\widehat\theta-q^*_{.025}].
$$

If the bootstrap estimates tend to fall above the observed estimate, the basic interval reflects that pattern below it.

### Studentized or Bootstrap-$t$ Interval

The studentized method bootstraps a standardized pivot:

$$
t_b^*=\frac{\widehat\theta_b^*-\widehat\theta}{SE_b^*}.
$$

It then uses quantiles of the $t_b^*$ distribution to construct the interval. This can achieve better coverage, but every bootstrap sample needs its own standard-error estimate. That may require a second bootstrap inside the first bootstrap, because apparently the Probability Gods enjoy nesting dolls.

### BCa Interval

The **bias-corrected and accelerated** interval adjusts the percentile cutoffs using two quantities:

- $z_0$, which describes median bias in the bootstrap distribution.
- $a$, the acceleration constant, which describes how rapidly the standard error changes with the true parameter.

The acceleration term is commonly estimated from jackknife values. The jackknife therefore returns inside the bootstrap, wearing a fake mustache and pretending it never left. BCa intervals are second-order accurate under suitable conditions and often useful for skewed statistics, but they can be unstable when the sample is tiny or the statistic is poorly behaved [@efron2003].

```{r boot-package}
median_statistic <- function(data, indices) {
  median(data[indices])
}

set.seed(346)
BootObject <- boot::boot(
  data = Delay,
  statistic = median_statistic,
  R = 5000
)

boot::boot.ci(
  BootObject,
  type = c("norm", "basic", "perc", "bca")
)
```

::: {.callout-important title="Coverage Has Two Tails"}
A nominal 95% interval should miss the true value about 2.5% of the time below and 2.5% above across repeated samples. An interval that misses 5% entirely on one side still has 95% total coverage, but it is systematically off-center. Better bootstrap intervals try to improve both overall coverage and tail balance [@efron2003].
:::

::: {.callout-warning title="A 95% Confidence Interval Is Not a 95% Probability Statement About This Finished Interval"}
After a frequentist interval has been calculated, the fixed population parameter is either inside it or outside it. The 95% refers to the long-run success rate of the procedure across repeated samples. Hoekstra and colleagues gave six false interpretations of confidence intervals to 442 first-year students, 34 master's students, and 118 researchers. Only 2%, 0%, and 3%, respectively, rejected all six false statements [@hoekstra2014]. Confidence intervals are useful, but the word *confidence* apparently encourages our brains to manufacture meanings.
:::

## Bootstrapping a Regression Slope

For independent observations, a **case bootstrap** or **pairs bootstrap** resamples rows. Each resampled participant keeps their outcome and all predictors attached.

```{r bootstrap-regression}
set.seed(347)
RegressionData <- data.frame(
  Sleep = runif(90, 3, 9)
)
RegressionData$Memory <-
  42 + 4 * RegressionData$Sleep +
  rnorm(90, 0, 3 + RegressionData$Sleep)

OriginalModel <- lm(Memory ~ Sleep, data = RegressionData)
OriginalSlope <- coef(OriginalModel)["Sleep"]

BootSlopes <- replicate(5000, {
  rows <- sample(seq_len(nrow(RegressionData)), replace = TRUE)
  coef(lm(Memory ~ Sleep, data = RegressionData[rows, ]))["Sleep"]
})

c(
  OriginalSlope = OriginalSlope,
  BootstrapSE = sd(BootSlopes),
  quantile(BootSlopes, c(.025, .975))
)
```

The row bootstrap is not the only regression bootstrap.

### Case or Pairs Bootstrap

Resample complete rows $(Y_i,X_i)$. This preserves the relationship among all variables within each case and allows the observed predictor values to vary across bootstrap samples. It is a defensible default when cases were independently sampled from a population.

### Residual Bootstrap

Fit the model, hold $X$ fixed, resample centered residuals, and create

$$
Y_i^*=\widehat Y_i+e_i^*.
$$

This treats the design matrix as fixed and assumes the residuals are exchangeable. Ordinary residual resampling is therefore questionable under heteroscedasticity because small-variance and large-variance residuals are thrown into one communal bucket.

```{r residual-bootstrap}
set.seed(348)
FittedValues <- fitted(OriginalModel)
CenteredResiduals <- residuals(OriginalModel) - mean(residuals(OriginalModel))

ResidualSlopes <- replicate(3000, {
  YStar <- FittedValues + sample(CenteredResiduals, replace = TRUE)
  coef(lm(YStar ~ RegressionData$Sleep))[2]
})

sd(ResidualSlopes)
```

### Wild Bootstrap

A wild bootstrap keeps each residual attached to its predictor location but multiplies it by a random weight with mean zero and variance one:

$$
Y_i^*=\widehat Y_i+e_iw_i.
$$

This preserves the way residual size changes across $X$ better than shuffling residuals. Wild-bootstrap ideas grew from work on resampling regression under heteroscedasticity [@wu1986].

```{r wild-bootstrap}
set.seed(349)
WildSlopes <- replicate(3000, {
  Weights <- sample(c(-1, 1), nrow(RegressionData), replace = TRUE)
  YStar <- FittedValues + residuals(OriginalModel) * Weights
  coef(lm(YStar ~ RegressionData$Sleep))[2]
})

sd(WildSlopes)
```

### Parametric Bootstrap

A parametric bootstrap simulates new outcomes from the fitted probability model. For a Gaussian regression, we might simulate

$$
Y_i^*\sim N(\widehat Y_i,\widehat\sigma^2).
$$

For logistic regression, we might simulate binary outcomes using each fitted probability. This method can be efficient when the model is correct. If the model is wrong, the simulation faithfully reenacts the wrong model thousands of times.

## Preserve the Dependence Structure

If observations are repeated, nested, clustered, or ordered in time, resampling individual rows usually destroys the design.

- **Cluster bootstrap:** Resample whole participants, classrooms, families, clinics, or other independently sampled clusters.
- **Multilevel bootstrap:** Resample at more than one level when the sampling process justifies it.
- **Block bootstrap:** Resample adjacent blocks of a time series so short-range dependence remains intact [@singhxie2008].
- **Model-based bootstrap:** Simulate from a fitted mixed model or time-series model while preserving its assumed structure.

::: {.callout-important title="Resample the Unit That Was Sampled"}
If trials are nested within people, treating trials as independent resampling units manufactures a fictional sample size. If entire classrooms were sampled, resampling individual students pretends the classroom did not exist. Preserve the design. The bootstrap is flexible, not psychic.
:::

## Bootstrap Estimation Is Not Automatically a Hypothesis Test

Resampling the observed data estimates uncertainty around the observed statistic. A hypothesis test asks what results would occur under a null model. Those are different jobs.

To test a null hypothesis, we may need to:

- Center outcomes so the null value is true.
- Permute group labels under exchangeability.
- Simulate from a restricted model that imposes the null.
- Compare full and reduced models using a parametric bootstrap.

::: {.callout-warning title="A Bootstrap Distribution Is Not Automatically a Null Distribution"}
Dividing a bootstrap estimate by its bootstrap standard error and feeding the result to an ordinary $t$ table is not a universal bootstrap test. First decide what must be true under the null, then create resamples in which that null is actually true.
:::

## How Many Replications?

A few hundred replications may be enough to teach the idea or debug code. Standard errors commonly need at least a few thousand. Confidence-interval endpoints depend on tail quantiles and often need more. Run enough replications that repeating the analysis with a different seed does not change the reported conclusion or the meaningful digits.

The number of bootstrap replications $B$ controls **Monte Carlo error** in the calculation. It does not increase the original sample size. More replications do not repair a biased sample, weak measurement, or a resampling scheme that ignores the design. Ten thousand wrong resamples are wrong with excellent numerical stability.

## What the Bootstrap Can and Cannot Do

The bootstrap is useful when:

- A theoretical sampling distribution is hard to derive.
- A statistic has a skewed or otherwise awkward sampling distribution.
- We need standard errors or intervals for medians, indirect effects, complex effect sizes, or model predictions.
- We want to see how sensitive an estimate is to sampling variation.

The bootstrap cannot manufacture information that was not in the original data. It struggles when:

- The sample is extremely small.
- Rare but important parts of the population are absent.
- The estimator sits on a boundary or behaves discontinuously.
- The resampling plan does not preserve clustering or time dependence.
- The model is badly misspecified.

Efron called attention to both the power and the limits of the plug-in principle: the empirical distribution is a powerful stand-in for the population, but it can inherit and sometimes amplify the sample's pathologies [@efron2003]. In technical language, garbage in, garbage out. In Alex language, the computer cannot rescue a study that was thrown into the volcano before the analysis began.

## The Short Story

The jackknife removes one observation at a time and is useful for studying bias, standard errors, and influence. The bootstrap samples with replacement from the empirical distribution and can approximate the sampling distribution of a much wider range of statistics. Choose the resampling unit to match the design, choose the interval to match the estimator, use enough replications for stability, and remember that computation is not forgiveness.
