---
title: "Paired-Samples t-Test"
fontsize: 11pt
geometry: margin=.5in
format:
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5.5,
  fig.height = 3.5,
  fig.align = "center",
  fig.pos = "H",
  results = "hold"
)
```

::: {.callout-note title="The Same People Have Returned"}
Use a paired-samples t-test when the **same people are measured twice** or when observations form meaningful pairs. The two scores from one person know each other. Pretending they are independent is statistical identity theft.

The central idea is simple: turn each pair into one difference score, then ask whether the average difference is zero. That is the whole test. Nobody needs a random effect yet. The random effects are waiting in the next chapter with paperwork.
:::

## Why Pair the Scores?

An **independent-samples** design compares different people. Half the students receive chocolate during a lecture and half receive nothing except the lecture itself, which already seems cruel. Each student contributes one score.

A **repeated-measures** design measures the same students twice. Every student attends one lecture without chocolate and one with chocolate. Each student is now their own control.

A **matched-pairs** design uses different people who were deliberately matched on something important, such as age or baseline ability. This can help, but matched strangers are not magical clones. If you place two undergraduates beside each other because they have the same GPA, one may still be a morning person while the other communicates only through exhausted grunting.

Repeated scores tend to be correlated because stable characteristics follow people into both conditions. A student who generally looks at the screen for a long time may do that with or without chocolate. Another student may stare out the window in both conditions while contemplating escape. Pairing lets us remove some of those stable person-to-person differences.

## A Chocolate-Fixation Study

Suppose we ask whether eating a small piece of chocolate during a lecture changes **fixation time**: the average number of seconds a student's eyes remain on the screen per glance. We measure 48 students twice:

- once with no chocolate;
- once with chocolate.

This example is simulated. Do not distribute mystery candy to human participants and announce that a textbook authorized it. Allergies, informed consent, and research ethics remain annoyingly real.

```{r simulate-paired-data}
set.seed(343)

n <- 48
participant_effect <- rnorm(n, mean = 0, sd = 1.20)

paired_wide <- data.frame(
  participant = factor(seq_len(n)),
  no_chocolate = 5.00 + participant_effect + rnorm(n, 0, 0.80),
  chocolate = 5.75 + participant_effect + rnorm(n, 0, 0.80)
)

paired_wide$difference <-
  paired_wide$chocolate - paired_wide$no_chocolate

paired_long <- tidyr::pivot_longer(
  paired_wide,
  cols = c(no_chocolate, chocolate),
  names_to = "condition",
  values_to = "fixation"
)

paired_long$condition <- factor(
  paired_long$condition,
  levels = c("no_chocolate", "chocolate"),
  labels = c("No chocolate", "Chocolate")
)
```

The **wide data** have one row per person and one column per condition. This shape makes the pairing obvious and works nicely for a paired t-test. The **long data** have two rows per person. We use that shape for the picture below and will need it again when we reach mixed regression.

## Look at the Pairs Before Testing Them

```{r plot-pairs}
library(ggplot2)

ggplot(
  paired_long,
  aes(x = condition, y = fixation, group = participant)
) +
  geom_line(alpha = 0.30, color = "gray55") +
  geom_point(alpha = 0.60, color = "steelblue4") +
  stat_summary(
    aes(group = 1), fun = mean, geom = "line",
    color = "firebrick", linewidth = 1.3
  ) +
  stat_summary(
    aes(group = 1), fun = mean, geom = "point",
    color = "firebrick", size = 3
  ) +
  labs(
    x = NULL,
    y = "Mean fixation time (seconds)",
    title = "Every gray line is one exploited volunteer"
  ) +
  theme_minimal(base_size = 11)
```

The red line shows the mean change. The gray lines show something the means hide: some people change more than others, and a few may move in the opposite direction. If you plot only two bars, all those people are compressed into rectangular coffins.

## The Paired t-Test Is a One-Sample t-Test in Disguise

For each person, compute a difference score:

$$
D_i=Y_{i,\text{chocolate}}-Y_{i,\text{no chocolate}}
$$

Then ask whether the population mean difference is zero:

$$
t=\frac{M_D-0}{S_D/\sqrt{n}}
$$

This is the one-sample t-test from the earlier chapter. The paired t-test does not compare two unrelated piles of scores. It reduces every pair to one difference and tests the set of differences.

```{r paired-t-test}
paired_result <- t.test(
  paired_wide$chocolate,
  paired_wide$no_chocolate,
  paired = TRUE
)

paired_result
```

### R Has Become Picky About the Formula Method

The formula method of `t.test()` does not accept `paired = TRUE`, so this will fail:

```{r incorrect-formula-syntax, eval=FALSE}
t.test(fixation ~ condition, data = paired_long, paired = TRUE)
```

Use the two paired vectors, as above. Even better, make the pairing visible by calculating the differences yourself:

```{r difference-score-test}
difference_result <- t.test(paired_wide$difference, mu = 0)
difference_result
```

The t statistic, degrees of freedom, and p-value match. R has not discovered two separate truths. We asked the same question twice.

::: {.callout-important title="The Unit of Analysis Is the Pair"}
We began with 96 scores, but the test has 48 difference scores. Therefore, $df=n-1=47$. Counting every measurement as an independent person would double the apparent sample size and create confidence unsupported by the design. The spreadsheet may have 96 rows. The study still has 48 humans.
:::

## Effect Size

A common standardized effect for paired data is Cohen's $d_z$:

$$
d_z=\frac{M_D}{S_D}
$$

```{r paired-effect-size}
d_z <- mean(paired_wide$difference) /
  sd(paired_wide$difference)

d_z
```

This standardizes the mean change by the standard deviation of the **difference scores**. Say which version of $d$ you used because repeated-measures effect sizes have several possible denominators. Calling all of them "Cohen's d" and wandering away is how future readers begin sharpening office supplies.

## Why Pairing Can Help

The variance of a difference score is:

$$
S_D^2=S_1^2+S_2^2-2rS_1S_2
$$

When the two measurements are positively correlated, the last term subtracts variability. Stable person differences cancel: the consistently attentive student is compared with themself, and the student planning an elaborate escape is also compared with themself.

This often gives a smaller standard error than an independent-samples analysis. Pairing does not guarantee more power, however. If the repeated scores barely correlate, little stable variation is removed. Pairing also cannot repair a terrible measure. Measuring attention by counting how often someone blinks at the ceiling remains a terrible plan in both conditions.

## Assumptions That Actually Matter

For the paired-samples t-test:

- the observations form genuine pairs;
- different pairs are independent of one another;
- the outcome is quantitative;
- the **difference scores** have no severe outliers;
- for small samples, the population distribution of difference scores is reasonably normal.

The test does **not** require the two condition variances to be equal. It operates on one set of differences. Inspect those differences, not two separate normality tests performed as ritual offerings.

```{r inspect-differences, fig.width=5.5, fig.height=2.8}
ggplot(paired_wide, aes(x = difference)) +
  geom_histogram(binwidth = 0.35, color = "white", fill = "steelblue4") +
  geom_vline(xintercept = 0, linetype = 2, color = "firebrick") +
  labs(
    x = "Chocolate minus no-chocolate fixation time",
    y = "Number of students"
  ) +
  theme_minimal(base_size = 11)
```

## Reporting the Result

For the present study, a concise report could be:

> Students showed longer fixation times during the chocolate condition ($M =$ `r round(mean(paired_wide$chocolate), 2)`, $SD =$ `r round(sd(paired_wide$chocolate), 2)`) than during the no-chocolate condition ($M =$ `r round(mean(paired_wide$no_chocolate), 2)`, $SD =$ `r round(sd(paired_wide$no_chocolate), 2)`). The mean paired difference was `r round(mean(paired_wide$difference), 2)` seconds, 95% CI [`r round(paired_result$conf.int[1], 2)`, `r round(paired_result$conf.int[2], 2)`], $t(`r unname(paired_result$parameter)`) =$ `r round(unname(paired_result$statistic), 2)`, $p$ `r ifelse(paired_result$p.value < .001, "< .001", paste("=", round(paired_result$p.value, 3)))`, $d_z =$ `r round(d_z, 2)`.

The statistic cannot prove that chocolate *caused* better attention unless the study was properly randomized and alternative explanations were controlled. It certainly cannot prove that replacing every lecture with a bucket of candy is sound educational policy.

## When Two Measurements Are Not Enough

One difference score works beautifully when each person has two measurements and the question is about their average change. With three or more occasions, missing observations, both within-person and between-person predictors, or people who change at different rates, one difference score cannot represent the whole design.

That is where [mixed regression](Chapter_Mixed_Regression.qmd) begins. It keeps the repeated observations, tells the model which rows belong to the same human, and allows people to differ without creating one indicator variable for every person who wandered into the study.

## The Short Story

- Use a paired-samples t-test when observations form genuine pairs.
- The same person measured twice is the most common paired design.
- Compute one difference score for each pair.
- A paired-samples t-test is a one-sample t-test on those differences.
- The unit of analysis is the pair, not the individual measurement.
- Inspect the distribution and outliers of the **differences**.
- Report the mean change, confidence interval, p-value, and the specific paired effect size you used.
- When the design grows beyond one clean pair of scores, continue to the mixed-regression chapter.

The grand lesson is not "memorize another command." It is "keep the two scores attached to the human who produced them." The humans already know they are the same person. R needs paperwork.
