---
title: "Categorical-by-Categorical Interactions: The 2-by-2 Design"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.width = 4.5)
knitr::opts_chunk$set(fig.height = 3.5)
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(fig.show = "hold")
```

```{r load-libraries}
library(tidyverse)
library(ggplot2)
library(emmeans)
library(gridExtra)
```


<!-- ---Alex--- Light-touch opening, per your call. The "what changes / what stays the
     same" scaffolding is untouched because the review file calls it deliberate; only the
     recap paragraphs that used to open the chapter have moved into second position. The
     new opening also plants the dichotomization problem so that CH17-P01's callout is a
     payoff rather than a fresh interruption. The 119 is real, not a round number: the
     filter keeps 81 of 200. -->

## The Design Psychology Cannot Stop Running

Open a psychology journal to a random page. There is a very good chance the analysis you land on is a 2 × 2: two variables, two levels each, four groups, and one figure with two lines that either are or are not parallel. Drug versus placebo crossed with young versus old. Threat crossed with reassurance. Word lists crossed with sleep. It is the most-run design in the field, and if you go to graduate school in this discipline you will run one within the year.

So let us run one. Same friendship study as the prior chapter, same hypothesis, same 200 students.

There is one obstacle. A 2 × 2 needs two categorical variables, and Being Nice is not categorical. It is a perfectly good continuous rating from 0 to 10. So we are going to keep the least agreeable students, keep the most agreeable students, throw the 119 people in the middle of the scale directly into the bin, and call what survives **Low Nice** and **High Nice**.

Yes, that is roughly as bad as it sounds. We will come back to exactly how bad, and to why this book is allowed to do it and you mostly are not, at the moment we actually do it.

## Where We Are and Where We Are Going

In the prior chapter you learned how to detect, estimate, and interpret an interaction between a *continuous* predictor and a *categorical* one. In the friendship study, Being Nice (0–10 scale) was continuous and Going Out (Stay in vs. Go out) was categorical. The interaction told you that the **slope** of Being Nice on friendship closeness was different for students who went out compared to those who stayed in.

That kind of interaction is very common in psychology research. But so is a second kind: interactions where **both predictors are categorical**. This is the foundation of the **two-way factorial ANOVA**, and the 2 × 2 is its simplest and most common version.

This chapter assumes you have already read the prior chapter material. You know what an interaction is, you know how to interpret the interaction model equation, and you know how to use `anova()` to compare a main effects model to an interaction model. What is new here is what changes, and what stays the same, when both predictors are categorical.

### What Changes from prior chapter

The biggest conceptual change is this: **there is no continuous variable to center.** In the prior chapter centering was necessary because the intercept needed a meaningful zero point on the continuous scale. When both predictors are categorical, every predictor already takes on a small set of discrete values, and the "zero point" is simply the reference category of each variable, a specific cell in the design, not an abstract mean.

Everything else follows from that:

- No centering is needed. Both variables just need dummy coding.
- The interaction coefficient is no longer a "difference in slopes." There are no slopes here. Instead, $B_3$ is the **difference in group differences**, sometimes called the "difference of differences." We will unpack this carefully.
- The follow-up to the interaction is **simple effects analysis**: testing, within each level of one variable, whether the levels of the other variable differ significantly. We use `pairs()` from the `emmeans` package instead of `emtrends()`.
- Effect sizes for pairwise comparisons are expressed as **Cohen's *d***, not unstandardized slopes.

What **stays the same**: the interaction is still about "it depends." The visual signature is still non-parallel lines (or unequal gaps in a bar chart). The model comparison is still done with `anova()` and $\Delta R^2$. The APA write-up still follows the same structure.

### What the 2×2 Looks Like as a Table

Before we touch any data, it helps to lay out the full design as a 2 × 2 table. Each cell represents a unique combination of the two categorical predictors, and we expect a separate (estimated) mean outcome in each cell.

|  | **Stay in** | **Go out** |
|---|---|---|
| **Low Nice** | $\mu_{11}$: Low Nice × Stay in | $\mu_{12}$: Low Nice × Go out |
| **High Nice** | $\mu_{21}$: High Nice × Stay in | $\mu_{22}$: High Nice × Go out |

The interaction question is: **is the difference between High Nice and Low Nice the same in both columns (Going Out groups)?** Or equivalently: **is the difference between Go out and Stay in the same in both rows (Niceness groups)?**

Both questions ask the same thing. The answer is the interaction.


## The Core Concept: The "Difference of Differences"

### What It Means When Both Predictors Are Categorical

In the prior chapter, the interaction coefficient $B_3$ told you how much steeper the slope of Being Nice was for the "Go out" group compared to the "Stay in" group. Slopes are a continuous concept: the rate at which friendship closeness increases as niceness increases by one unit.

When Being Nice is categorical (Low vs. High), there are no slopes. Instead, there is a **gap**: the difference in predicted friendship closeness between High Nice and Low Nice students. The interaction now asks whether that gap is the same size in both Going Out groups.

Here is the idea in formula form. Define:

- **Effect of Niceness for Stay-in students** = $\mu_{21} - \mu_{11}$ (High Nice – Low Nice, among those who stay in)
- **Effect of Niceness for Go-out students** = $\mu_{22} - \mu_{12}$ (High Nice – Low Nice, among those who go out)

The **interaction** is the difference between these two effects:

$$B_3 = (\mu_{22} - \mu_{12}) - (\mu_{21} - \mu_{11})$$

This is the "difference of differences." If there is no interaction, the effect of Niceness is the same size in both Going Out groups, and $B_3 = 0$. If there is an interaction, the effect of Niceness is larger in one group than the other, and $B_3 \neq 0$.

Equivalently (and this is worth pausing on), you can write it the other way:

$$B_3 = (\mu_{22} - \mu_{21}) - (\mu_{12} - \mu_{11})$$

This version asks: is the effect of Going Out (Go out – Stay in) the same for High Nice and Low Nice students? The two formulations are mathematically identical and represent the same interaction. Which framing you use depends on which story you want to tell. In our study, we are primarily interested in whether the **benefit of being nice depends on whether you go out**, so we will usually use the first framing.

### Visual Hypotheses: What We Expect to See

As in the prior chapter, we can sketch out our two competing possibilities before running a single model.

```{r hypothesis-plots, fig.width=7, fig.height=3.25, echo=FALSE}
#| fig-cap: "The 2 × 2 version of the two hypotheses. With only two levels on the x axis, non-parallel no longer means different slopes; it means differently sized gaps."
#| fig-alt: "Two side-by-side line graphs, each with Low Nice and High Nice on the x axis and predicted closeness from 0 to 75 on the y axis, and two connected points per Going Out group. In the left panel, Hypothesis A No Interaction, the Stay in line rises from 15 to 20 and the Go out line rises from 20 to 25, so the two lines are parallel and the vertical gap stays at 5. In the right panel, Hypothesis B Interaction Present, both lines start at 15 for Low Nice, but the Stay in line rises only to 16 while the Go out line jumps to 61, opening a large gap at the right."

## Hypothesis A: No interaction (equal gaps)
df_A <- data.frame(
  NiceGroup = rep(c("Low Nice", "High Nice"), 2),
  GoingOut  = rep(c("Stay in", "Go out"), each = 2),
  Predicted = c(15, 20, 20, 25)
)
df_A$NiceGroup <- factor(df_A$NiceGroup, levels = c("Low Nice", "High Nice"))
df_A$GoingOut  <- factor(df_A$GoingOut,  levels = c("Stay in", "Go out"))

p_A <- ggplot(df_A, aes(x = NiceGroup, y = Predicted,
                         color = GoingOut, group = GoingOut)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Stay in" = "#5b8db8", "Go out" = "#e07b54")) +
  scale_y_continuous(limits = c(0, 75)) +
  labs(title = "Hypothesis A: No Interaction",
       subtitle = "Parallel lines: same gap for both groups",
       x = "Niceness Group", y = "Predicted Closeness (0-100)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 8, color = "grey40"))

## Hypothesis B: Interaction (unequal gaps)
df_B <- data.frame(
  NiceGroup = rep(c("Low Nice", "High Nice"), 2),
  GoingOut  = rep(c("Stay in", "Go out"), each = 2),
  Predicted = c(15, 16, 15, 61)
)
df_B$NiceGroup <- factor(df_B$NiceGroup, levels = c("Low Nice", "High Nice"))
df_B$GoingOut  <- factor(df_B$GoingOut,  levels = c("Stay in", "Go out"))

p_B <- ggplot(df_B, aes(x = NiceGroup, y = Predicted,
                         color = GoingOut, group = GoingOut)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Stay in" = "#5b8db8", "Go out" = "#e07b54")) +
  scale_y_continuous(limits = c(0, 75)) +
  labs(title = "Hypothesis B: Interaction Present",
       subtitle = "Non-parallel lines: unequal gaps",
       x = "Niceness Group", y = "Predicted Closeness (0-100)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 8, color = "grey40"))

gridExtra::grid.arrange(p_A, p_B, ncol = 2)
```

Notice the visualization has changed from the prior chapter. Instead of regression lines across a continuous X-axis, we now have two points connected by a line for each Going Out group. The key visual signature of an interaction is still **non-parallel lines**. But "non-parallel" now means the **gaps between the two groups are different sizes** across the niceness levels, rather than different slopes.

- In **Hypothesis A**, the two lines are parallel: the gap between "Go out" and "Stay in" is the same for Low Nice and High Nice students. Being nice adds a fixed amount regardless of going out.
- In **Hypothesis B**, the lines fan out: the gap between "Go out" and "Stay in" is tiny for Low Nice students but enormous for High Nice students. This is our interaction hypothesis: being nice only pays off substantially if you are also out there meeting people.


## Before We Run the Models: Dummy Coding Both Variables

### No Centering Required

In the prior chapter, you had to center the continuous predictor (Being Nice) before building the interaction model. Centering shifted the scale so that zero represented the average person, making the intercept interpretable.

When both predictors are categorical, there is nothing to center. Each predictor already takes on discrete values. The "zero point" for each predictor is simply its **reference category**, whichever group we designate as the baseline.

This is actually simpler than the prior chapter. The only setup step is making sure both variables are properly factored in R, with the reference groups ordered first.

### Dummy Coding Both Variables

We have two categorical predictors, each with two levels:

| Variable | Levels | Reference group (coded 0) | Comparison group (coded 1) |
|---|---|---|---|
| **NiceGroup** | Low Nice, High Nice | Low Nice | High Nice |
| **GoingOut** | Stay in, Go out | Stay in | Go out |

With this coding, the **four coefficients** in the interaction model have the following meanings:

| Coefficient | What it represents |
|---|---|
| $B_0$ (Intercept) | Predicted closeness for **Low Nice × Stay in** students, the baseline cell |
| $B_1$ (NiceGroup: High Nice) | High Nice – Low Nice **among Stay-in students** (simple effect of niceness for stay-in group) |
| $B_2$ (GoingOut: Go out) | Go out – Stay in **among Low Nice students** (simple effect of going out for low-nice group) |
| $B_3$ (NiceGroup × GoingOut) | The "difference of differences": does the effect of niceness differ by going-out group? |

Notice that $B_1$ and $B_2$ are no longer main effects in the traditional sense: they are **simple effects**, each representing the effect of one variable *holding the other variable at its reference level*. This is the same logic as in the prior chapter, where $B_1$ was the slope for the reference group only. Here, $B_1$ is the niceness gap for the reference going-out group only.

### Recovering All Four Cell Means

One of the useful features of the 2 × 2 interaction model is that you can reconstruct **every cell mean** directly from the coefficients. Here is the formula for each cell:

| Cell | Formula | Plain meaning |
|---|---|---|
| Low Nice × Stay in | $B_0$ | Baseline |
| High Nice × Stay in | $B_0 + B_1$ | Baseline + niceness effect (Stay in) |
| Low Nice × Go out | $B_0 + B_2$ | Baseline + going-out effect (Low Nice) |
| High Nice × Go out | $B_0 + B_1 + B_2 + B_3$ | All four terms |

The final cell is the most important: for High Nice students who also go out, the prediction is the baseline plus the simple effect of niceness, plus the simple effect of going out, **plus the interaction term**. The interaction term $B_3$ adjusts the prediction for the cell where both variables are at their non-reference level. If there is no interaction ($B_3 = 0$), the effects of the two variables add up independently and the model is purely additive.


## The Data

We return to our 200 first-semester students. Recall that we measured friendship closeness (0–100 feelings thermometer), agreeableness (0–10), and whether they went out regularly.

For this analysis, we focus on the **extremes of the niceness distribution**: students who scored below 2 on the agreeableness scale (near 0 = **Low Nice**) and students who scored above 8 (**High Nice**). Everyone in the middle of the scale is dropped, which is most of the sample, and we keep only the contrast between the most and least agreeable people in it.

::: {.callout-warning title="Why We Are Allowed to Do This, and You Are Mostly Not"}
I am about to butcher a perfectly good continuous variable, on purpose, in front of you. I am doing it because the 2 × 2 is how a large share of psychology structures its experiments, and you need to be able to read, run, and interpret one.

Doing this to *real* continuous data is a different matter. Median splits and extreme-groups designs throw away information you paid to collect, shrink your statistical power, and can manufacture effects that are not there [@maccallum2002]. Notice what happens below: 200 participants become 81. We are discarding well over half the study to make a picture that is easier to draw.

The rule is simple. **If your variable is continuous, use the previous chapter's model.** Categorize when the variable is genuinely categorical, or when someone else already categorized it before you arrived, not because a 2 × 2 is easier to explain in a talk.
:::

Here is the simulation, including the filter that does the damage. This is the same 200 students and the same seed as the prior chapter, so `Friend.Data.Full` below is identical to the dataset we used there:

```{r simulate-data, echo=TRUE}
library(tidyverse)

set.seed(42)
n <- 200
X <- runif(n, 0, 10)
Z <- rbinom(n, 1, 0.5)
Y <- 0.05 * X + 0 * Z + 5 * X * Z + 15 + rnorm(n, sd = 7)
Friend.Data.Full <- data.frame(BeingNice = X, GoingOut = Z, Closeness = Y)
Friend.Data.Full$GoingOut_ <- factor(Friend.Data.Full$GoingOut,
                                     levels = c(0, 1),
                                     labels = c("Stay in", "Go out"))

# Create niceness group for extremes
Friend.Data.Full$NiceGroup_ <- ifelse(Friend.Data.Full$BeingNice < 2,  "Low Nice",
                               ifelse(Friend.Data.Full$BeingNice > 8, "High Nice", NA))

# Keep only extreme groups
Friend.Data <- Friend.Data.Full %>%
  filter(!is.na(NiceGroup_)) %>%
  mutate(NiceGroup_ = factor(NiceGroup_, levels = c("Low Nice", "High Nice")))

n_low_stayin  <- sum(Friend.Data$NiceGroup_ == "Low Nice"  & Friend.Data$GoingOut_ == "Stay in")
n_low_goout   <- sum(Friend.Data$NiceGroup_ == "Low Nice"  & Friend.Data$GoingOut_ == "Go out")
n_high_stayin <- sum(Friend.Data$NiceGroup_ == "High Nice" & Friend.Data$GoingOut_ == "Stay in")
n_high_goout  <- sum(Friend.Data$NiceGroup_ == "High Nice" & Friend.Data$GoingOut_ == "Go out")
```

The resulting dataset has `r nrow(Friend.Data)` participants spread across four cells:

```{r cell-counts, echo=TRUE}
Friend.Data %>%
  group_by(NiceGroup_, GoingOut_) %>%
  summarise(n = n(),
            M_Close  = round(mean(Closeness), 2),
            SD_Close = round(sd(Closeness), 2),
            .groups = "drop")
```

The cell counts are not equal, which is typical when you extract extremes from a continuous distribution rather than randomly assigning participants to conditions. Unequal cells are no problem for the regression estimates. (Classic ANOVA arithmetic assumes balanced cells; with unequal n the ANOVA world descends into a fight about "Types of sums of squares" that regression coefficients let you skip. If you ever meet Type III SS in the wild, that is what the fight was about, and we will see it happen later in this chapter.)

The cell means already tell an interesting story. Among Low Nice students, going out is worth something but not much: about 16 on the thermometer if they stay in, about 23 if they go out. Among High Nice students the same comparison is enormous, roughly 19 against 60. That is the difference of differences we just defined, sitting in the raw means before we have fit anything, and it matches Hypothesis B.

## A First Look at the Data

Before running any models, let's visualize the four cells directly.

```{r first-look, fig.width=5, fig.height=3.5, echo=FALSE}
#| fig-cap: "The four cells before any model is fit. Jittered raw scores with the observed cell means connected by group."
#| fig-alt: "A plot with Low Nice and High Nice on the x axis and Friendship Closeness 0 to 100 on the y axis, showing jittered raw data points and two connected cell-mean lines. The blue Stay in line is nearly flat, from about 16 to 19. The orange Go out line rises steeply from about 23 to about 60. The Go out High Nice cluster of points sits well above every other cell."

cell_means <- Friend.Data %>%
  group_by(NiceGroup_, GoingOut_) %>%
  summarise(M = mean(Closeness), .groups = "drop")

ggplot(Friend.Data, aes(x = NiceGroup_, y = Closeness, color = GoingOut_)) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1.2) +
  geom_line(data = cell_means, aes(y = M, group = GoingOut_), linewidth = 1.2) +
  geom_point(data = cell_means, aes(y = M), size = 3.5) +
  scale_color_manual(values = c("Stay in" = "#5b8db8", "Go out" = "#e07b54")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Friendship Means by Group",
       x = "Niceness Group", y = "Friendship Closeness (0-100)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The pattern is striking. The "Stay in" line (blue) is almost flat: Low Nice and High Nice students who stay in have close to the same closeness scores. The "Go out" line (orange) rises sharply: High Nice students who go out form far closer friendships than Low Nice students who go out. The two lines are clearly **non-parallel**, which is the visual signature of an interaction. This picture already suggests Hypothesis B.


## Building the Models

### Model 1: The Additive Model

We begin with the main effects (additive) model. This model says the two variables each have an independent effect on friendship closeness, and those effects add up without interacting.

$$\widehat{Closeness} = B_0 + B_1 \cdot NiceGroup + B_2 \cdot GoingOut + \varepsilon$$

```{r additive-model, echo=TRUE}
Model.1 <- lm(Closeness ~ NiceGroup_ + GoingOut_, data = Friend.Data)
summary(Model.1)
```

```{r r2-store, include=FALSE}
R2.1 <- summary(Model.1)$r.squared
R2.2.temp <- summary(lm(Closeness ~ NiceGroup_ * GoingOut_, data = Friend.Data))$r.squared
```

The additive model estimates a single effect of NiceGroup (averaged across both Going Out groups) and a single effect of GoingOut (averaged across both Niceness groups). These are conventional main effects.

Notice that the R² here (`r round(R2.1, 3)`) is already reasonably high, driven primarily by the Going Out main effect. But as we will see, forcing the effects to be additive is a poor fit for the actual pattern in the data.

### Visualizing the Additive Model

```{r additive-plot, fig.width=4.5, fig.height=3.5}
#| fig-cap: "The additive model's estimated cell means, with error bars and the raw data behind them. The lines are parallel because the model is not permitted to draw anything else."
#| fig-alt: "A plot with Low Nice and High Nice on the x axis and predicted closeness 0 to 100 on the y axis, showing two parallel rising lines with error bars over faint raw data. The blue Stay in line runs from about 9 to 27 and the orange Go out line from about 33 to 51, both at the same angle. The raw points do not match: the actual Go out High Nice points cluster near 60, well above the line drawn for them, and the Stay in points sit near 16 to 19 rather than on their line."

Add.Em <- emmeans(Model.1, ~ NiceGroup_ | GoingOut_)
Add.Plot <- as.data.frame(Add.Em)

ggplot(Add.Plot, aes(x = NiceGroup_, y = emmean,
                     color = GoingOut_, group = GoingOut_)) +
  geom_point(data = Friend.Data,
             aes(x = NiceGroup_, y = Closeness, color = GoingOut_),
             alpha = 0.25, size = 1, position = position_jitter(width = 0.1)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
  scale_color_manual(values = c("Stay in" = "#5b8db8", "Go out" = "#e07b54")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Additive Model: Parallel Lines (forced)",
       x = "Niceness Group", y = "Predicted Closeness (0-100)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The additive model forces the two lines to be parallel. The gap between "Go out" and "Stay in" is identical for Low Nice and High Nice students. But look at the raw data: the "Go out" students in the High Nice group (orange, right side) are far higher than what the parallel-line model predicts for them, and the "Stay in" students in the same group are not as high as the model suggests. The parallel assumption is clearly wrong.

### Model 2: The Interaction Model

Now we add the interaction term:

$$\widehat{Closeness} = B_0 + B_1 \cdot NiceGroup + B_2 \cdot GoingOut + B_3 \cdot NiceGroup \times GoingOut + \varepsilon$$

```{r interaction-model, echo=TRUE}
Model.2 <- lm(Closeness ~ NiceGroup_ * GoingOut_, data = Friend.Data)
summary(Model.2)
```

```{r r2-store2, include=FALSE}
R2.1 <- summary(Model.1)$r.squared
R2.2 <- summary(Model.2)$r.squared
aov_comp  <- anova(Model.1, Model.2)
delta_F   <- round(aov_comp$F[2], 2)
delta_df1 <- aov_comp$Df[2]
delta_df2 <- aov_comp$Res.Df[2]
delta_p   <- aov_comp$`Pr(>F)`[2]
```

### Does the Interaction Model Explain More Variance?

```{r model-comparison, echo=TRUE}
knitr::kable(anova(Model.1, Model.2), digits = 3,
             caption = "Model Comparison: Additive vs. Interaction")
```

```{r delta-r2, echo=FALSE}
cat("Additive model R-squared:    ", round(R2.1, 3), "\n")
cat("Interaction model R-squared: ", round(R2.2, 3), "\n")
cat("Change in R-squared:         ", round(R2.2 - R2.1, 3), "\n")
```

The interaction model explains substantially more variance. The *F*-test ($\Delta R^2 =$ `r round(R2.2 - R2.1, 3)`, $F(`r delta_df1`, `r delta_df2`) =$ `r delta_F`, $p < .001$) confirms that this improvement is far larger than would be expected by chance. Adding the interaction term was worth it.


## Interpreting the Interaction Model

### The Plot First

```{r interaction-plot, fig.width=4.5, fig.height=3.5}
#| fig-cap: "The interaction model's estimated cell means. Freed to give each row its own gap, the model reproduces the four observed cell means exactly."
#| fig-alt: "A plot with Low Nice and High Nice on the x axis and predicted closeness 0 to 100 on the y axis, showing two clearly non-parallel lines with error bars over faint raw data. The blue Stay in line is nearly flat, from about 16 to 19. The orange Go out line climbs steeply from about 23 to about 60. The raw points now sit tightly around their own cell means."

Inter.Em <- emmeans(Model.2, ~ NiceGroup_ | GoingOut_)
Inter.Plot <- as.data.frame(Inter.Em)

ggplot(Inter.Plot, aes(x = NiceGroup_, y = emmean,
                       color = GoingOut_, group = GoingOut_)) +
  geom_point(data = Friend.Data,
             aes(x = NiceGroup_, y = Closeness, color = GoingOut_),
             alpha = 0.25, size = 1, position = position_jitter(width = 0.1)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
  scale_color_manual(values = c("Stay in" = "#5b8db8", "Go out" = "#e07b54")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Interaction Model: Non-Parallel Lines",
       x = "Niceness Group", y = "Predicted Closeness (0-100)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

This is the interaction model's predicted means for each cell. Notice the non-parallel lines: the "Stay in" line is nearly flat (very small gap between Low Nice and High Nice), while the "Go out" line rises steeply (a large gap between Low Nice and High Nice). Compare this to the additive model's forced parallel lines. The story told by these two models is completely different.

### Walking Through Each Coefficient

```{r reprint-model}
summary(Model.2)
```

```{r coef-store, include=FALSE}
B0 <- round(coef(Model.2)["(Intercept)"], 2)
B1 <- round(coef(Model.2)["NiceGroup_High Nice"], 2)
B2 <- round(coef(Model.2)["GoingOut_Go out"], 2)
B3 <- round(coef(Model.2)["NiceGroup_High Nice:GoingOut_Go out"], 2)
SE3 <- round(summary(Model.2)$coefficients["NiceGroup_High Nice:GoingOut_Go out", "Std. Error"], 2)
t3  <- round(summary(Model.2)$coefficients["NiceGroup_High Nice:GoingOut_Go out", "t value"], 2)
p3  <- summary(Model.2)$coefficients["NiceGroup_High Nice:GoingOut_Go out", "Pr(>|t|)"]
```

### The Intercept ($B_0 =$ `r B0`)

The intercept is the predicted closeness score for the **baseline cell**: Low Nice students who Stay in (both predictors at 0). This is a real, specific group of people in the dataset. A thoroughly disagreeable person who never leaves home still lands at about `r B0` points on the friendship thermometer, presumably because they live near someone and physical proximity does the work even for the unsociable.

### $B_1$: Effect of High Niceness **Among Stay-in Students** ($B_1 =$ `r B1`)

$B_1$ is the difference in predicted closeness between High Nice and Low Nice students **within the Stay-in group**. Looking at the "Stay in" line in the plot, this gap is small: $B_1 =$ `r B1` points on a 100-point scale, and non-significant.

For students who stay home, being highly agreeable provides virtually no additional benefit for friendship formation compared to being low in agreeableness. Without social exposure, personality differences do not matter much.

This is a **simple effect**: it is the effect of NiceGroup *within one specific level* of GoingOut (the reference level, Stay in). It is not a main effect.

### $B_2$: Effect of Going Out **Among Low Nice Students** ($B_2 =$ `r B2`)

$B_2$ is the difference in predicted closeness between Go-out and Stay-in students **within the Low Nice group**. Students low in agreeableness who go out score about `r B2` points higher on the thermometer than their equally disagreeable counterparts who stay in. Small, but real: this one is significant.

This is also a **simple effect**: the effect of GoingOut *within one specific level* of NiceGroup (the reference level, Low Nice).

### $B_3$: The Interaction Term ($B_3 =$ `r B3`)

This is the heart of the analysis. $B_3$ tells you how much **larger** the effect of niceness is for the Go-out group compared to the Stay-in group.

We can reconstruct the simple effect of Niceness for each Going Out group:

| Group | Simple effect of Niceness | Formula |
|---|---|---|
| Stay in | `r B1` | $B_1$ |
| Go out | `r round(B1 + B3, 2)` | $B_1 + B_3 =$ `r B1` $+$ `r B3` |

The difference between these two simple effects is $B_3 =$ `r B3`. For Go-out students, being highly agreeable is worth about `r round(B1 + B3, 2)` more points of closeness than being low in agreeableness. For Stay-in students, the same comparison yields only `r B1` points. The interaction coefficient captures this dramatic difference.

Here is the full cell mean summary:

| Cell | Predicted mean | Derivation |
|---|---|---|
| Low Nice × Stay in | `r B0` | $B_0$ |
| High Nice × Stay in | `r round(B0 + B1, 2)` | $B_0 + B_1$ |
| Low Nice × Go out | `r round(B0 + B2, 2)` | $B_0 + B_2$ |
| High Nice × Go out | `r round(B0 + B1 + B2 + B3, 2)` | $B_0 + B_1 + B_2 + B_3$ |


## Following Up the Interaction: Simple Effects

The interaction term tells us the two niceness gaps are significantly different between Going Out groups. But we also want to know: **within each Going Out group, is the difference between High Nice and Low Nice statistically significant?**

We use `pairs()` from the `emmeans` package to test each pairwise comparison within each level of the other variable.

### Simple Effects of Niceness within Each Going-Out Group

```{r simple-effects, echo=TRUE}
library(emmeans)

Fit.2 <- emmeans(Model.2, ~ NiceGroup_ | GoingOut_)
pairs(Fit.2, adjust = "none") %>% summary(infer = TRUE)
```

```{r simple-effects-store, include=FALSE}
se_out <- pairs(Fit.2, adjust = "none") %>% summary(infer = TRUE)
se_df  <- as.data.frame(se_out)
b_stayin <- round(se_df$estimate[se_df$GoingOut_ == "Stay in"], 2)
b_goout  <- round(se_df$estimate[se_df$GoingOut_ == "Go out"],  2)
p_stayin <- se_df$p.value[se_df$GoingOut_ == "Stay in"]
p_goout  <- se_df$p.value[se_df$GoingOut_ == "Go out"]
t_stayin <- round(se_df$t.ratio[se_df$GoingOut_ == "Stay in"], 2)
t_goout  <- round(se_df$t.ratio[se_df$GoingOut_ == "Go out"],  2)
se_stayin <- round(se_df$SE[se_df$GoingOut_ == "Stay in"], 2)
se_goout  <- round(se_df$SE[se_df$GoingOut_ == "Go out"],  2)
```

- **Stay-in group:** The difference between High Nice and Low Nice students is `r b_stayin` points, which is **not** statistically significant ($p =$ `r round(p_stayin, 3)`). For students who stay in, agreeableness does not significantly predict friendship closeness.

- **Go-out group:** The difference between High Nice and Low Nice students is `r b_goout` points, which is **highly** significant ($p <$ .001). Among students who go out, being highly agreeable is associated with substantially closer friendships than being low in agreeableness.

### Effect Sizes: Cohen's *d*

Pairwise comparisons in this framework can be accompanied by Cohen's *d* as an effect size measure. We use the model's residual standard deviation as the pooled standard deviation:

```{r effect-sizes, echo=TRUE}
eff_size(Fit.2,
         sigma = sigma(Model.2),
         edf   = df.residual(Model.2),
         method = "pairwise")
```

```{r d-store, include=FALSE}
d_out   <- eff_size(Fit.2, sigma = sigma(Model.2), edf = df.residual(Model.2))
d_df    <- as.data.frame(d_out)
d_stay  <- round(abs(d_df$effect.size[d_df$GoingOut_ == "Stay in"]), 2)
d_go    <- round(abs(d_df$effect.size[d_df$GoingOut_ == "Go out"]),  2)
```

The Cohen's *d* values confirm the pattern:

- **Stay in:** $d =$ `r d_stay`, which by Cohen's rough conventions is a small-to-medium effect, and which our test could not distinguish from zero at this sample size. Note that these two statements are not the same statement, and the gap between them is the whole reason we threw away 119 people. With 46 stay-in students, an effect this size is exactly the kind that goes undetected.
- **Go out:** $d =$ `r d_go`. That is not a typo, and "large" does not really cover it. Cohen's benchmark for large is 0.8; this is over five standard deviations. Effects like this do not happen in real personality research, they happen in simulated data where the author picked the parameters, and you should be suspicious of any real paper reporting one.


## APA Write-Up

Here is how you would report these results following APA style. The structure parallels the prior chapter: model comparison, interaction coefficient, then simple effects.


```{r apa-vals, include=FALSE}
sm2    <- summary(Model.2)$coefficients
sm1_f  <- summary(Model.1)$fstatistic
m1_F   <- round(sm1_f[1], 2)
m1_df1 <- sm1_f[2]
m1_df2 <- sm1_f[3]

cell_desc <- Friend.Data %>%
  group_by(NiceGroup_, GoingOut_) %>%
  summarise(M = round(mean(Closeness), 1),
            SD = round(sd(Closeness), 1), .groups = "drop")

m_ln_si  <- cell_desc$M[cell_desc$NiceGroup_ == "Low Nice"  & cell_desc$GoingOut_ == "Stay in"]
sd_ln_si <- cell_desc$SD[cell_desc$NiceGroup_ == "Low Nice"  & cell_desc$GoingOut_ == "Stay in"]
m_ln_go  <- cell_desc$M[cell_desc$NiceGroup_ == "Low Nice"  & cell_desc$GoingOut_ == "Go out"]
sd_ln_go <- cell_desc$SD[cell_desc$NiceGroup_ == "Low Nice"  & cell_desc$GoingOut_ == "Go out"]
m_hn_si  <- cell_desc$M[cell_desc$NiceGroup_ == "High Nice" & cell_desc$GoingOut_ == "Stay in"]
sd_hn_si <- cell_desc$SD[cell_desc$NiceGroup_ == "High Nice" & cell_desc$GoingOut_ == "Stay in"]
m_hn_go  <- cell_desc$M[cell_desc$NiceGroup_ == "High Nice" & cell_desc$GoingOut_ == "Go out"]
sd_hn_go <- cell_desc$SD[cell_desc$NiceGroup_ == "High Nice" & cell_desc$GoingOut_ == "Go out"]
```

Cell means and standard deviations are presented in Table 1. Among students who stayed in, friendship closeness was similar regardless of agreeableness group (Low Nice: *M* = `r m_ln_si`, *SD* = `r sd_ln_si`; High Nice: *M* = `r m_hn_si`, *SD* = `r sd_hn_si`). Among students who went out, High Nice students formed substantially closer friendships than Low Nice students (Low Nice: *M* = `r m_ln_go`, *SD* = `r sd_ln_go`; High Nice: *M* = `r m_hn_go`, *SD* = `r sd_hn_go`).

A hierarchical multiple regression was conducted to examine whether social engagement (Going Out) moderated the relationship between agreeableness group (Low vs. High Nice) and friendship closeness. Niceness group and Going Out (both dummy coded with Low Nice and Stay in as reference groups) were entered in Step 1 as main effects; their product was entered in Step 2. The additive model (Step 1) was statistically significant, $F(`r m1_df1`, `r m1_df2`) =$ `r m1_F`, $p < .001$, $R^2 =$ `r round(R2.1, 3)`. Adding the interaction term (Step 2) produced a statistically significant improvement in model fit, $\Delta R^2 =$ `r round(R2.2 - R2.1, 3)`, $F(`r delta_df1`, `r delta_df2`) =$ `r delta_F`, $p < .001$, indicating that the effect of agreeableness on friendship closeness differed significantly depending on whether students went out.

In the full interaction model ($R^2 =$ `r round(R2.2, 3)`, adjusted $R^2 =$ `r round(summary(Model.2)$adj.r.squared, 3)`), the interaction between Niceness Group and Going Out was statistically significant, $b =$ `r B3`, $SE =$ `r SE3`, $t(`r delta_df2`) =$ `r t3`, $p < .001$. To interpret the interaction, simple effects of Niceness Group were examined separately within each level of Going Out. Among students who stayed in, agreeableness group did not significantly predict friendship closeness, $b =$ `r b_stayin`, $SE =$ `r se_stayin`, $t(`r delta_df2`) =$ `r t_stayin`, $p =$ `r round(p_stayin, 3)`, $d =$ `r d_stay`. Among students who went out, High Nice students formed significantly closer friendships than Low Nice students, $b =$ `r b_goout`, $SE =$ `r se_goout`, $t(`r delta_df2`) =$ `r t_goout`, $p < .001$, $d =$ `r d_go`. These results indicate that the benefit of high agreeableness for friendship formation is conditional on social engagement: being a warm and agreeable person leads to substantially closer friendships only among students who are regularly out meeting others.


**Things to notice about this write-up:**

- We report **Cohen's *d*** alongside the simple effects, since the comparisons are between two categorical groups. This is the appropriate effect size for a two-group mean difference.
- The **model comparison** ($\Delta R^2$, *F*-test) is still the primary test of the interaction, exactly as in the prior chapter.
- We describe the interaction in plain language: *"the benefit of high agreeableness ... is conditional on social engagement."*
- The **main effects** from the interaction model are not heavily emphasized. In the presence of a significant interaction, they represent effects at the reference level of the other variable (simple effects), not population-level averages.


## APA-Style Figure

```{r apa-figure, fig.width=5, fig.height=4, echo=TRUE}
#| fig-alt: "A publication-style figure with Agreeableness Group, Low Nice and High Nice, on the x axis and Predicted Friendship Closeness 0 to 100 on the y axis. Two non-parallel lines connect model-estimated cell means with 95 percent confidence error bars, over faint jittered raw data. The blue Stay in line is nearly flat from about 16 to 19; the orange Go out line rises steeply from about 23 to about 60. The legend, titled Social Engagement, sits above the plot."

# APA interaction plot: categorical × categorical
library(ggplot2)

Fig.Data <- as.data.frame(emmeans(Model.2, ~ NiceGroup_ | GoingOut_))

ggplot(Fig.Data, aes(x = NiceGroup_, y = emmean,
                     color = GoingOut_, group = GoingOut_)) +
  # Raw data (faint)
  geom_point(data = Friend.Data,
             aes(x = NiceGroup_, y = Closeness, color = GoingOut_),
             alpha = 0.20, size = 1,
             position = position_jitter(width = 0.12)) +
  # Confidence intervals
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.08, linewidth = 0.8) +
  # Connecting lines
  geom_line(linewidth = 1.2) +
  # Cell mean points
  geom_point(size = 3.5) +
  scale_color_manual(values = c("Stay in" = "#5b8db8",
                                "Go out"  = "#e07b54"),
                     labels = c("Stay in", "Go out")) +
  scale_y_continuous(limits = c(0, 100),
                     breaks = seq(0, 100, 20)) +
  labs(x     = "Agreeableness Group",
       y     = "Predicted Friendship Closeness (0-100)",
       color = "Social Engagement") +
  theme_bw() +
  theme(legend.position   = "top",
        legend.title      = element_text(size = 9),
        legend.text       = element_text(size = 9),
        axis.title        = element_text(size = 10),
        panel.grid.minor  = element_blank())
```

*Figure 1.* Predicted friendship closeness as a function of agreeableness group and social engagement. Points represent model-estimated cell means; error bars represent 95% confidence intervals; faint points represent raw data. Among students who went out (*Go out*; orange), highly agreeable students formed substantially closer friendships than less agreeable students (*d* = `r d_go`). Among students who stayed in (*Stay in*; blue), agreeableness group was not significantly associated with friendship closeness (*d* = `r d_stay`).


## The ANOVA Costume

I have been rude about ANOVA since the preface, so it is only fair to introduce you properly, once, so you know what it is when someone inflicts it on you.

Everything we just did has a second name. Run the exact same model through `anova()` and it comes back dressed as a two-way factorial ANOVA:

```{r anova-costume, echo=TRUE}
anova(Model.2)
```

That is the table a paper means when it says "a 2 × 2 ANOVA revealed a significant interaction." Same data, same model, same `Model.2` object we have had this whole time. Nothing was refit.

```{r anova-costume-store, include=FALSE}
aov_tab <- anova(Model.2)
F_int   <- round(aov_tab$`F value`[3], 2)
# Square the UNROUNDED t, or this lands 0.01 away from F and ruins the point
t3_raw  <- summary(Model.2)$coefficients["NiceGroup_High Nice:GoingOut_Go out", "t value"]
t3_sq   <- round(t3_raw^2, 2)
carAov  <- car::Anova(Model.2, type = 2)
F_nice_I  <- round(aov_tab$`F value`[1], 2)
F_nice_II <- round(carAov$`F value`[1], 2)
F_go_I    <- round(aov_tab$`F value`[2], 2)
F_go_II   <- round(carAov$`F value`[2], 2)
F_int_II  <- round(carAov$`F value`[3], 2)
```

Look at the interaction row. $F =$ `r F_int`. Now square the *t* for $B_3$ from the regression output: $t =$ `r round(t3_raw, 3)`, and $t^2 =$ `r t3_sq`. These are the same number, because they are the same test. The ANOVA table is not giving you a second opinion. It is giving you the first opinion in a different font.

### The Sums of Squares Fight, Live

Earlier I promised you would meet Type III sums of squares. Here it is, and our unequal cells are what make it visible.

`anova()` computes **Type I** (sequential) sums of squares: each term is tested after the terms above it. The `car` package computes **Type II**, where each main effect is tested after the other main effect. With balanced cells these give identical answers and nobody argues. Our cells are not balanced.

```{r anova-types, echo=TRUE}
library(car)

Anova(Model.2, type = 2)
```

Compare the two tables:

| Term | Type I (`anova`) | Type II (`car::Anova`) |
|---|---|---|
| NiceGroup | `r F_nice_I` | `r F_nice_II` |
| GoingOut | `r F_go_I` | `r F_go_II` |
| NiceGroup × GoingOut | `r F_int` | `r F_int_II` |

The interaction is identical, `r F_int` both times. So is Going Out, because it was entered second and was therefore already adjusted for Niceness Group. The one that moves is Niceness Group, from `r F_nice_I` down to `r F_nice_II`, purely because Type I tested it first and handed it credit for variance it shares with Going Out.

Both numbers are correct answers. They are answers to different questions. That is the entire fight, and it exists only because the cells are unequal, which happened only because we dichotomized. The regression coefficients never had this problem: $B_1$ was always the niceness gap within the Stay-in group, unambiguously, and nobody had to pick a Roman numeral.

*This is what I mean when I say regression scales and ANOVA does not. It is the same model either way; one of them just makes you argue about it.*

## Comparison: Continuous × Categorical (the prior chapter) vs. Categorical × Categorical (This chapter)

It is worth pausing to compare the two approaches we have now seen.

| Feature | Prior chapter (Continuous × Categorical) | This chapter (Categorical × Categorical) |
|---|---|---|
| Being Nice | Continuous (0–10) | Categorical (Low vs. High) |
| Centering | Required | Not needed |
| Interaction coefficient | Difference in slopes | Difference of differences |
| Follow-up | `emtrends()`, simple slopes | `pairs()`, simple effects |
| Effect size | Unstandardized slope (*b*) | Cohen's *d* |
| Sample used | All 200 participants | ~80 (extremes only) |

The categorical version uses fewer participants (only the extremes of the niceness distribution), which means less statistical power. The continuous version squeezes more information out of the data because it uses every participant and models the full gradient of niceness effects.

However, the categorical × categorical design is conceptually simpler and maps directly onto the two-way factorial ANOVA framework that many researchers in psychology use. Understanding both is essential.

::: {.callout-important title="An Interaction Is a Difference of Differences"}
First compare the groups inside one condition. Then compare them inside the other condition. The interaction asks whether those two differences are different. If you skip that last comparison, you have described two results but have not tested an interaction.
:::

## The Short Story

- **A 2 × 2 is four cells and four coefficients**, and the four coefficients rebuild the four cell means exactly. $B_0$, $B_0 + B_1$, $B_0 + B_2$, $B_0 + B_1 + B_2 + B_3$.
- **$B_0$ is the reference cell**, not a grand mean. It is one specific corner of the design: both variables at their reference level.
- **$B_1$ and $B_2$ are simple effects, not main effects.** Each is the effect of one variable *at the reference level of the other*. Move the reference level and these numbers change.
- **$B_3$ is the difference of differences, and it is the test.** How much bigger is the niceness gap for the go-out group than for the stay-in group? Nothing else in the table answers that.
- **Follow up with `pairs()` within levels, and report Cohen's *d*.** Simple effects say what the interaction looks like; they do not establish that it exists. The $\Delta R^2$ *F*-test does that, first.
- **It is all the same model.** `anova()` redraws the identical fit as a factorial ANOVA, and the interaction $F$ is just the $B_3$ $t$ squared.

And the punchline from the comparison table, because it is the thing most worth carrying out of these two chapters: **the continuous version uses all of your data.** We got a cleaner-looking 2 × 2 here by throwing away 119 of 200 students. If your predictor is continuous, keep it continuous and use the previous chapter's model. The 2 × 2 is a design you should be able to read fluently, not a shape you should force your data into.
