---
title: "Advanced: Mediation and Indirect Effects"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

## The Question Changes from Whether to How

Mediation asks whether the relationship between $X$ and $Y$ may operate through a third variable, $M$.

$$
X\longrightarrow M\longrightarrow Y.
$$

Suppose a study-skills workshop increases practice time, which then improves exam performance. The proposed mediator is practice time.

Baron and Kenny's classic paper helped make the distinction familiar: a **moderator** changes the strength or direction of a relationship, whereas a **mediator** represents a proposed process through which one variable affects another [@baronkenny1986]. The definitions were useful. The original step-by-step testing ritual has not aged as well.

::: {.callout-warning title="A Mediator Is a Causal Claim Wearing Three Variables"}
Regression can estimate paths consistent with mediation. It cannot prove temporal order, eliminate unmeasured confounding, or show that changing the mediator would change the outcome. Causal mediation requires design and assumptions, not merely three significant coefficients standing in a trench coat.
:::

## The Path Model

The mediator model is:

$$
M=i_M+aX+e_M.
$$

The outcome model is:

$$
Y=i_Y+c'X+bM+e_Y.
$$

The indirect effect is:

$$
ab.
$$

The total effect in a simple linear system is:

$$
c=c'+ab.
$$

The labels mean:

- $a$: expected change in $M$ for a one-unit change in $X$.
- $b$: expected change in $Y$ for a one-unit change in $M$, controlling for $X$.
- $c'$: direct relationship between $X$ and $Y$, controlling for $M$.
- $ab$: indirect effect through $M$.
- $c$: total relationship between $X$ and $Y$ before adding $M$.

For ordinary linear models without interactions and with the same observations, $c=c'+ab$. With nonlinear models, interactions, missing-data differences, or other complications, this simple decomposition may not hold on every scale.

```{r mediation-path-diagram, fig.width=7, fig.height=3.3}
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5))

rect(0.7, 1.8, 2.5, 3.2, col = "#D9EAF7", border = "#1F4E79", lwd = 2)
rect(4.1, 1.8, 5.9, 3.2, col = "#FDE0C5", border = "#D55E00", lwd = 2)
rect(7.5, 1.8, 9.3, 3.2, col = "#D8F0E5", border = "#007F5F", lwd = 2)
text(c(1.6, 5, 8.4), 2.5, c("Workshop\nX", "Practice\nM", "Exam\nY"), cex = 1.05)

arrows(2.55, 2.75, 4.0, 2.75, length = .12, lwd = 2)
arrows(5.95, 2.75, 7.4, 2.75, length = .12, lwd = 2)
arrows(2.45, 1.65, 7.55, 1.65, length = .12, lwd = 2, xpd = NA)
text(3.25, 3.15, "a", cex = 1.2)
text(6.75, 3.15, "b", cex = 1.2)
text(5, 1.22, "c' (direct)", cex = 1.05)
text(5, 4.25, "Indirect effect = a x b", cex = 1.1)
```

## Simulated Workshop Data

```{r mediation-data}
set.seed(343)
n <- 220
MediationData <- data.frame(
  Workshop = rbinom(n, 1, .5)
)

MediationData$Practice <-
  3 + 2.2 * MediationData$Workshop + rnorm(n, 0, 1.6)

MediationData$Exam <-
  58 + 2.5 * MediationData$Workshop +
  3.2 * MediationData$Practice + rnorm(n, 0, 7)
```

```{r mediation-models}
MediatorModel <- lm(Practice ~ Workshop, data = MediationData)
OutcomeModel <- lm(Exam ~ Workshop + Practice, data = MediationData)
TotalModel <- lm(Exam ~ Workshop, data = MediationData)

a <- coef(MediatorModel)["Workshop"]
b <- coef(OutcomeModel)["Practice"]
c_prime <- coef(OutcomeModel)["Workshop"]
c_total <- coef(TotalModel)["Workshop"]

c(a = a, b = b, indirect = a * b, direct = c_prime, total = c_total)
```

## Do Not Require Every Historical Step

The traditional Baron-and-Kenny procedure required a sequence of significant paths, including a significant total effect. Modern mediation analysis focuses on the indirect effect itself [@preacher2008]. A total effect can be small when indirect pathways operate in opposite directions or when the direct and indirect effects cancel.

Historically, researchers were told to:

1. Regress $Y$ on $X$ and require a significant total effect $c$.
2. Regress $M$ on $X$ and require a significant $a$ path.
3. Regress $Y$ on $X$ and $M$ and require a significant $b$ path.
4. Declare "complete" mediation if $c'$ became nonsignificant, or "partial" mediation if it merely became smaller.

These regressions remain useful for understanding the system. The gatekeeping rules are the problem. A nonsignificant total effect does not prohibit an indirect effect. The direct and indirect effects can point in opposite directions and cancel. Significance also depends on sample size, so making "complete mediation" depend on a nonsignificant $c'$ rewards low power.

::: {.callout-note title="The Indirect Effect Is the Target"}
Do not infer mediation merely because $a$ and $b$ are individually significant. Test and report $ab$ with an interval that reflects its skewed sampling distribution.
:::

### Joint Significance and the Sobel Test

If both $a$ and $b$ are clearly different from zero, their joint significance provides evidence consistent with an indirect effect. However, it does not directly estimate the uncertainty of $ab$.

The Sobel approach uses a delta-method standard error. A common version is

$$
SE_{ab}\approx\sqrt{b^2SE_a^2+a^2SE_b^2},
$$

and then tests $z=ab/SE_{ab}$. The problem is that a product of two approximately normal coefficients is usually not normal, especially when either path is near zero. The resulting interval is forced to be symmetric when the sampling distribution is often skewed. This is why bootstrap intervals largely replaced the Sobel test for ordinary mediation [@preacher2008].

## Bootstrap the Indirect Effect

The product $ab$ is often asymmetric, especially in smaller samples. A bootstrap interval avoids forcing the product into a symmetric normal approximation.

```{r mediation-bootstrap, fig.width=7, fig.height=4.5}
set.seed(344)
B <- 4000

IndirectBoot <- replicate(B, {
  rows <- sample(seq_len(nrow(MediationData)), replace = TRUE)
  d <- MediationData[rows, ]
  a_star <- coef(lm(Practice ~ Workshop, data = d))["Workshop"]
  b_star <- coef(lm(Exam ~ Workshop + Practice, data = d))["Practice"]
  a_star * b_star
})

quantile(IndirectBoot, c(.025, .975))

hist(
  IndirectBoot, breaks = 35,
  col = "#9ECAE1", border = "white",
  xlab = "Bootstrapped indirect effect",
  main = "The indirect effect gets its own sampling distribution"
)
abline(v = 0, lty = 2, lwd = 2)
abline(v = a * b, col = "#D55E00", lwd = 3)
```

The code resamples complete rows so each participant's workshop, practice, and exam values stay together. In clustered or repeated data, resample at the appropriate higher-level unit or use a model-based procedure. Do not split a participant into three independent ghosts.

### Percentile and BCa Intervals

The histogram above supports a percentile interval by taking its 2.5th and 97.5th percentiles. A BCa interval additionally adjusts for median bias and acceleration. Neither method rescues a badly identified causal model. See the advanced bootstrapping chapter for the jackknife, percentile, basic, studentized, and BCa logic.

### The `mediation` Package Vocabulary

The `mediation` package describes the indirect effect as the **average causal mediation effect** (ACME), the direct effect as the **average direct effect** (ADE), and their sum as the total effect. Those names emphasize causal estimands, which means the assumptions deserve more attention, not less.

```{r mediation-package, eval=FALSE}
library(mediation)

MediationFit <- mediate(
  model.m = MediatorModel,
  model.y = OutcomeModel,
  treat = "Workshop",
  mediator = "Practice",
  boot = TRUE,
  boot.ci.type = "bca",
  sims = 5000
)

summary(MediationFit)
plot(MediationFit)
```

Five thousand simulations are used here for a final analysis. A few hundred are acceptable while testing code, but bootstrap tail endpoints dislike being estimated from spare change.

## What Must Be True for a Causal Interpretation?

At minimum, we need defensible assumptions about:

- Temporal order: $X$ precedes $M$, which precedes $Y$.
- No unmeasured confounding of $X\rightarrow M$.
- No unmeasured confounding of $M\rightarrow Y$.
- No mediator–outcome confound created by $X$.
- Correct functional forms and measurement.
- No major post-treatment selection problems.

Randomizing $X$ helps with paths beginning at $X$. It does not automatically randomize the mediator. Participants who practice more may differ in motivation, sleep, prior knowledge, terror of disappointing their mother, or several variables we did not measure.

## Partial and Complete Mediation

Avoid treating “complete mediation” as a trophy awarded when $c'$ becomes nonsignificant. Significance depends on sample size and uncertainty. Report the direct effect, indirect effect, total effect, and their intervals. Let the estimates describe the pattern.

The **proportion mediated**, often calculated as $ab/c$, can be unstable or nonsensical when the total effect is close to zero or when direct and indirect effects have opposite signs. Report it only when the decomposition and scale make it meaningful. A ratio exploding because its denominator is nearly zero is algebra having a tantrum, not a discovery.

## Binary Mediators or Outcomes

The mediator and outcome models do not both have to be ordinary linear regressions. A binary mediator can use logistic regression, and a binary outcome can use a binomial model.

```{r binary-mediator-example}
set.seed(346)
n_binary <- 600
BinaryMediationData <- data.frame(
  Workshop = rbinom(n_binary, 1, .5)
)

BinaryMediationData$AskedForHelp <- rbinom(
  n_binary,
  1,
  plogis(-1 + 1.1 * BinaryMediationData$Workshop)
)

BinaryMediationData$Exam <-
  62 +
  2 * BinaryMediationData$Workshop +
  7 * BinaryMediationData$AskedForHelp +
  rnorm(n_binary, 0, 8)

BinaryMediatorModel <- glm(
  AskedForHelp ~ Workshop,
  data = BinaryMediationData,
  family = binomial()
)

BinaryOutcomeModel <- lm(
  Exam ~ Workshop + AskedForHelp,
  data = BinaryMediationData
)

summary(BinaryMediatorModel)
summary(BinaryOutcomeModel)
```

With nonlinear links, $ab$ formed from two raw coefficients may not equal a simple indirect effect on the probability or outcome scale. Use software that defines the estimand, calculates predictions on a meaningful scale, and states whether effects are conditional or averaged. Do not multiply a log-odds coefficient by a raw-score coefficient and declare the units “advanced.”

## Power for an Indirect Effect

Power depends jointly on paths $a$ and $b$, their standard errors, the distributions of the variables, collinearity, measurement reliability, and the resampling method. A study powered for the total $X\rightarrow Y$ association is not automatically powered for $ab$.

Because the product distribution is awkward, Monte Carlo simulation is often the clearest planning method:

1. Specify plausible values for $a$, $b$, $c'$, residual variances, and measurement reliability.
2. Simulate data under those assumptions.
3. Fit the planned mediator and outcome models.
4. Construct the planned indirect-effect interval.
5. Repeat the entire study many times and count how often the interval excludes zero.

Run several plausible scenarios. Power calculated from one heroic guessed effect size is optimism wearing a spreadsheet.

::: {.callout-important title="Measurement Error Attacks Mediation Twice"}
Poor measurement weakens the estimated $a$ or $b$ path and can distort the decomposition. Because the indirect effect is a product, damage to either path damages $ab$. Reliable measures and defensible timing are part of the power analysis, even when the software pretends the only input is $N$.
:::

## Moderated Mediation

An indirect effect may depend on another variable $W$. For example, the workshop may increase practice more strongly among students with enough available time:

$$
M=i_M+a_1X+a_2W+a_3XW+e_M.
$$

The conditional indirect effect becomes:

$$
(a_1+a_3W)b.
$$

This is not a separate magical procedure. It combines mediation with an interaction, then evaluates the indirect effect at meaningful values of the moderator.

The [moderated mediation chapter](Chapter_Moderated_Mediation.qmd) develops first-stage, second-stage, direct-path, and combined moderation; conditional indirect effects; bootstrap contrasts; and the index of moderated mediation.

::: {.callout-important title="More Arrows Require More Theory"}
Every added mediator, moderator, and covariate creates additional assumptions and alternative explanations. A path diagram is not evidence merely because the arrows are straight and the font is attractive.
:::

## Identification: What the Three Regressions Cannot See

For the indirect effect to support a causal mediation story, we need strong assumptions about confounding of the $X\rightarrow M$, $X\rightarrow Y$, and $M\rightarrow Y$ relations. We also need no mediator-outcome confound that was itself caused by $X$. The last condition is especially rude because controlling for the post-treatment confound can create additional bias.

Randomizing $X$ protects comparisons beginning at $X$. It does not randomize the mediator. If motivated students both practice more and perform better, the estimated $b$ path can mix the mediator with motivation. A beautiful indirect-effect interval does not reveal the missing variable.

## Competing Mediators and Temporal Order

With several mediators, the total indirect effect can be divided into specific indirect effects only under additional assumptions. Parallel mediators may influence one another; serial mediators require a defensible order. Measuring every variable in one survey and drawing arrows from left to right does not create time. Longitudinal measurement, experimental manipulation of mechanisms, and sensitivity analysis make the argument stronger.

## Mediation as a Structural Equation Model

Structural equation modeling (SEM) writes the same regressions as one system. In a path diagram:

- Squares usually represent observed variables.
- Circles usually represent latent variables.
- One-headed arrows represent regressions.
- Two-headed arrows represent variances or covariances.
- A triangle labeled 1 may represent a constant or intercept.

For simple mediation with observed variables, SEM does not magically change the scientific assumptions. Its advantages appear when we estimate several equations together, test multiple mediators, represent latent variables, handle correlated residuals, or impose equality constraints.

In `lavaan`, labels define paths and `:=` defines derived parameters:

```{r lavaan-mediation, eval=FALSE}
library(lavaan)

MediationSyntax <- '
  # Regressions
  Practice ~ a*Workshop
  Exam ~ cprime*Workshop + b*Practice

  # Derived effects
  indirect := a*b
  direct := cprime
  total := cprime + (a*b)
'

MediationSEM <- sem(
  model = MediationSyntax,
  data = MediationData,
  se = "bootstrap",
  bootstrap = 5000
)

summary(
  MediationSEM,
  standardized = TRUE,
  rsquare = TRUE,
  fit.measures = TRUE
)
```

For a just-identified three-variable path model, global fit indices may be uninformative because the model reproduces the available covariance information exactly. Adding the words *structural equation model* does not provide a free model-fit halo.

## The Short Story

Mediation estimates an indirect effect $ab$. Bootstrap its uncertainty, report direct and total effects, and separate “the data are consistent with this pathway” from “we proved this causal mechanism.” The second claim requires much more than regression output.
