---
title: "Pearson's Chi-Square"
fontsize: 11pt
geometry: margin=.5in
format:
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
    include-in-header:
      text: |
        \usepackage{amsmath}
---
```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(fig.width=4.25)
knitr::opts_chunk$set(fig.height=4.0)
knitr::opts_chunk$set(fig.align='center') 
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(results='hold') 
```


## Parametric vs. Non-Parametric

Statistical tests are often divided into parametric and nonparametric families. The difference concerns what the model assumes and what part of the data it compares.

### Parametric Tests (coming later)

Parametric tests describe populations with parameters such as means and variances. Particular tests make particular distributional and error assumptions; they do not all demand that the raw scores themselves be perfectly normal. They are powerful when their assumptions are reasonable.

You already know the ingredients (mean, SD) — later we’ll formalize how they connect to tests like t and F.

### Nonparametric Tests (this chapter and the next)

Nonparametric tests make fewer or different distributional assumptions and often ask questions about counts, signs, or ranks:

- They usually do not require normally distributed raw scores
- Work with ordinal or nominal data
- Also useful when parametric assumptions are violated
- Instead of means, they compare counts/frequencies or rank observations

They are not assumption-free. Independence, sampling, measurement, and the design still matter. The next chapter examines that entire nonparametric family in more detail.

### Chi-Square

Pearson's chi-square is often placed under the “nonparametric” umbrella because it does not model means with a normal error distribution. More precisely, it is a test for **categorical counts**. It is not the emergency version of a t-test to use whenever normality looks annoyed.

Rather than asking: “How far is this group mean from another group mean?”

Chi-square asks: “Do the observed counts differ from what we would expect by chance?”

In technical terms, chi-square hypothesis testing asks whether an observed pattern is consistent with a chance-based null hypothesis.

- We compare observed data to expected values under null hypothesis ($H_0$), which means we expect no difference from chance.
- Large discrepancies lead us to reject $H_0$ and provide support for the alternative hypothesis ($H_1$). 

*Note:* We will return the logic of hypothesis testing in more detail later in the semester.  

Chi-square is especially useful when:

- Outcomes are categories (yes/no, group A/B/C, correct/incorrect)
- Means and standard deviations either don’t exist or don’t make sense

There are two types of chi-square:

- One-Way Pearson's Chi-Square [Goodness of Fit]
- Two-Way Pearson's Chi-Squared [Test of Independence] 

## One-Way Pearson's Chi-Square [Goodness of Fit]

Have you been told: "Stay with your first answer on a multiple-choice test."  *So is changing answers more likely to be harmful?*

> Best studied the responses of 261 students in an introductory psychology course [@best1979]. He recorded the number of right-to-wrong (27), wrong-to-right (195), and wrong-to-wrong (39) answer changes for each student. Note: We will ignore wrong-to-wrong as he did in his first analysis.

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  27            | 195

Total of 222 observations

We will use the **One-Way Chi-Squared** [Goodness of fit]. It asks whether the relative frequencies observed in the categories of a sample frequency distribution are in agreement with the relative frequencies hypothesized to be true in the population.

### Hypothesis for Goodness of fit

There are two versions (but it does not change the math, just how you talk about it): 

- No preference: there will be no preference between different categories (e.g., do people prefer Coke or Pepsi)
-	No difference from a known population: This type hypothesis compares two populations (or representative samples)

If people were making chance guesses from a population of people who don't have any knowledge about material on the test (version 2 of the hypothesis):

- $H_0: P_{right-to-wrong} = .5, P_{wrong-to-right} = .5$ 
- $H_1: H_0$ is False 

*Note:* The probability you set can be whatever you want, but across all cells, they must add to 1. When you do not know what probability should be from a theoretical standpoint, you can simply do 1/Number of cells. 

### Find Expected Frequencies

If we expect by chance people would have gotten p = .50 on wrong-to-right and right to wrong, $f_e = pn$  

$F_{e_{right-to-wrong}}= .5*222=111$
$F_{e_{wrong-to-right}}= .5*222=111$

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  27            | 195
Expected  |  111           | 111

*Note:* These tables are often shown in papers (sometimes as proportions).

### Goodness of Fit Calculation by Hand

$$\chi^2 = \sum \frac{(f_o-f_e)^2}{f_e}$$

$\chi^2 = \frac{(27-111)^2}{111}+\frac{(195-111)^2}{111} = 127.14$

*Note:* Never enter proportions into the chi-square. You must work with frequencies.

### Test Against Chance Distribution

$\chi^2$ values close to zero are probably “chance,” and values that are larger are farther from chance.
Wait—how do I know that?

Look at our $f_o$ values. If the $f_o$ values were close to the $f_e$ values, then the total value we calculate would be very close to zero.

For example, if we actually found this pattern instead:

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  111           | 111
Expected  |  111           | 111

$\chi^2 = \frac{(111-111)^2}{111}+\frac{(111-111)^2}{111} = 0$

While it’s very rare for the observed frequencies to be exactly what we expect by chance, it’s often really close to zero.

To get a sense of how often it would be close to zero, imagine we created a new test and the multiple-choice question said:

> “Please select an answer, then erase it and select something else. There is one right answer.”

That’s all you get—no actual knowledge being tested. It’s a pure game of chance! In other words, people are totally guessing and are forced to change their answer.

Sometimes, purely by chance, they will go from Right-to-Wrong or Wrong-to-Right (and we will ignore all the Wrong-to-Wrong cases). What would that look like?

I can run 10,000 subjects to get a sense of what that might look like.

```{r}
#| echo: false
set.seed(42)
x <- rchisq(1e4, 1)
hist(x, prob=FALSE, main="Chance Histogram",
  xlab="Chi-square value", ylab="Number", xlim=c(0,20),breaks =5000)
#lines(density(x), col='red')
```

It looks like we get zero and values close to zero a lot—but also, purely by chance, we can get values all the way out to about 17.25. So even by chance, we can get a value that’s “far” from zero.

But does that happen often?

Before we answer that question, let’s zoom in closer to zero so we can see what’s going on more clearly.

```{r}
#| echo: false
hist(x, prob=FALSE, main="Chance Histogram",
  xlab="Chi-square value", ylab="Number", xlim=c(0,.25), breaks =5000)
#lines(density(x), col='red')
```
So how far from zero is still “chance”?

To answer that question, we would need more than 10,000 people. In fact, we would need the entire population. That would be impossible, but luckily, someone has done the math-magic to figure out what that would look like: 

```{r}
#| echo: false
curve(
  dchisq(x, df = 1),
  from = 0,
  to = 20,
  main = "Chi-Square Chance Distribution (df = 1)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)
```
Earlier, we looked at frequencies—how often different chi-square values occurred in a simulation.

Here, we switch to a probability density, which describes how likely different chi-square values are under the null hypothesis.

- The x-axis shows possible chi-square values.
- The y-axis shows relative likelihood (how often it will occur), not counts.

The important idea is that:

- Values near zero are very likely under chance
- Larger values are increasingly unlikely under chance

The total area under the curve equals 1, representing all possible outcomes under the null hypothesis.

### How do we set chance? 

How extreme does a chi-square value need to be before we stop believing it’s just chance?

One very common rule is the 95% / 5% rule.

>We say: “If the null hypothesis is true, then 95% of the time our chi-square values should fall below some cutoff.”

That cutoff defines the critical value. Any value beyond that point is rare enough that we start to doubt the chance explanation.

This does not mean:

- That chance is impossible past that value
- Or that we are 100% certain the result is “real”

It means:

- About 5% of the time, even when the null hypothesis is true, we will still see values this extreme by chance alone
- We accept that risk in exchange for having a decision rule

So when we say a result is “unlikely due to chance,” what we really mean is:

> “If this were just chance, we would expect to see something this extreme only about 5% of the time.”

```{r}
#| echo: false
df <- 1
crit <- qchisq(0.95, df)

curve(
  dchisq(x, df),
  from = 0,
  to = 20,
  main = "Chi-Square Chance Distribution (df = 1)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)

# Shade critical region (top 5%)
x_vals <- seq(crit, 20, length.out = 1000)
y_vals <- dchisq(x_vals, df)
polygon(
  c(x_vals, rev(x_vals)),
  c(y_vals, rep(0, length(y_vals))),
  col = rgb(1, 0, 0, 0.3),
  border = NA
)

# Add critical value line
abline(v = crit, lty = 2, lwd = 2)

# Label
text(
  crit + 1, 
  max(dchisq(x_vals, df)) * 0.6,
  labels = "5% of chance values\n(critical region)",
  adj = 0
)

```

The distribution changes as a function of the degrees of freedom [$df$ is the number of cells - 1]. Degrees of freedom tell us how much independent information is available to estimate variability in the data. If we have diferent number of conditions, we have a different distribution to compare it.  

```{r}
Crit.MCQ=qchisq(.05, df=1, lower.tail = FALSE)
```

We get the crit = `r Crit.MCQ`. Any $\chi^2$ we calculate above that value means it has only a 5% probability of being a chance result.  

### Our results? 

We ask where is our chi-square value relative to this chance distribution given our $df = 1$.  We calculated 127.14, and that is in the tail (the small part, well above chance value). We can use R or a table from a book to get the critical values for alpha (p) =.05. We don't need to make the graph every time as someone did this already using calculus and has given us all the values with a simple function in R called `qchisq`.  

This function answers the question:

- “What chi-square value cuts off a certain percentage of chance outcomes?”
  - p = .05
- We are defining the 5% region — the part of the distribution that happens only about 5% of the time under chance.

- df = 1
  - This tells R which chi-square distribution to use.
    - With two categories, the degrees of freedom is 1.

- lower.tail = FALSE
  - This is important:
    - FALSE means we want the upper tail of the distribution

```{r}
Crit.MCQ<-qchisq(p = .05, df=1, lower.tail = FALSE)
```

We get the crit = `r Crit.MCQ` < 124.14 (experimental value we got), so we reject the null hypothesis ($H_0$) and provide evidence for the alternative ($H_1$).    

### Report in APA

$\chi^2(df,N) = X.XX, p < or >= .XX$

Since our result was p-value < .05, as rejected our null. 

Students were significantly more likely to change an answer from wrong to right (87.83%) than from right to wrong (12.16%), $\chi^2(1, N = 222) = 124.14, p < .05$.

### Impact of higher DF

Note: shape of chance distribution changes as we have more df (say 6 cells to compare)

```{r}
#| echo: false
df <- 5
crit <- qchisq(0.95, df)

curve(
  dchisq(x, df),
  from = 0,
  to = 20,
  main = "Chi-Square Chance Distribution (df = 5)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)

# Shade critical region (top 5%)
x_vals <- seq(crit, 20, length.out = 1000)
y_vals <- dchisq(x_vals, df)
polygon(
  c(x_vals, rev(x_vals)),
  c(y_vals, rep(0, length(y_vals))),
  col = rgb(1, 0, 0, 0.3),
  border = NA
)

# Add critical value line
abline(v = crit, lty = 2, lwd = 2)

# Label
text(
  crit + 1, 
  max(dchisq(x_vals, df)) * 0.6,
  labels = "5% of chance values\n(critical region)",
  adj = 0
)

```

As you can see the distribution gets more spread out. So with larger DF our critical value increases

```{r}
#| echo: true
ChiCrit<-qchisq(.05, 1:8, lower.tail = FALSE)
plot(ChiCrit, main="Chi-Square Critical Values: alpha = .05",
  xlab="df", ylab="Critical Value", xlim=c(1,8), ylim=c(0,25))
```

### Goodness of Fit Calculation by Code

The way above if the hard way and shown you understand what the computer is doing. Instead, doing one-way chi-Squares in R is really easy! 

### Step 1: Build your Table

```{r}
#| echo: true
MCQ.Study <- as.table(rbind(c(27, 195)))
dimnames(MCQ.Study) <- list(Freq = c("Observed"),
                           Action = c("RtoW", "WtoR"))
MCQ.Study
```

### Step 2: Calculate Chi-squared

```{r}
#| echo: true
MCQ.chi <- chisq.test(MCQ.Study, p=c(.5, .5))
MCQ.chi
```

Again, in APA format, we would say, 

Students were significantly more likely to change an answer from wrong to right (87.83%) than from right to wrong (12.16%), $\chi^2(1, N = 222) = 124.14, p < .05$.



## Two-Way Pearson's Chi-Squared [Test of Independence] 

"Maybe we should only change our answers on easy items, but on hard items, we should trust ourselves."

> Best (1979) also examined the 1670 total number of responses across the 261 students and divided the items into easy/difficult in an introductory psychology course.  He recorded the number of right-to-wrong, wrong-to-right, and wrong-to-wrong (39) answer changes for each student. Again we will ignore wrong-to-wrong (17% of the responses) for simplicity.

Frequency | Right to Wrong | Wrong to Right 
---------|----------------|----------------
Easy     |  97            | 411
Difficult|  251           | 620

We need to do a Two-Way Chi-Square [Test of Independence]. It asks whether observed frequencies reflect the independence of two qualitative variables. Compares the actual observed frequencies of some phenomenon (in our sample) with the frequencies we would expect if there were no relationship at all between the two variables in the larger (sampled) population. We ask if the two variables are independent: knowledge of the value of one variable provides no information about the value of another variable.

### Hypothesis for Test of Independence

The null hypothesis is that for each, the value obtained for one variable is not related to or influenced by the second variable.  This idea can be expressed in two versions:

- Version 1: A single sample is measured on two separate variables. For example, knowing your personality will help predict your color preference (alternative). Knowing your personality will NOT help you predict your color preference (null) [Personality and color preference are measured per person]
    -	Null: The variables are not related. 
    -	Alternative: The variables are related. 

- Version 2: Two (or more) samples are measured and compared to see if there are differences. For example, the same proportions of extroverts (group 1) and introverts (group 2) prefer the same color (null).  If the proportions are different for the two groups, then you have the alternative hypothesis
    -	Null: The samples are independent 
    -	Alternative: The samples are not independent

Here we are working with Version 1 (one group taking two types of items)

### Null Hypothesis ($H_0$)

- Answer-changing behavior is independent of item difficulty.
- The proportion of right-to-wrong and wrong-to-right changes is the same for easy and difficult items

Knowing whether an item is easy or difficult provides no information about how answers are changed

### Alternative Hypothesis ($H_1$)

Answer-changing behavior depends on item difficulty.

- Students are more likely to benefit from changing answers on easy items

- On difficult items, answer changes are less likely to improve performance

Knowing whether an item is easy or difficult does provide information about how answers are changed

Simply put: 

- $H_0:$ The item difficulty is not related to how a person changes their answer 
- $H_1:$ The item difficulty is related to how a person changes their answer 

### Find Expected Frequencies

This is little more complex than the goodness of fit test

### Find row/col sums

Frequency | Right to Wrong | Wrong to Right | Sum
---------|----------------|----------------|----
Easy     |  97            | 411            | **508**
Difficult|  251           | 620            | **871**
Sum      |  **348**       | **1031**       | **1379** 

### Step 2: Expected frequencies per cell

$$f_e = \frac{col\,total*row\,total}{overall\,total}$$

Frequency | Right to Wrong | Wrong to Right | Sum
---------|----------------|----------------|----
Easy     |  97 (128.1972) | 411 (379.8028) | **508**
Difficult|  251 (219.8028)| 620 (651.1972) | **871**
Sum      |  **348**       | **1031**       | **1379** 


### Test of Independence Calculation by Hand

$$\chi^2 = \sum \frac{(f_o-f_e)^2}{f_e}$$

$\chi^2 = \frac{(97-128.1972)^2}{128.1972} + \frac{(411-379.8028)^2}{379.8028} +\frac{(251-219.8028)^2}{219.8028}+\frac{(620-651.1972)^2}{651.1972}$

$\chi^2 = 16.08$

### Test Against Distribution

$df = (Row-1)(Col-1)$
$df = (2-1)(2-1) = 1$

Given our $df = 1$.  We calculated 16.08 we can use R or a table to get the critical values for alpha = .05 

```{r}
Crit.MCQ=qchisq(.05, df=1, lower.tail = FALSE)
```

We get the crit = `r Crit.MCQ` < 16.08, so we reject the null.   

### Report in APA

$\chi^2(df,N) = X.XX, p < or >= .XX, V = .XX$ 

In APA format:

Item difficulty was related to how students changed their answers, $\chi^2(1, N = 1379) = 16.08, p < .05$.

[That is all we can say statistically.  Normally people present a table in percentages to unpack this result, **but** You have to decide if you want to do row or column percentages. People do not present both]. I think the row percent make more sense in this case we want to compare across item difficulty. 

% Row    | Right to Wrong | Wrong to Right | %
---------|----------------|----------------|----
Easy     |  19.1%         |  80.9%         | 100%
Difficult|  28.8%         |  71.2%         | 100%
Col Total|  25.2%         |  74.8%         | 100% 


### Test of Independence Calculation by Code

Again the code for R is very easy, so I will show you how it works: 

### Step 1: Build Your table

```{r}
#| echo: true
MCQ.Study.2 <- as.table(rbind(c(97, 411), c(251, 620)))
dimnames(MCQ.Study.2) <- list(Social = c("Easy","Diff"),
                              Change = c("RtoW", "WtoR"))
                             
MCQ.Study.2
```

### Step 2: Calculate chi-square

In tests of independence, R calculates the expected frequencies for you. For a $2 \times 2$ table, `chisq.test()` applies **Yates's continuity correction** by default when `correct = TRUE`. The correction tries to improve a continuous chi-square approximation to discrete counts, but it can be overly conservative. It is not a ritual that must always be performed. Use it deliberately, report whether you used it, and consider Fisher's exact test when expected counts are sparse.

$$\chi^2_{Yates} = \sum \frac{(|f_o-f_e| - .5)^2}{f_e}$$

*Note*: You can view the observed and expected frequencies. For larger sparse tables, `chisq.test(..., simulate.p.value = TRUE)` can estimate a p-value by simulation.

```{r}
#| eval: true
#| echo: true
MCQ.chi.2 <- chisq.test(MCQ.Study.2,correct = TRUE)
MCQ.chi.2
```


### Row Proportion Code
```{r}
#| eval: true
#| echo: true
prop.table(MCQ.Study.2,1)
```

### Col Proportion Code
```{r}
#| eval: true
#| echo: true
prop.table(MCQ.Study.2,2)
```


### Effect Size: Because an Asterisk Is Not a Personality

The chi-square test tells us whether the observed table is difficult to explain under the null hypothesis. It does **not** tell us whether the association is large enough to matter. With enough observations, a microscopic association can earn an asterisk and begin acting important at faculty meetings.

For a $2 \times 2$ table, **phi** ($\phi$) is:

$$
\phi = \sqrt{\frac{\chi^2}{N}}
$$

For a table of any size, use **Cramer's V**:

$$
V = \sqrt{\frac{\chi^2}{N\min(r-1,c-1)}}
$$

Here, $r$ and $c$ are the numbers of rows and columns. Cramer's V ranges from 0, meaning no association, to 1, meaning perfect association. Do not automatically translate `.10`, `.30`, and `.50` into small, medium, and large without thinking about the design and research area. Those labels are rough landmarks, not sacred commandments delivered on stone tablets by Cohen.

```{r}
#| echo: true
chi_uncorrected <- chisq.test(MCQ.Study.2, correct = FALSE)
n_total <- sum(MCQ.Study.2)

phi <- sqrt(as.numeric(chi_uncorrected$statistic) / n_total)
cramers_v <- sqrt(
  as.numeric(chi_uncorrected$statistic) /
    (n_total * min(dim(MCQ.Study.2) - 1))
)

c(phi = phi, cramers_v = cramers_v)
```

For these data, $V = .11$: there is an association, but it is not a gigantic one. The asterisk has not transformed into a flaming sword.

### Odds Ratios for a 2-by-2 Table

An **odds ratio** compares the odds of an outcome between two groups. In our table, the odds of changing an easy item from right to wrong rather than wrong to right are $97/411$. For difficult items, those odds are $251/620$.

$$
OR = \frac{97/411}{251/620} = 0.58
$$

An odds ratio of 1 would mean equal odds. Here, the odds of a right-to-wrong rather than wrong-to-right change were about 42% lower for easy items than for difficult items. If you reverse the comparison, the odds for difficult items are about $1/0.58 = 1.71$ times those for easy items. Odds ratios are direction-sensitive little creatures, so always say which group and outcome are in the numerator.

Base R gives us the odds ratio and its confidence interval with Fisher's exact test:

```{r}
#| echo: true
fisher_result <- fisher.test(MCQ.Study.2)
fisher_result$estimate
fisher_result$conf.int
```

The estimated odds ratio is 0.58, 95% CI [0.44, 0.77].

### Power for Chi-Square

Power asks how often a study would detect an association of a specified size if that association really exists. Small samples give fickle chance more room to cause trouble. A weak association plus twelve people and one undergraduate who clicked every response while watching TikTok is not a sturdy research program.

For chi-square tests, Cohen's $w$ is commonly used in power calculations. In a $2 \times 2$ table it is equivalent to phi and Cramer's V. The function below uses only base R. It finds the total sample size needed for 80% power when $w = .1062$, approximately the association observed here.

```{r}
#| echo: true
chisq_power <- function(n, w, df, alpha = .05) {
  critical_value <- qchisq(1 - alpha, df = df)
  1 - pchisq(critical_value, df = df, ncp = n * w^2)
}

required_n <- uniroot(
  function(n) chisq_power(n, w = .1062, df = 1) - .80,
  interval = c(2, 10000)
)$root

ceiling(required_n)
```

The answer is about 696 total observations. That number is not permission to choose an effect size after seeing the results. Plan power using the smallest effect worth detecting, prior evidence, or a defensible range of possible effects.

### Reporting the Result

A compact report can include the test, effect size, and a table of interpretable percentages:

> Item difficulty was associated with answer-change direction, $\chi^2(1, N = 1379) = 15.57$, $p < .001$, $V = .11$. Compared with difficult items, easy items had lower odds of being changed from right to wrong rather than wrong to right, $OR = 0.58$, 95% CI [0.44, 0.77].

The chi-square value above uses Yates's correction, matching the test already run in this chapter. The effect-size calculation uses the ordinary Pearson chi-square. State your choices so readers do not have to conduct forensic accounting on your results section.

### Chi-Square Issues

- Chi-square works with **counts**, not percentages, means, or the emotional intensity with which you typed the numbers.
- The observations must be independent. One person should not quietly appear in several cells unless the design and model explicitly handle repeated observations.
- The approximation becomes unreliable when expected counts are very small. A common guideline is that no expected count should be below 1 and no more than 20% should be below 5, but this is a guideline rather than a commandment. For a small $2 \times 2$ table, use Fisher's exact test. For larger sparse tables, a simulated p-value may be useful.
- Chi-square cannot be negative because every discrepancy is squared. It can equal zero when every observed count equals its expected count, an event normally witnessed immediately before the Probability Gods demand another sacrifice.
- A cell contributes more when its observed and expected counts differ by a large amount **relative to the expected count**.
- A significant omnibus test says that some association exists. It does not identify every cell responsible for it. Follow-up comparisons create a multiple-testing problem, which the advanced chapter later in this book handles before the asterisks multiply and form a union.

::: {.callout-important title="Counts First; Interpretation Second"}
A chi-square test asks whether the observed **counts** differ from the counts expected under the null model. After that test, inspect the percentages, residuals, and effect size to learn where the association lives and whether anybody should care.
:::
