---
title: "Moderated Mediation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 3.75,
  fig.align = "center"
)
```

::: {.callout-warning title="Historical Lecture"}
This chapter has not yet been adapted to match the format and style of the rest of the book. It is included in its original lecture structure for readers who need to conduct and interpret a moderated mediation analysis.
:::

## Moderated Mediation

- Moderation examines factors that influence the strength of a relationship between $X$ and $Y$.
- Mediation examines an intermediate pathway connecting $X$ and $Y$.
- Moderated mediation asks whether an indirect effect depends on a moderator.
- In practical terms, what if parenting style moderated the mediation through trust in authority from the previous chapter?
- Strict parents might foster respect for authority, whereas permissive parents might foster opposition to authority.
- We will measure parenting style from -3 to 3, ranging from permissive to authoritative.
- Our hypothesis is that the indirect effect through trust is stronger for children with authoritative parents than for children with permissive parents.
- Children with authoritative parents might also be more well-behaved in general, so we will examine moderation of both direct and indirect paths.

```{r}
library(mediation)

set.seed(42)

# Path a strength
a <- .4

# Path b strength
b <- .4

# Path c' strength
cp <- .01

# People
n <- 268

# Normal distribution of time (minutes)
X <- rnorm(n, 5, 2)

# Moderator
Mod <- runif(n, -3, 3)

# Mediator
M <- a * X * Mod + rnorm(n, 0, 1)

# Equation used to create Y
Y <- cp * X * Mod + b * M + M * Mod + rnorm(n, sd = 1)

# Build the data frame
Marshmallow.Mod <- data.frame(
  Success = Y,
  Time = X,
  Trust = M,
  Parents = Mod
)
```

### Moderation on Paths $a$, $b$, and $c'$

- In this example, we think that parenting style moderates the direct path ($c'$) and both parts of the indirect path ($a$ and $b$).
- We therefore include interactions with the moderator in both models.
- The mediator model becomes $M \sim X * Mod$.
- The outcome model becomes $Y \sim M * Mod + X * Mod$.
- This corresponds to Hayes's PROCESS Model 59.

```{r}
Mod.Med.Model.2 <- lm(
  Trust ~ Time * Parents,
  data = Marshmallow.Mod
)
summary(Mod.Med.Model.2)

Mod.Med.Model.3 <- lm(
  Success ~ Trust * Parents + Time * Parents,
  data = Marshmallow.Mod
)
summary(Mod.Med.Model.3)
```

- In the `mediation` package, we list the moderator as a covariate and specify the values at which we want to estimate the effects.
- We will use one standard deviation below and above the mean. We could also use zero to examine the effect at the average parenting-style score.
- This allows us to examine the conditional direct and indirect effects.
- Let's look at permissive parents first.

```{r}
Permissive <- mean(Marshmallow.Mod$Parents) - sd(Marshmallow.Mod$Parents)
Permissive

Mod.Med.Boot.BCa.1 <- mediate(
  Mod.Med.Model.2,
  Mod.Med.Model.3,
  covariates = list(Parents = Permissive),
  boot = TRUE,
  boot.ci.type = "bca",
  sims = 200,
  treat = "Time",
  mediator = "Trust"
)
summary(Mod.Med.Boot.BCa.1)
plot(Mod.Med.Boot.BCa.1)
```

- Now, let's look at authoritative parents.

```{r}
Authoritative <- mean(Marshmallow.Mod$Parents) + sd(Marshmallow.Mod$Parents)
Authoritative

Mod.Med.Boot.BCa.2 <- mediate(
  Mod.Med.Model.2,
  Mod.Med.Model.3,
  covariates = list(Parents = Authoritative),
  boot = TRUE,
  boot.ci.type = "bca",
  sims = 200,
  treat = "Time",
  mediator = "Trust"
)
summary(Mod.Med.Boot.BCa.2)
plot(Mod.Med.Boot.BCa.2)
```

- To test whether the conditional direct or indirect effects differ across moderator values, we can use `test.modmed()`.

```{r}
Mod.Med.Boot.BCa.3 <- mediate(
  Mod.Med.Model.2,
  Mod.Med.Model.3,
  boot = TRUE,
  boot.ci.type = "bca",
  sims = 200,
  treat = "Time",
  mediator = "Trust"
)

test.modmed(
  Mod.Med.Boot.BCa.3,
  covariates.1 = list(Parents = Permissive),
  covariates.2 = list(Parents = Authoritative),
  sims = 200
)
```

- In this simulation, the conditional indirect effects differ, but the conditional direct effects do not.
- The data are therefore consistent with the indirect effect being moderated by parenting style in the predicted direction.
- The 200 bootstrap draws keep this lecture example quick. Use substantially more draws for a real analysis.
