---
title: "Generalized Linear Models"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 3.75,
  fig.align = "center"
)
```

::: {.callout-warning title="Historical Lecture"}
This chapter has not yet been adapted to match the format and style of the rest of the book. It is included in its original lecture structure for readers who need to use generalized linear models.
:::

## Generalized Linear Models (GLMs)

- **This is an entire area of regression, and we could spend a full semester on it. Today's goal is a crash course on the basics of the most common GLM: logistic regression.**
- So far, you have been using a special case of the GLM in which we assume that the underlying distribution is Gaussian.
- Now we will expand our approach to examine the larger family.
- When you run a GLM, you need to state the family and link function you will use.

### Family and Link

- Linear regression assumes that the DV is Gaussian, with mean $\mu$, standard deviation $\sigma$, and a possible response range from $-\infty$ to $\infty$.
- GLMs allow us to examine categorical and count responses that cannot satisfy the same requirements as linear regression.
- Other distributions, such as the binomial and Poisson, require a different relationship between the mean and variance.
- We use a link function to connect the mean of the DV to the linear predictor.
- Here are common families used in psychology.

Family    | Variance  | Link
----------| -------- | -------
Gaussian  | Gaussian | identity
Binomial  | Binomial | logit/probit
Poisson   | Poisson  | log


## Linear Regression on a Binomial DV

- Let's simulate what happens when we analyze a binomial DV as if it were a plain old regression.
- My grandmother had a very hard time predicting gender in the 90s because clothing became big and baggy, men wore earrings and also started sporting long hair
- Let's simulate her prediction process: gender (DV; 0 = male, 1 = female) based on the hair length (IV; -3 to 3) and bagginess of clothing (IV; -3 to 3) of the person she was looking at.
- The goal is to predict gender, not the hair length of each gender, so *t*-tests are out. We can also add other predictors later.

```{r}
library(ggplot2)

set.seed(42)
n=30
x = runif(n,-3,3) #  Hair length centered [-3 short, 3 = long]
j = runif(n,-3,3) #  Baggy clothing centered [-3 not baggy, 3 = really baggy]
z =  .8*x - .2*j         
pr = 1/(1+exp(-z))  # pass through an inv-logit function
y = rbinom(n,1,pr)  # response variable
 
 #now feed it to glm:
LogisticStudy1= data.frame(Gender=y,HairLength=x, baggy=j)

ggplot(LogisticStudy1, aes(x=HairLength, y=Gender)) + geom_point() + 
  stat_smooth(method="lm", formula=y~x, se=FALSE)+
  theme_classic()
ggplot(LogisticStudy1, aes(x=baggy, y=Gender)) + geom_point() + 
  stat_smooth(method="lm", formula=y~x, se=FALSE)+
  theme_classic()

```

\pagebreak

```{r}
LM.1 <- lm(Gender ~ HairLength + baggy, data = LogisticStudy1)
summary(LM.1)

```

- The intercept reflects the predicted value of gender at the mean hair length and mean clothing bagginess.
- The slope for hair says that as hair gets longer, the model predicts a greater probability of female.
- The slope for bagginess says that as clothes get baggier, the model predicts a greater probability of male, but the slope is not significant.
- How do the residuals look?

```{r}
plot(LM.1, which =1)
```

- That looks odd because the fitted values are continuous, but the response is binomial.
- Let's see how well we predicted the result for each individual.


```{r}
LogisticStudy1$Predicted.Value<-predict(LM.1)
summary(LogisticStudy1$Predicted.Value)
```

- Yikes. The model predicted values outside the possible range.
- Let's say any prediction above .5 is female and anything at or below .5 is male.
- Then we will examine a contingency table comparing predicted and observed results.


```{r}
LogisticStudy1$Predicted.Gender<-ifelse(LogisticStudy1$Predicted.Value > .5,1,0)
C.Table<-with(LogisticStudy1,
   table(Predicted.Gender, Gender))
C.Table
```

- Correct predictions are 0-0 and 1-1; incorrect predictions are mismatches.

```{r}
PercentPredicted<-(C.Table[1,1]+C.Table[2,2])/sum(C.Table)*100
PercentPredicted
```

- The model did a good job making predictions in this simple case, but the ordinary $R^2$ is not appropriate for a binary outcome.

### Summary

- Using linear regression on these data produced predictions outside the bounded range, a violation of homoscedasticity, and an $R^2$ that is not appropriate for this outcome.
- Instead, let's try a GLM. First, we need to understand the binomial distribution and logit link function.

## Logistic Regression

- Logistic regression is the regression model we use here to analyze a binomial DV.

### Binomial Distribution

- The binomial probability mass function can be expressed as follows:

$$p(n\mid N) = \binom{N}{n}p^n(1-p)^{N-n}$$

- Here, $n$ is the number of successes in $N$ trials at a specified probability $p$.
- The distribution changes as a function of the underlying probability of obtaining a 0 or 1.
- The plot below shows $N=10$ trials, with probability changing from 0 to 1 in increments of .25.

```{r, fig.width=7.5, fig.height=3.0}

par(mfrow=c(1, 5))
for(p in seq(0, 1, len=5))
{
    x <- dbinom(0:10, size=10, p=p)
    barplot(x, names.arg=0:10, space=0)
}

```

- As the number of trials increases, the distribution looks more normal except near the boundaries. Here are 100 trials.

```{r, fig.width=7.5, fig.height=3.0}
par(mfrow=c(1, 5))
for(p in seq(0, 1, len=5))
{
    x <- dbinom(0:100, size=100, p=p)
    barplot(x, names.arg=0:100, space=0)
}
```

### Logit Link Function

- We can bound our results by making the best-fitting curve asymptotic to the response boundaries.
- To make this work, we need to switch from a straight line to a sigmoid curve.
- The linear predictor can range from $-\infty$ to $\infty$, so we transform the probability with the logit link.

$$\operatorname{logit}(p) = \log\left(\frac{p}{1-p}\right)$$

- I will use $\log$ for the natural logarithm to be consistent with R. Some texts write the natural logarithm as $\ln$.

```{r, fig.width=3.5, fig.height=3.0}
logit.Transform<-function(p) {log(p/(1-p)) }

plot(logit.Transform(seq(0,1,.0001)),seq(0,1,.0001),
     main="Logit Transform",ylim = c(0,1),
  xlab="Logit",ylab="Probability")

```


### Fit the Logistic Regression

- $\operatorname{logit}(Gender) = B_1(HairLength) + B_2(baggy) + B_0$
- We can accomplish this by changing the function in R from `lm()` to `glm()` and specifying the family as `binomial(link = "logit")`.
- First, let's fit the model and then make sense of the parameters.

```{r}
LR.1 <- glm(
  Gender ~ HairLength + baggy,
  data = LogisticStudy1,
  family = binomial(link = "logit")
)
summary(LR.1)

```


```{r, fig.width=3.5, fig.height=3.0}
library(ggplot2)
ggplot(LogisticStudy1, aes(x=HairLength, y=Gender)) + geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE)+
  theme_classic()

ggplot(LogisticStudy1, aes(x=baggy, y=Gender)) + geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE)+
  theme_classic()

```

### Plot in Probabilities

- The y-axis shows the predicted probability of gender as a function of hair length.

```{r, fig.width=5.5, fig.height=3.0}
library(effects)
PredictedData<-allEffects(LR.1)
plot(PredictedData)
```

- How did we get the probability from the logit?

### Interpret Coefficients

- Raw coefficients are in log-odds, which are difficult to interpret directly.
- We can transform them into odds or predicted probabilities.
- The odds of success are the ratio of the probability of success to the probability of failure.
- A 50% probability corresponds to odds of 1 to 1.
 $$Odds = e^{logit(x)}$$ 

```{r, fig.width=3.5, fig.height=3.0}
L.to.O.Transform<-function(p) {exp(logit.Transform(p))}

plot((logit.Transform(seq(0,.99,.01))),L.to.O.Transform(seq(0,.99,.01)),
     main="Logit to Odds Transform",
  xlab="Logit",ylab="Odds")

```

- Here are our regression estimates expressed as odds ratios.

```{r, fig.width=3.5, fig.height=3.0}
LR.1.Odds.Ratios <- exp(coef(LR.1))
LR.1.Odds.Ratios
```


- Physicians often discuss odds, but psychologists usually prefer probabilities.
$$P = \frac{Odds} {1 + Odds}$$

```{r, fig.width=3.5, fig.height=3.0}
O.to.P.Transform <- function(odds) odds / (1 + odds)
odds <- seq(0, 20, .1)

plot(odds, O.to.P.Transform(odds),
     main="Odds to Probability Transform",
  xlab="Odds",ylab="Probability")

```

- We can convert the complete regression equation to obtain predicted probabilities directly.

$$p_{(Gender)} =\frac{e^{(B_1(HairLength)+B_2(baggy) + B_0})}{1+e^{(B_1(HairLength)+B_2(baggy) + B_0})}$$

- Or, more simply:

$$p_{(Gender)} =\frac{1}{1+e^{-(B_1(HairLength)+B_2(baggy) + B_0})}$$



```{r, fig.width=3.5, fig.height=3.0}
prediction_grid <- data.frame(
  HairLength = c(-2, 0, 2),
  baggy = 0
)
prediction_grid$Predicted.Probability <- predict(
  LR.1,
  newdata = prediction_grid,
  type = "response"
)
prediction_grid
```

- Report the coefficients from the logistic regression, and translate important results into odds ratios or predicted probabilities.
- A coefficient cannot be converted into a probability by itself because the probability depends on the complete linear predictor, including the intercept and values of the other predictors.
- As in linear regression, add the relevant predictor terms first and then apply the inverse-logit transformation shown above.

### Wait: What About My R-Squared?

- Ordinary least-squares $R^2$ does not carry over unchanged to these models.
- A binary outcome's conditional variance depends on its mean, so homoscedasticity does not apply in the same way.
- Researchers have developed several pseudo-$R^2$ measures to describe fit or improvement over a null model.
- Common versions include Cox and Snell, Nagelkerke, McFadden, and Efron.
- An accessible description of these measures is available from [UCLA Statistical Consulting](https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faq-what-are-pseudo-r-squareds/).
- Because many of these measures compare a fitted model with a restricted or null model, we first need to examine model fitting for GLMs.

#### Deviance Testing

- Ordinary least squares provides the fitting criterion for linear regression. GLMs commonly use maximum likelihood instead.
- In OLS regression, we used this decomposition: $$SS_{Resid} = SS_Y - SS_{Regression}$$
- In other words, residual = observed value - fitted value.
- For most GLMs, the fitting algorithm iterates until it finds the parameter values that maximize the likelihood.
- The null model contains only an intercept. A more complex model contains the intercept plus predictors.

$$\text{Likelihood ratio} = \frac{L_{Simple}}{L_{Complex}}$$

- The likelihood-ratio statistic is:

$$G^2 = -2\log\left(\frac{L_{Simple}}{L_{Complex}}\right) = 2(LL_{Complex}-LL_{Simple})$$

First, I will show you pseudo-$R^2$. Then we will examine how to test differences in model fit.

- Null deviance:

$$D_{Null} = -2[\log(L_{Null}) - \log(L_{Saturated})]$$

- Model deviance:

$$D_{K} = -2[\log(L_{K}) - \log(L_{Saturated})]$$

- This plays a role similar to residual sums of squares in OLS.

#### Pseudo-R-Squared

- We need deviance scores for some pseudo-$R^2$ measures. For example:
$$R_L^2 = \frac{D_{Null} -D_{k}}{D_{Null}}$$
- SPSS reports Cox and Snell, so people use it often.
- Nagelkerke rescales Cox and Snell so that its maximum can reach 1.
- McFadden's version is also common.
$$R_{McFadden}^2 = 1 - \frac{LL_{k}}{LL_{Null}}$$
 
- We can obtain several measures from the `pscl` package.
- `llh` is the log-likelihood from the fitted model.
- `llhNull` is the log-likelihood from the intercept-only restricted model.
- $G^2 = 2(LL_{K} - LL_{NULL})$ is the likelihood-ratio statistic comparing the fitted and null models.
- McFadden = McFadden pseudo-$R^2$
- r2ML = Cox & Snell pseudo-$R^2$
- r2CU = Nagelkerke pseudo-$R^2$

```{r, echo=TRUE, warning=FALSE,message=FALSE}
library(pscl)
pR2(LR.1)
```
 
#### Wait, Why Do I See *z* and Not *t*-Values?

- Tests of individual predictors use Wald *z*-statistics rather than *t*-tests.
- The Wald *z*-statistic for a coefficient is:

$$z = \frac{B_j}{SE_{B_j}}$$

- Squaring this statistic gives a one-degree-of-freedom Wald chi-square statistic:

$$\chi^2_{Wald} = \left(\frac{B_j}{SE_{B_j}}\right)^2$$

- You can also request sequential likelihood-ratio tests. The order of predictors matters for this table.

```{r, echo=TRUE, warning=FALSE,message=FALSE}
anova(LR.1, test="Chisq")
```
 
### Hierarchical Testing

- Sequential testing can be difficult if you have many predictors.
- Instead of testing a change in ordinary $R^2$, we compare the deviance of nested models.
- A likelihood-ratio test compares the models using a chi-square reference distribution.

```{r, echo=TRUE, warning=FALSE,message=FALSE}
LR.Model.1<-glm(Gender~baggy,data=LogisticStudy1, family=binomial(link = "logit"))
LR.Model.2<-glm(Gender~baggy+HairLength,data=LogisticStudy1, family=binomial(link = "logit"))
anova(LR.Model.1,LR.Model.2,test = "Chisq")
```

- Here, Model 2 has lower residual deviance and fits better, so hair length improved prediction.

### How Well Am I Predicting?
- Model 1

```{r, echo=TRUE, warning=FALSE,message=FALSE}
fitted.results <- predict(LR.Model.1,newdata=LogisticStudy1,type='response')
fitted.results <- ifelse(fitted.results > 0.5,1,0)
misClasificError <- mean(fitted.results != LogisticStudy1$Gender)
print(paste('Accuracy = ',round(1-misClasificError,3)))
```
- Model 2

```{r, echo=TRUE, warning=FALSE,message=FALSE}
fitted.results <- predict(LR.Model.2,newdata=LogisticStudy1,type='response')
fitted.results <- ifelse(fitted.results > 0.5,1,0)
misClasificError <- mean(fitted.results != LogisticStudy1$Gender)
print(paste('Accuracy = ',round(1-misClasificError,3)))
```

- This is not the only way to examine accuracy.
- We have correct responses, misses, and false alarms. Remember our Type I and Type II error boxes.
- To visualize this relationship, we will examine a receiver operating characteristic (ROC) curve.
- We will calculate the area under the curve as a measure of discrimination. Values closer to 1 indicate better discrimination, whereas .5 indicates chance-level discrimination.

```{r, echo=TRUE, warning=FALSE,message=FALSE}
library(ROCR)
fitted.results <- predict(LR.Model.2,newdata=LogisticStudy1,type='response')
pr <- prediction(fitted.results, LogisticStudy1$Gender)

prf <- performance(pr, measure = "tpr", x.measure = "fpr")
plot(prf)

auc <- performance(pr, measure = "auc")
auc <- auc@y.values[[1]]
print(paste('Area under the Curve = ',round(auc,3)))

```


### Some New Things to Keep in Mind

- Should I use a logistic model? Is a sigmoid the right shape for the relationship?
- The Hosmer-Lemeshow test compares observed event proportions with predicted probabilities across subgroups using a chi-square statistic.
- Historically, a nonsignificant result was treated as evidence of adequate fit. Do not rely on this test alone; also inspect calibration, discrimination, and model specification.

```{r, echo=TRUE, warning=FALSE,message=FALSE}
library(MKmisc)
HLgof.test(fit = fitted(LR.Model.2), obs = LogisticStudy1$Gender)
```

