---
title: "Multiple Regression: Statistical Control"
geometry: margin=.5in
fontsize: 11pt
format:
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.width = 4.25)
knitr::opts_chunk$set(fig.height = 4.0)
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(fig.show = "hold")
```

```{r load libraries}
library(MASS)
library(tidyverse)
library(ggplot2)
library(GGally)
library(sjPlot)
```


## Introduction: Why One Predictor Is Never Enough

People are complicated. When a student does well on an exam, is it because they studied? Because they weren't anxious? Because they found the material interesting? It's all of the above — and more.

So far we've used **simple linear regression** (1 DV, 1 IV). But real prediction involves multiple factors at once. **Multiple regression** lets us predict an outcome from two or more predictors simultaneously.

The key concept: **statistical control** — estimating the association between X and Y while *holding other measured variables constant*. This lets us ask not just "does studying predict performance?" but "does studying predict performance **above and beyond** what anxiety alone explains?"

Without statistical control, you can't untangle those two things. With multiple regression, you can.

## A Study on Exam Performance

Let's ground this in a specific example. Imagine you are a researcher interested in what predicts student exam performance. After reading the literature, you identify two strong candidates:

1. **Study Hours** (X1) — how many hours per week a student spends studying for the exam. More studying should lead to better scores.
2. **Test Anxiety** (X2) — how anxious a student feels about the exam (measured on a 0–100 scale, where 0 = no anxiety and 100 = extreme anxiety). More anxiety is expected to hurt performance.

You recruit 100 college students and collect these measures before a major exam. After the exam, you record their score (0–100).

### The Complication: Our Two Predictors Are Correlated

Here is the twist. It turns out that Study Hours and Test Anxiety are **not independent of each other**. Students who study more tend to feel *less* anxious — perhaps because preparation builds confidence, or because anxious students have trouble sitting down to study in the first place.

This means:

- Students with high Study Hours tend to have low Test Anxiety.
- Students with low Study Hours tend to have high Test Anxiety.

In other words, Study Hours and Test Anxiety are **negatively correlated** with each other.

This creates a problem. If you only look at Study Hours → Exam Score, some of the apparent benefit of studying might actually be coming from lower anxiety — not from the studying itself. And if you only look at Anxiety → Exam Score, some of the apparent harm of anxiety might be coming from studying *less*, not from the anxiety per se.

Multiple regression lets us disentangle these overlapping influences.

### Expected Correlations

Here are the correlations we expect to find in our data:

| | Study Hours | Test Anxiety | Exam Score |
|---|---|---|---|
| **Study Hours**  | —     | −.35  | +.45 |
| **Test Anxiety** | −.35  | —     | −.40 |
| **Exam Score**   | +.45  | −.40  | —    |

- Study Hours *positively* predicts exam scores ($r = .45$): more studying, better scores.
- Test Anxiety *negatively* predicts exam scores ($r = -.40$): more anxiety, lower scores.
- Study Hours and Test Anxiety are *negatively correlated* ($r = -.35$): students who study more tend to be less anxious.

That third correlation is the key complication — and the reason we need multiple regression.

### Simulating the Data

```{r simulate data}
set.seed(42)

n   <- 100
rY1 <-  0.45   # Study Hours -> Exam Score
rY2 <- -0.40   # Test Anxiety -> Exam Score
r12 <- -0.35   # Study Hours <-> Test Anxiety

Sigma <- matrix(c(
   1,   rY1, rY2,
  rY1,   1,  r12,
  rY2,  r12,  1
), ncol = 3)

raw <- as.data.frame(mvrnorm(n, mu = c(0, 0, 0), Sigma = Sigma, empirical = TRUE))
colnames(raw) <- c("Exam", "StudyHours", "Anxiety")

# Rescale to realistic units
dat <- data.frame(
  Exam       = round(raw$Exam       * 12 + 75, 1),  # Mean = 75, SD = 12
  StudyHours = round(raw$StudyHours *  3 +  8, 1),  # Mean = 8 hrs/week, SD = 3
  Anxiety    = round(raw$Anxiety    * 10 + 50, 1)   # Mean = 50 on 0-100 scale, SD = 10
)
```

Always look at your data before running any analysis. Let's check the first few rows and the summary statistics.

```{r first look, echo=TRUE}
head(dat)
summary(dat)
```

And here is a pairs plot showing all three pairwise relationships simultaneously. Each panel in the lower triangle shows a scatterplot, and the diagonal shows histograms for each variable.

```{r pairs plot, fig.width = 5.5, fig.height = 5, echo=TRUE}
GGally::ggpairs(
  dat,
  lower = list(continuous = "smooth"),
  diag  = list(continuous = "barDiag")
) + theme_bw()
```

Here is what that code asks R to do:

- `GGally::ggpairs()` builds a matrix containing every pair of variables. The `GGally::` prefix tells R exactly which package owns the function.
- `dat` is the data frame whose variables enter the matrix.
- `lower = list(continuous = "smooth")` places scatterplots and fitted smooth lines in the lower triangle.
- `diag = list(continuous = "barDiag")` places a histogram for each variable on the diagonal.
- `theme_bw()` removes some visual clutter so the relationships are easier to inspect.

The plot is compact because each variable takes a turn as both an x-axis and a y-axis. Read one panel at a time. Do not stare at the whole matrix until a causal theory appears.

You can see all three correlations in the scatterplots:

- **Bottom left:** Study Hours and Test Anxiety slope downward — students who study more tend to be less anxious.
- **Bottom middle:** Test Anxiety and Exam Score slope downward — more anxious students tend to score lower.
- **Bottom right:** Study Hours and Exam Score slope upward — more studying is associated with higher scores.

## Understanding Overlapping Variance

Before we run the multiple regression, we need to understand *why* having two correlated predictors creates the problem we described above. The key concept is **overlapping variance**.

### The Venn Diagram Approach

When you use simple regression with a single predictor, things are straightforward: X either predicts Y or it does not. We can visualize this as two circles — one for X and one for Y — where the *overlap* represents the shared variance between them. As you already know, that overlap, when squared, gives us $R^2$.

When we add a second predictor, things get more interesting. Now we have three circles — X1, X2, and Y — and some of the circles overlap with each other:

```{r ballantine-overlap}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
#| fig-cap: "A Ballantine for Study Hours, Test Anxiety, and Exam Score. The areas are conceptual, not drawn to scale."

plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1), asp = 1)

symbols(
  x = c(0.50, 0.34, 0.66),
  y = c(0.40, 0.63, 0.63),
  circles = c(0.30, 0.27, 0.27),
  inches = FALSE,
  add = TRUE,
  bg = c(
    adjustcolor("#009E73", alpha.f = 0.30),
    adjustcolor("#0072B2", alpha.f = 0.30),
    adjustcolor("#E69F00", alpha.f = 0.30)
  ),
  fg = c("#006B4F", "#004C7A", "#A66B00")
)

text(0.17, 0.90, "Study Hours\n(X1)", font = 2, col = "#004C7A")
text(0.83, 0.90, "Test Anxiety\n(X2)", font = 2, col = "#A66B00")
text(0.50, 0.035, "Exam Score (Y)", font = 2, col = "#006B4F")

text(0.34, 0.43, "a", cex = 1.5, font = 2)
text(0.66, 0.43, "b", cex = 1.5, font = 2)
text(0.50, 0.58, "c", cex = 1.5, font = 2)
text(0.50, 0.19, "e", cex = 1.5, font = 2)
```

In this diagram, we can label four distinct regions:

- **Region a** = variance in Y that is **uniquely explained by X1** (Study Hours), above and beyond X2.
- **Region b** = variance in Y that is **uniquely explained by X2** (Test Anxiety), above and beyond X1.
- **Region c** = variance in Y that **both X1 and X2 share** — neither predictor can claim exclusive credit for this part.
- **Region e** = variance in Y that is **unexplained** by either predictor (residual error).

The total variance in Y explained by the model is $a + b + c$.

### The Problem with Region c

Region $c$ is the heart of the matter. It represents variance in exam scores that both Study Hours and Anxiety are *jointly* associated with, because the two predictors are correlated with each other.

Here is the problem that region $c$ creates:

- If you only look at **Study Hours → Exam** (simple regression), the observed relationship includes region $a$ *and* region $c$. Part of what looks like a "Study Hours effect" is actually shared with Anxiety, because students who study more tend to be less anxious, and less anxious students score higher.
- If you only look at **Anxiety → Exam** (simple regression), the observed relationship includes region $b$ *and* region $c$. Part of what looks like an "Anxiety effect" is actually shared with Study Hours, for the same reason.

So the bivariate slopes mix unique and shared information. In this particular example, each slope is larger in absolute value when examined alone because it also has access to variance shared with the other predictor.

**Multiple regression solves this** by estimating each predictor's unique contribution. In the multiple regression model, Study Hours gets credit only for region $a$, and Anxiety gets credit only for region $b$. Region $c$ is acknowledged as shared variance and included in $R^2$, but it is not falsely assigned to either predictor.

In these simulated data, the multiple-regression slopes will be **smaller** than the corresponding simple-regression slopes. That is not a universal rule. Depending on the correlation pattern, adding a control can make a slope shrink, grow, or reverse. The controlled slopes answer a different question: what does each predictor contribute conditionally, given the other predictor in the model?

## The Multiple Regression Equation

You already know the simple regression equation:

$$Y = B_0 + B_1 X + e$$

**Multiple regression** extends this to two (or more) predictors:

$$Y = B_0 + B_1 X_1 + B_2 X_2 + e$$

Where:

- $B_0$ = intercept (predicted Y when all predictors = 0)
- $B_1$ = slope for X1 (Study Hours)
- $B_2$ = slope for X2 (Test Anxiety)
- $e$ = residual error (observed − predicted)

In our example:

$$\widehat{Exam} = B_0 + B_1 \cdot StudyHours + B_2 \cdot Anxiety$$

The equation looks very similar to simple regression. But the **interpretation of the slopes has changed in an important way**.

### What "Controlling For" Means

In simple regression, $B_1$ tells you: *"For each one-unit increase in X, how much does Y change?"*

In multiple regression, $B_1$ tells you something more precise: *"For each one-unit increase in X1, how much does Y change — **among people who have the same value of X2**?"*

This is what **statistical control** means. The slope for Study Hours in our multiple regression model describes the Study Hours–Exam association *as if we were comparing students at the same anxiety level.* We are estimating that association after holding anxiety **constant** — or equivalently, after **controlling for** anxiety.

Think of it this way: imagine sorting all 100 students into groups based on their anxiety score. Within the low-anxiety group, every student has roughly the same anxiety, so exam differences cannot be attributed to measured anxiety differences inside that group. They may be associated with study hours, or with other variables we did not measure. Now imagine doing the same thing for the medium-anxiety group and the high-anxiety group. Multiple regression performs a continuous version of this comparison for every value of Anxiety to isolate the unique predictive contribution of Study Hours.

**The key insight:** a slope in multiple regression is the association between that predictor and Y after accounting for the predictor's linear relationship with the other predictors in the model. It is the unique predictive contribution of X1 above and beyond what X2 already explains. It is not automatically a causal effect.

## Running the Analysis in R

Let's run the analysis step by step. We will start with simple regression for each predictor alone, then combine them. This approach lets you see exactly what changes — and why — when you add the second predictor.

### Step 1: Study Hours Alone

```{r model 1, echo=TRUE}
Model.1 <- lm(Exam ~ StudyHours, data = dat)
summary(Model.1)
```

So, looking at Study Hours alone:

- For every additional hour of studying per week, exam scores increase by about `r round(Model.1$coefficients[2], 2)` points.
- This is a statistically significant positive effect.
- Study Hours alone explains about `r round(summary(Model.1)$r.squared * 100, 1)`% of the variance in exam scores ($R^2 =$ `r round(summary(Model.1)$r.squared, 3)`).

### Step 2: Test Anxiety Alone

```{r model 2, echo=TRUE}
Model.2 <- lm(Exam ~ Anxiety, data = dat)
summary(Model.2)
```

Looking at Test Anxiety alone:

- For every 1-point increase in test anxiety, exam scores decrease by about `r abs(round(Model.2$coefficients[2], 2))` points.
- This is also statistically significant.
- Anxiety alone explains about `r round(summary(Model.2)$r.squared * 100, 1)`% of the variance ($R^2 =$ `r round(summary(Model.2)$r.squared, 3)`).

### Step 3: Both Predictors Together

Now for the key step — entering both predictors simultaneously. In R, we add a second predictor using the `+` sign in the formula:

```{r model 3, echo=TRUE}
Model.3 <- lm(Exam ~ StudyHours + Anxiety, data = dat)
summary(Model.3)
```

## Interpreting the Multiple Regression Output

Let's work through what changed when we added both predictors.

### The Slopes Have Shrunk — Why?

Compare the slope estimates across all three models:

| Predictor | Simple Regression (Alone) | Multiple Regression (Controlling for the Other) |
|---|---|---|
| Study Hours | `r round(Model.1$coefficients[2], 3)` | `r round(Model.3$coefficients[2], 3)` |
| Test Anxiety | `r round(Model.2$coefficients[2], 3)` | `r round(Model.3$coefficients[3], 3)` |

**Both slopes are smaller in absolute value** when the two predictors are in the model together. This is exactly what the Venn diagram predicts. Here is why:

When Study Hours was the *only* predictor (Model 1), its prediction included region $a$ *and* region $c$ — its unique association with exam performance plus variance shared with Anxiety. Once both predictors enter Model 3, the Study Hours coefficient is based on Study Hours variation that Anxiety does not linearly predict. Region $c$ still contributes to the model's overall $R^2$, but neither coefficient can claim it as uniquely its own.

The same logic applies in reverse: Anxiety's slope shrinks from `r round(Model.2$coefficients[2], 3)` to `r round(Model.3$coefficients[3], 3)` because some of its bivariate association with Exam Score overlaps with lower Study Hours.

This shrinkage is **not a problem**, nor is the model cold. Model 3 is answering the conditional question we asked. Its slopes estimate each predictor's unique association with Exam Score given the other measured predictor. They are not automatically "truer," more causal, or less biased; that depends on the research design, measurement quality, and whether controlling for that variable makes theoretical sense.

### Interpreting Each Controlled Slope

Here is how to interpret each slope in Model 3 in plain language:

**Study Hours** ($B_1 =$ `r round(Model.3$coefficients[2], 2)`):

> "For students with the *same* level of test anxiety, each additional hour of weekly study is associated with a `r round(Model.3$coefficients[2], 2)`-point increase in exam score."

**Test Anxiety** ($B_2 =$ `r round(Model.3$coefficients[3], 2)`):

> "For students who study the *same* number of hours per week, each one-point increase in test anxiety is associated with a `r abs(round(Model.3$coefficients[3], 2))`-point decrease in exam score."

The phrase **"for students with the same [other variable]"** is the key to interpreting any slope in multiple regression. When you see a slope reported in a multiple regression model, it always carries that "holding everything else constant" qualifier — even when it is not stated explicitly.

### The Intercept

The intercept ($B_0 =$ `r round(Model.3$coefficients[1], 2)`) is the predicted exam score when **both** Study Hours = 0 **and** Anxiety = 0. Since no student in our sample studied for zero hours *and* had zero anxiety simultaneously, this is not a prediction we would ever actually make. Like in simple regression, the intercept just anchors the equation mathematically. It becomes more meaningful when predictors are mean-centered (but we will save that for another time).

### Making a Prediction

Now that we have the full equation, we can make predictions. For a student who studies 10 hours per week and has an anxiety score of 45:


```{r prediction example, echo=TRUE}
# Predicted exam score for a student with 10 study hours and anxiety = 45
B0 <- Model.3$coefficients[1]
B1 <- Model.3$coefficients[2]
B2 <- Model.3$coefficients[3]

predicted <- B0 + B1 * 10 + B2 * 45
```

This student would be predicted to score about `r round(B0 + B1 * 10 + B2 * 45, 1)` on the exam.

## How Well Does the Model Fit? R²

Let's look at how the model's explanatory power ($R^2$) changes across our three models.

```{r r-squared comparison, echo=TRUE}
r2_model1 <- summary(Model.1)$r.squared
r2_model2 <- summary(Model.2)$r.squared
r2_model3 <- summary(Model.3)$r.squared
```

```{r,r-squared comparison2, echo=FALSE}

cat("Model 1 (Study Hours only): R-squared =", round(r2_model1, 3), "\n")
cat("Model 2 (Anxiety only):     R-squared =", round(r2_model2, 3), "\n")
cat("Model 3 (Both predictors):  R-squared =", round(r2_model3, 3), "\n")
```

The full model ($R^2 =$ `r round(r2_model3, 3)`) explains more variance than either predictor alone. But notice something: the combined model does **not** explain as much as the simple sum of the two individual $R^2$ values (`r round(r2_model1, 3)` + `r round(r2_model2, 3)` = `r round(r2_model1 + r2_model2, 3)`). Why not?

Go back to the Venn diagram. When Study Hours was the only predictor, it explained regions $a + c$. When Anxiety was the only predictor, it explained regions $b + c$. Together, they explain regions $a + b + c$. But notice that region $c$ was already counted in Model 1's $R^2$. Adding Anxiety on top of Study Hours can only add region $b$ to what was already explained — it cannot double-count region $c$.

This is why $R^2$ does not add up neatly when predictors are correlated. The combined model explains more than either predictor alone, but less than the naive sum of their individual $R^2$ values.

### Breaking Down the Venn Diagram Numerically

We can actually estimate the size of each region in the Venn diagram using simple subtraction:

```{r regions, echo=TRUE}
# Unique variance of Study Hours (region a)
region_a <- r2_model3 - r2_model2

# Unique variance of Anxiety (region b)
region_b <- r2_model3 - r2_model1

# Shared variance (region c)
region_c <- r2_model1 + r2_model2 - r2_model3
```

```{r regions2, echo=FALSE}
cat("Region a - unique to Study Hours:", round(region_a, 3), "\n")
cat("Region b - unique to Anxiety:    ", round(region_b, 3), "\n")
cat("Region c - shared variance:      ", round(region_c, 3), "\n")
cat("Total R-squared:                 ", round(r2_model3, 3), "\n")
```


In our data:

- Study Hours has about `r round(region_a * 100, 1)`% unique variance in exam scores — its unique predictive contribution after accounting for anxiety (region $a$).
- Anxiety has about `r round(region_b * 100, 1)`% unique variance — its unique predictive contribution after accounting for study hours (region $b$).
- About `r round(region_c * 100, 1)`% of the variance in exam scores is shared between the two predictors — variance that both contribute to together, but neither can claim exclusively (region $c$).

This numerical breakdown puts the Venn diagram concept in concrete terms. The **squared semi-partial correlations** correspond directly to regions $a$ and $b$. The regression slopes are based on the same unique predictor variation, but express the expected outcome change in Exam Score units.

### Adding Predictors Always Increases R²

There is an important quirk about $R^2$: it will **always increase** when you add another predictor to the model — even if that predictor is completely useless random noise. This is simply because any variable, no matter how irrelevant, will share at least a tiny sliver of variance with Y by chance in any given sample.

This means if you kept piling more and more predictors into a model, $R^2$ would keep creeping upward — not because those predictors were genuinely meaningful, but just because of sampling variability.

### Adjusted R²

**Adjusted $R^2$** corrects for this problem by applying a penalty for each additional predictor:

$$\text{Adjusted } R^2 = 1 - \left(1 - R^2\right) \frac{n - 1}{n - k - 1}$$

Where $n$ = sample size and $k$ = number of predictors in the model. With a small sample and many predictors, the penalty is larger. With a large sample and few predictors, the penalty is small.

Raw $R^2$ describes how well the model fits **this sample**. Adjusted $R^2$ asks a slightly less flattering question: after penalizing us for the number of predictors we used, how much variance does the model appear to explain? Because raw $R^2$ capitalizes on accidental relationships in the current sample, adjusted $R^2$ is usually closer to what we might expect in the population or a new sample. It is not a prophecy, but it is less easily impressed.

Adjusted $R^2$ can increase when a new predictor improves the model enough to overcome the penalty. It can decrease when the added predictor contributes too little. This makes it a fairer comparison tool when models contain different numbers of predictors.

```{r adjusted r2, echo=TRUE}
adj_r2_model1 <- summary(Model.1)$adj.r.squared
adj_r2_model3 <- summary(Model.3)$adj.r.squared
```

```{r adjusted r2 print, echo=FALSE}
cat("Model 1 Adjusted R-squared:", round(adj_r2_model1, 3), "\n")
cat("Model 3 Adjusted R-squared:", round(adj_r2_model3, 3), "\n")
```


Our adjusted $R^2$ increased from `r round(adj_r2_model1, 3)` to `r round(adj_r2_model3, 3)` when we added Anxiety. This confirms that Anxiety is genuinely useful — after the penalty for adding a predictor, the model still explains meaningfully more variance.

People commonly report **both** $R^2$ and adjusted $R^2$. Raw $R^2$ describes the variance explained in the observed sample. Adjusted $R^2$ provides the more conservative estimate after accounting for model size and is often the better guess about how much explanatory power will survive beyond this particular group of exploited undergraduates.

## Visualizing Multiple Regression

Tables of numbers are useful, but plots help build intuition. Let's visualize what is happening.

### The Study Hours Slope — Before and After Controlling

This first pair of plots shows the predicted relationship between Study Hours and Exam Score in the two models. The key question is: how does the slope change when we control for Anxiety?

```{r plot study before, fig.width=4.5, fig.height=3.5}
plot_model(
  Model.1,
  type       = "pred",
  terms      = "StudyHours",
  colors     = "blue",
  title      = "Study Hours and Exam\n(Without controlling for Anxiety)",
  axis.title = c("Study Hours per Week", "Predicted Exam Score")
) +
  coord_cartesian(ylim = c(55, 95)) +
  theme_bw()
```

```{r plot study after, fig.width=4.5, fig.height=3.5}
plot_model(
  Model.3,
  type       = "pred",
  terms      = "StudyHours",
  colors     = "blue",
  title      = "Study Hours and Exam\n(Controlling for Anxiety)",
  axis.title = c("Study Hours per Week", "Predicted Exam Score")
) +
  coord_cartesian(ylim = c(55, 95)) +
  theme_bw()
```

The slope becomes slightly shallower after controlling for Anxiety. This is statistical control in action: some of the original Study Hours effect was shared with (and partially driven by) lower anxiety levels in high-studying students. The controlled slope is more precise — it captures only what studying contributes, independent of anxiety.

### The Anxiety Slope — Before and After Controlling

```{r plot anxiety before, fig.width=4.5, fig.height=3.5}
plot_model(
  Model.2,
  type       = "pred",
  terms      = "Anxiety",
  colors     = "red",
  title      = "Anxiety and Exam\n(Without controlling for Study Hours)",
  axis.title = c("Test Anxiety (0-100)", "Predicted Exam Score")
) +
  coord_cartesian(ylim = c(55, 95)) +
  theme_bw()
```

```{r plot anxiety after, fig.width=4.5, fig.height=3.5}
plot_model(
  Model.3,
  type       = "pred",
  terms      = "Anxiety",
  colors     = "red",
  title      = "Anxiety and Exam\n(Controlling for Study Hours)",
  axis.title = c("Test Anxiety (0-100)", "Predicted Exam Score")
) +
  coord_cartesian(ylim = c(55, 95)) +
  theme_bw()
```

The same pattern appears in reverse: the negative slope for Anxiety becomes slightly smaller after we account for the fact that highly anxious students also tend to study less. The controlled slope describes Anxiety's conditional association with performance while Study Hours is held constant.

### The Full Model: Both Predictors Together

We can also visualize both predictors simultaneously. The plot below shows how predicted exam scores change with Study Hours, for students at three different anxiety levels (low, average, and high).

```{r plot full model, fig.width=5, fig.height=4}
plot_model(
  Model.3,
  type         = "pred",
  terms        = c("StudyHours", "Anxiety [40, 50, 60]"),
  title        = "Predicted Exam Score by Study Hours and Anxiety Level",
  legend.title = "Test Anxiety",
  axis.title   = c("Study Hours per Week", "Predicted Exam Score")
) +
  coord_cartesian(ylim = c(55, 95)) +
  theme_bw()
```

Each line represents students at a different anxiety level: low anxiety (40), average anxiety (50), and high anxiety (60). Notice:

- **More studying always helps** — all three lines slope upward, regardless of anxiety level.
- **Higher anxiety always hurts** — the lines are shifted downward for higher-anxiety students.
- **The lines are parallel** — the predicted difference associated with each additional study hour is the same at every anxiety level. This is what it means for the terms to be **additive** in our model: the model contains separate Study Hours and Anxiety slopes but no interaction between them.

This plot gives you a visual intuition for what "controlling for" means: the slope of each line within one anxiety group represents the Study Hours association holding Anxiety constant — exactly what the multiple regression coefficient estimates.

## A Note on What Did Not Change

In Model 3, **both predictors remain statistically significant** after controlling for each other. The fact that both slopes shrank does not mean either predictor is unimportant. Both retain unique predictive associations with Exam Score: Study Hours contributes above and beyond Anxiety (region $a$), and Anxiety contributes above and beyond Study Hours (region $b$).

The meaning: **a slope that shrinks when you add a control variable is not automatically a sign of a weak finding.** It means the controlled and uncontrolled models answer different questions. In this example, the controlled slope is based on predictor variance not linearly explained by the other predictor.

Adding a control variable can also make a slope grow, disappear, reverse, or change statistical significance. Those patterns can reflect confounding, suppression, collinearity, measurement problems, or a poorly chosen control. The next chapter takes apart what statistical control is actually doing.

## APA Write-Up

Here is how you would write up these results in APA format:


Prior to the main analysis, zero-order correlations were examined. Study hours positively correlated with exam performance ($r = .45$, $p < .001$), and test anxiety negatively correlated with exam performance ($r = -.40$, $p < .001$). The two predictors were also significantly correlated with each other ($r = -.35$, $p < .001$), indicating that students who studied more tended to report lower anxiety.

A multiple linear regression was conducted with exam score as the outcome and study hours and test anxiety as predictors. The overall model was statistically significant, $F(2, 97) =$ `r round(summary(Model.3)$fstatistic[1], 2)`, $p < .001$, $R^2 =$ `r round(summary(Model.3)$r.squared, 3)`, adjusted $R^2 =$ `r round(summary(Model.3)$adj.r.squared, 3)`.

After controlling for test anxiety, study hours significantly predicted exam performance, $b =$ `r round(Model.3$coefficients[2], 2)`, $SE =$ `r round(summary(Model.3)$coefficients[2, 2], 2)`, $t(97) =$ `r round(summary(Model.3)$coefficients[2, 3], 2)`, $p < .001$. After controlling for study hours, test anxiety significantly predicted exam performance, $b =$ `r round(Model.3$coefficients[3], 2)`, $SE =$ `r round(summary(Model.3)$coefficients[3, 2], 2)`, $t(97) =$ `r round(summary(Model.3)$coefficients[3, 3], 2)`, $p <$ `r ifelse(summary(Model.3)$coefficients[3, 4] < .001, ".001", round(summary(Model.3)$coefficients[3, 4], 3))`.

Together, these results indicate that Study Hours and Test Anxiety each uniquely predict Exam Score in this model: students who study more and experience less anxiety tend to perform better, and both associations remain after accounting for the correlation between the predictors.


A few things to notice in this write-up:

- We tend to report the zero-order correlations first, so readers understand the bivariate relationships before seeing the controlled effects.
- We use the phrase **"after controlling for..."** (or equivalently, "above and beyond" or "holding constant") to signal that a multiple regression slope is being reported.
- We report $F$, $R^2$, and adjusted $R^2$ for the overall model, then report $b$, $SE$, $t$, and $p$ for each individual predictor.
- We do not report Cohen's $d$ here. In regression, $b$ (the unstandardized slope) is the primary effect size. The overall model's $R^2$ tells us the total proportion of variance explained.

::: {.callout-tip title="Next: We Take Control Apart"}
The next chapter, [Statistical Control: What Are We Actually Removing?](Chapter_Control.qmd), uses these same students to show exactly how residuals, semi-partial correlation, partial correlation, standardized beta, and change in $R^2$ fit together. No new imaginary undergraduates will be recruited until the current ones have finished the exam.
:::
