---
title: "Advanced: Regression Diagnostics and Influential Weirdos"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

## Advanced Topic: Who Is This For?

This chapter is for graduate students and advanced undergraduates who already know how to fit and interpret a regression model. We now ask a more alarming question: **Should we trust the model we just fit?**

Regression will always give you coefficients if the mathematics permits it. It does not pause to ask whether one participant entered a reaction time of 94 years or whether the residuals resemble a startled squid.

::: {.callout-warning title="A Significant Model Can Still Be a Bad Model"}
A small p-value does not certify linearity, equal residual variance, independence, honest measurement, or sensible causal reasoning. It only answers a narrower question under the model's assumptions.
:::

## A Deeply Responsible Therapy Study

Suppose we predict symptom improvement from therapy attendance and baseline symptom severity. Most participants behave normally. Participant 60 attends nearly every session and reports a suspiciously enormous improvement, possibly because they misunderstood the scale or wanted the research assistant to stop calling.

```{r diagnostic-data}
set.seed(343)
n <- 60
TherapyData <- data.frame(
  Sessions = runif(n, 0, 12),
  Baseline = rnorm(n, 60, 10)
)

TherapyData$Improvement <-
  1.8 * TherapyData$Sessions +
  0.18 * TherapyData$Baseline +
  rnorm(n, 0, 5)

TherapyData$Sessions[60] <- 12
TherapyData$Baseline[60] <- 45
TherapyData$Improvement[60] <- 60

TherapyModel <- lm(Improvement ~ Sessions + Baseline, data = TherapyData)
summary(TherapyModel)
```

## Residuals: What the Model Failed to Explain

For person $i$,

$$
e_i = Y_i-\widehat{Y}_i.
$$

A residual is the observed outcome minus the model's prediction. Positive residuals mean the person scored above the prediction; negative residuals mean they scored below it.

The ordinary linear model assumes:

$$
Y_i=\beta_0+\beta_1X_{1i}+\cdots+\beta_kX_{ki}+\epsilon_i,
$$

where the errors have an expected value of zero, are independent, and have approximately constant variance. Normality matters primarily for conventional small-sample inference, not because every raw variable must form a perfect bell.

```{r residual-fitted, fig.width=7, fig.height=4.5}
plot(
  fitted(TherapyModel), rstandard(TherapyModel),
  pch = 19, col = "#0072B2",
  xlab = "Fitted improvement",
  ylab = "Standardized residual",
  main = "Residuals should not organize a protest"
)
abline(h = 0, lwd = 2)
abline(h = c(-2, 2), lty = 2, col = "#D55E00")
text(fitted(TherapyModel)[60], rstandard(TherapyModel)[60], "60", pos = 4)
```

Look for curves, funnels, clusters, and isolated points. A random cloud around zero is comforting. A visible pattern means the model has left information lying on the floor.

::: {.callout-note title="Raw Residuals Versus Standardized Residuals"}
Raw residuals remain in the outcome's units. Standardized or studentized residuals divide by an estimated residual standard deviation, making unusually large residuals easier to identify. Values beyond roughly $\pm2$ deserve inspection. That is an investigation threshold, not an automatic deletion command.
:::

## Linearity

Linear regression assumes the mean outcome is a linear combination of the predictors. If the true relationship bends, a straight line may average over the bend and lie everywhere with great consistency.

Possible responses include transformations, polynomial terms, splines, or a model with a more appropriate link function. The graph should guide the question; repeatedly adding powers until an asterisk appears is polynomial fishing.

## Homoscedasticity

**Homoscedasticity** means that residual spread is roughly constant across fitted values. A funnel shape indicates **heteroscedasticity**: the model predicts some portions of the outcome more precisely than others.

Heteroscedasticity does not automatically bias ordinary least-squares slopes, but it can make conventional standard errors and confidence intervals inaccurate. We can reconsider the outcome scale, use heteroscedasticity-consistent standard errors, or fit a model that explicitly represents changing variance.

## Normality of the Errors

```{r residual-qq, fig.width=6.5, fig.height=4.5}
qqnorm(rstandard(TherapyModel), pch = 19, col = "#0072B2",
       main = "Q-Q plot of standardized residuals")
qqline(rstandard(TherapyModel), col = "#D55E00", lwd = 2)
```

A Q–Q plot compares residual quantiles with those expected under a normal distribution. Mild deviations are ordinary. Severe tail departures may signal outliers, a wrong outcome distribution, or a model that has ignored important structure.

::: {.callout-warning title="Do Not Test Normality and Surrender"}
With a large sample, a formal normality test can reject harmless deviations. With a small sample, it may miss serious ones. Inspect the residuals, consider the design, and ask whether the inferential procedure is robust enough for the observed problem.
:::

## Leverage, Influence, and Cook's Distance

An observation has high **leverage** when its predictor values are unusual. It is **influential** when removing it meaningfully changes the fitted model. A person can have a large residual without much leverage, or high leverage while sitting obediently on the fitted line.

Cook's distance combines residual size and leverage. One common expression is:

$$
D_i=
\frac{e_i^2}{p\widehat{\sigma}^2}
\frac{h_{ii}}{(1-h_{ii})^2},
$$

where $p$ is the number of fitted coefficients and $h_{ii}$ is leverage.

```{r influence-plot, fig.width=7, fig.height=4.5}
CookD <- cooks.distance(TherapyModel)
plot(
  CookD, type = "h", lwd = 2, col = "#0072B2",
  xlab = "Participant", ylab = "Cook's distance",
  main = "Which participant is driving the statistical bus?"
)
abline(h = 4 / n, lty = 2, col = "#D55E00", lwd = 2)
text(60, CookD[60], "60", pos = 3)
```

Rules such as $D_i>4/N$ are screening devices. They tell us where to look, not whom to throw into the statistical volcano.

## Multicollinearity

When predictors are strongly related, the model struggles to separate their unique slopes. Coefficients may become unstable, standard errors inflate, and signs may change when another predictor enters.

For predictor $j$, the variance inflation factor is:

$$
VIF_j=\frac{1}{1-R_j^2},
$$

where $R_j^2$ comes from predicting $X_j$ with the other predictors. A large VIF means the other predictors nearly reconstruct that predictor.

Multicollinearity is not repaired by staring harder at the output. Possible solutions include improving the design, collecting more informative cases, combining redundant measures when theory supports it, or narrowing the question.

## Independence and Clustering

Ordinary regression assumes residuals are independent. Repeated observations from one person, students within classrooms, siblings within families, and responses to the same stimulus are not independent simply because the spreadsheet gave them separate rows.

If residuals share a source, use a design-aware method such as a mixed-effects model. We will do that in the graduate mixed-model section.

## A Diagnostic Workflow

1. Plot the raw data.
2. Fit the theoretically justified model.
3. Plot residuals against fitted values.
4. inspect Q–Q behavior and residual tails.
5. Examine leverage and influence.
6. Check predictor redundancy.
7. Ask whether observations are independent.
8. Investigate suspicious cases against the original records.
9. Report consequential problems and sensitivity analyses.

::: {.callout-important title="Deleting the Problem Is Not Diagnosing the Problem"}
Never remove an observation merely because it weakens the hypothesis. Correct impossible values, explain defensible exclusions, and compare conclusions with and without influential cases. If one person reverses the entire result, that instability is part of the result.
:::

## The Mathematics Under the Diagnostic Pictures

Ordinary least squares produces fitted values through the **hat matrix**:

$$
\widehat{\mathbf y}=\mathbf H\mathbf y,
\qquad
\mathbf H=\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'.
$$

The diagonal value $h_{ii}$ is observation $i$'s leverage. Leverage comes from an unusual predictor pattern, not an unusual outcome by itself. A point can have high leverage and still sit obediently on the regression line. It becomes influential when its leverage combines with a meaningful residual.

Studentized residuals scale each residual by an estimate of its own uncertainty. Cook's distance then asks how much the fitted model would change if observation $i$ vanished:

$$
D_i \propto
\frac{r_i^2}{p}
\frac{h_{ii}}{1-h_{ii}}.
$$

The exact cutoffs are screening rules, not automatic deletion warrants. Inspect the case, verify the data, refit with and without it, and explain what changed.

## When the Variance Changes

Heteroscedasticity does not usually bias the OLS slope when the mean model is correct, but it can damage the ordinary standard errors. Possible responses include modeling the variance, transforming the outcome when scientifically sensible, using a model family that matches the outcome, or reporting heteroscedasticity-consistent standard errors. The picture tells you a problem exists; theory tells you which response makes sense.

## The Short Story

Diagnostics ask whether the model's leftovers behave as expected and whether a few observations control the conclusion. Plot residuals, inspect influence, respect clustering, and do not allow a significant coefficient to distract you from a visibly terrible model.
