---
title: "R Help and Review: Tidyverse Without Tears"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

## Why Is This at the End of the Book?

This is not a second statistics course hiding behind a fake mustache. It is a compact reference for the R operations you will repeatedly forget five minutes after learning them. That is normal. Professional R users also look things up constantly; they have merely developed more dignified facial expressions while doing it.

The **tidyverse** is a collection of R packages designed around consistent ways of storing, transforming, summarizing, and plotting data. We will focus on two:

- `dplyr` for manipulating data.
- `ggplot2` for making graphs.

Install them once on a computer:

```{r}
#| label: install-packages
#| eval: false
install.packages(c("dplyr", "ggplot2"))
```

Load them again whenever you begin a new R session. This chunk really does run, and it is what makes every other chunk in the chapter work:

```{r}
#| label: load-packages
library(dplyr)
library(ggplot2)
```

Installing is moving the furniture into your apartment. Loading is taking the furniture out of the closet so you can sit on it. Do not reinstall the sofa every morning.

## Our Deeply Serious Caffeine Study

The old lecture expected a file called `caffeine.csv` to materialize from a working directory last seen in 2019. We will simulate the data instead, which makes the example reproducible and removes the need to perform archaeological research inside Alex's Dropbox.

In this completely fictional study, 160 students consume different amounts of caffeine. We record sleep, anxiety, hatred of group projects, and whether the student begins hearing colors. No students actually consumed these doses. The IRB has stopped returning our calls for unrelated reasons.

```{r}
#| label: simulate-caffeine-data
set.seed(343)

n_students <- 160

lab_students <- data.frame(
  student_id = seq_len(n_students),
  caffeine_mg = sample(c(0, 100, 200, 400), n_students, replace = TRUE),
  sleep_hours = round(pmin(10, pmax(2, rnorm(n_students, 6.2, 1.4))), 1),
  group_project_hatred = sample(1:10, n_students, replace = TRUE)
)

color_probability <- plogis(
  -4 + .010 * lab_students$caffeine_mg - .20 * (lab_students$sleep_hours - 6)
)

lab_students$hears_colors <- factor(
  ifelse(rbinom(n_students, 1, color_probability) == 1, "Yes", "No"),
  levels = c("No", "Yes")
)

lab_students$anxiety <- round(pmin(
  10,
  pmax(
    0,
    4 + .009 * lab_students$caffeine_mg -
      .30 * lab_students$sleep_hours +
      .12 * lab_students$group_project_hatred +
      rnorm(n_students, 0, 1.2)
  )
), 1)

head(lab_students)
```

Because `set.seed(343)` fixes the random-number sequence, everyone gets the same fictional students. The Probability Gods are still fickle, but for this example we temporarily confiscated their dice.

## Data Frames

A data frame is a rectangular collection of variables:

- Each **row** is one observational unit, here one student.
- Each **column** is one variable.
- Each **cell** is one student's value on one variable.

This structure is often called **tidy data** when each variable has its own column, each observation has its own row, and each value has its own cell.

Useful first inspections include:

```{r}
#| label: inspect-data
dim(lab_students)
names(lab_students)
str(lab_students)
summary(lab_students)
```

`str()` is especially useful. It reveals whether R thinks a variable is numeric, character, or a factor before you spend 45 minutes accusing the regression model of personal betrayal.

## The Native Pipe

Modern R uses `|>` to pass the result of one step into the next step:

```{r}
#| label: pipe-head
lab_students |>
  head(3)
```

Read it as **"and then."**

> Start with `lab_students`, **and then** keep highly caffeinated students, **and then** select the variables we want.

```{r}
#| label: pipe-chain
lab_students |>
  filter(caffeine_mg >= 200) |>
  select(student_id, caffeine_mg, anxiety, hears_colors) |>
  head()
```

Older R code often uses `%>%`, the pipe from `magrittr` and `dplyr`. It is still valid and appears throughout older lectures, published code, and the ancient inscriptions found beneath BSB. For everyday `dplyr` work the two mean the same thing, so when you find `%>%` on Stack Overflow you can read it as `|>` and carry on. This book uses `|>` everywhere, including the earlier chapters, which were swept so you would never have to wonder whether the difference mattered.

## The Five Verbs You Will Use Constantly

### `select()`: Keep Columns

```{r}
#| label: select-keep
small_data <- lab_students |>
  select(student_id, caffeine_mg, sleep_hours, anxiety)

head(small_data)
```

Remove columns with a minus sign:

```{r}
#| label: select-drop
lab_students |>
  select(-student_id) |>
  head()
```

### `filter()`: Keep Rows

```{r}
#| label: filter-yes
chromatic_students <- lab_students |>
  filter(hears_colors == "Yes")

nrow(chromatic_students)
```

Combine conditions with `&` for **and** and `|` for **or**:

```{r}
#| label: filter-and
lab_students |>
  filter(caffeine_mg >= 200 & sleep_hours < 6) |>
  head()
```

Inside `filter()`, a comma means the same thing as `&`: every condition must be true. So `filter(caffeine_mg >= 200, sleep_hours < 6)` returns exactly the rows above. You will see both, including in the solutions at the end of this chapter, and neither is more correct than the other.

Use `==` to test equality. One `=` assigns an argument; two `==` ask whether values are equal. This tiny distinction has destroyed entire afternoons.

### `mutate()`: Create or Change Columns

```{r}
#| label: mutate-columns
lab_students <- lab_students |>
  mutate(
    caffeine_per_hour_sleep = caffeine_mg / sleep_hours,
    extremely_caffeinated = caffeine_mg >= 400
  )

head(lab_students)
```

`mutate()` keeps the existing variables and adds or replaces columns. The expression is calculated once per row.

### `arrange()`: Sort Rows

```{r}
#| label: arrange-rows
lab_students |>
  arrange(desc(anxiety)) |>
  select(student_id, caffeine_mg, sleep_hours, anxiety) |>
  head(10)
```

These are the students most likely to email at 2:13 a.m. with the subject line `QUICK QUESTION!!!!!`.

### `summarise()`: Collapse Many Rows into Answers

```{r}
#| label: summarise-all
lab_students |>
  summarise(
    n = n(),
    mean_anxiety = mean(anxiety),
    sd_anxiety = sd(anxiety),
    median_sleep = median(sleep_hours),
    proportion_hearing_colors = mean(hears_colors == "Yes")
  )
```

Unlike `mutate()`, `summarise()` reduces many rows to a smaller summary. `n()` counts rows. A logical statement such as `hears_colors == "Yes"` becomes `TRUE` or `FALSE`; its mean is therefore the proportion that is `TRUE`.

## Group, Then Summarize

`group_by()` tells later verbs to operate separately within categories.

```{r}
#| label: group-summarise
caffeine_summary <- lab_students |>
  group_by(caffeine_mg) |>
  summarise(
    n = n(),
    mean_anxiety = mean(anxiety),
    sd_anxiety = sd(anxiety),
    mean_sleep = mean(sleep_hours),
    proportion_hearing_colors = mean(hears_colors == "Yes"),
    .groups = "drop"
  )

caffeine_summary
```

`.groups = "drop"` removes the grouping after the summary. This prevents the invisible ghost of a previous grouping decision from possessing later analyses.

For simple frequency tables, `count()` is faster:

```{r}
#| label: count-table
lab_students |>
  count(caffeine_mg, hears_colors)
```

For the same summaries across several columns, use `across()`:

```{r}
#| label: across-columns
lab_students |>
  group_by(caffeine_mg) |>
  summarise(
    across(
      c(sleep_hours, anxiety, group_project_hatred),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )
```

## Reading a Real Data File

Quarto renders the book from the project directory. Use a project-relative path instead of `setwd()`:

```{r}
#| label: read-csv-example
#| eval: false
caffeine_data <- read.csv("data/caffeine.csv")
```

This says the file lives in a folder named `data` inside the project. Hard-coding a path to one person's Dropbox guarantees that the code will fail on another computer, next semester, or immediately after Alex reorganizes his folders during an avoidable burst of optimism.

## `ggplot2`: Build a Graph in Layers

A `ggplot` usually has three core ingredients:

1. **Data**: the data frame.
2. **Aesthetics**: variables mapped to position, color, shape, or size with `aes()`.
3. **Geometries**: the marks placed on the graph, such as points, bars, or boxes.

```{r}
#| label: ggplot-basic
#| fig-alt: "A bare ggplot scatterplot of anxiety against caffeine, with the default grey panel and unedited axis labels caffeine_mg and anxiety. Because caffeine takes only four values, the points stack into four vertical stripes at 0, 100, 200 and 400, with nothing at 300. Each stripe climbs a little higher than the last."
ggplot(
  data = lab_students,
  mapping = aes(x = caffeine_mg, y = anxiety)
) +
  geom_point()
```

The `+` adds layers to a ggplot. It is not interchangeable with the pipe. The pipe moves data through operations; the plus sign adds components to a graph. R is very committed to making punctuation carry emotional weight.

### Scatterplot with a Model Line

```{r}
#| label: ggplot-scatter-model
#| fig-cap: "Anxiety across caffeine doses. Jitter separates students who otherwise occupy the same caffeine value."
#| fig-alt: "The same data with the four stripes fanned out by jitter, so individual students are visible instead of stacked. A black regression line with a narrow grey confidence band rises steadily from about 2.7 anxiety at zero caffeine to about 6.2 at 400 mg. Points are colored by whether the student hears colors: the orange Yes points are rare and cluster at the high-caffeine end, with only one at zero."
ggplot(lab_students, aes(x = caffeine_mg, y = anxiety, color = hears_colors)) +
  geom_jitter(width = 12, height = 0, alpha = .65, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = c("No" = "#1F4E79", "Yes" = "#D55E00")) +
  labs(
    x = "Caffeine (mg)",
    y = "Anxiety rating",
    color = "Hears colors",
    title = "At some point the espresso begins explaining the wallpaper"
  ) +
  theme_minimal(base_size = 12)
```

`geom_jitter()` moves overlapping points by a tiny random amount horizontally so we can see them. It does not change the caffeine condition stored in the data.

### Boxplots, Correctly Explained

```{r}
#| label: ggplot-boxplot
#| fig-cap: "Anxiety distributions at each caffeine dose."
#| fig-alt: "Four boxplots, one per caffeine dose, marching upward from left to right. Medians rise from about 2.5 at zero, to 4.0 at 100 mg, 4.2 at 200 mg, and 6.3 at 400 mg. The zero-caffeine box is the tallest and has long whiskers reaching from 0 to about 5.4; the higher doses have tighter boxes. Orange dots beyond the whiskers mark a handful of outlying students at 100 and 200 mg."
ggplot(lab_students, aes(x = factor(caffeine_mg), y = anxiety)) +
  geom_boxplot(fill = "#9ECAE1", color = "#1F4E79", outlier.color = "#D55E00") +
  labs(
    x = "Caffeine (mg)",
    y = "Anxiety rating",
    title = "Boxes: useful, compact, and frequently lied about"
  ) +
  theme_minimal(base_size = 12)
```

The line inside each box is the **median**. The bottom and top of the box are the first and third quartiles, so the box contains the middle **50%** of scores. Its height is the interquartile range (IQR). The whiskers ordinarily extend to the most extreme observed values within 1.5 IQRs of the box; points beyond them are plotted separately.

Those separate points are not automatically errors, criminals, or volunteers for deletion. They are observations that deserve inspection and an explanation.

### Plotting a Summary

```{r}
#| label: ggplot-summary-col
#| fig-cap: "Simulated proportion of students reporting chromatic audition by caffeine dose."
#| fig-alt: "Pink bars showing the proportion of students hearing colors at each caffeine dose, on a y axis fixed from 0 to 1. The bars are tiny at 0 mg (.03), grow to .07 at 200 mg, and jump to .31 at 400 mg. There is no bar at all at 100 mg, because in this simulated sample not one of the 44 students at that dose reported hearing colors, so the bar has height zero."
ggplot(caffeine_summary, aes(x = factor(caffeine_mg), y = proportion_hearing_colors)) +
  geom_col(fill = "#CC79A7", width = .70) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Caffeine (mg)",
    y = "Proportion hearing colors",
    title = "Eventually teal starts whispering"
  ) +
  theme_minimal(base_size = 12)
```

`geom_col()` uses the y-values already calculated in `caffeine_summary`. `geom_bar()` would count raw rows instead. This difference is small, annoying, and worth remembering.

## A Compact Workflow

Most routine work follows this sequence:

```{r}
#| label: compact-workflow
analysis_data <- lab_students |>
  filter(!is.na(anxiety)) |>
  mutate(caffeine_group = factor(caffeine_mg)) |>
  select(student_id, caffeine_group, sleep_hours, anxiety, hears_colors)

analysis_summary <- analysis_data |>
  group_by(caffeine_group) |>
  summarise(
    n = n(),
    mean = mean(anxiety),
    sd = sd(anxiety),
    .groups = "drop"
  )

analysis_summary
```

1. Start with the original data.
2. Filter only with a defensible rule.
3. Create or recode needed variables.
4. Select the variables used in the analysis.
5. Check and summarize the result.
6. Graph the data before asking a model to explain them.

Keep the original data unchanged and create a new analysis object. Overwriting the only copy of a variable is how `Final_Data_REAL_v7_USE_THIS_ONE.csv` enters the world.

## Practice: Make the Caffeine Data Confess

Try these before looking at the solutions:

1. Keep students who slept fewer than six hours and consumed at least 200 mg of caffeine.
2. For each caffeine dose, calculate sample size, median anxiety, IQR of anxiety, and the proportion hearing colors.
3. Create a scatterplot of sleep and anxiety, coloring points by whether the student hears colors.
4. Find the ten students with the highest caffeine-per-hour-of-sleep values.

::: {.callout-tip collapse="true"}
## Solutions, because suffering has limits

```{r}
#| label: practice-solutions
#| fig-alt: "Solution scatterplot of anxiety against hours of sleep, points colored by whether the student hears colors, with a separate regression line and wide grey confidence band per group. Both lines are nearly flat, drifting down only slightly across the sleep range; the Yes group's line sits about a point higher but its band is very wide because there are few such students. Sleep does much less work here than caffeine did."
sleep_deprived_and_caffeinated <- lab_students |>
  filter(sleep_hours < 6, caffeine_mg >= 200)

practice_summary <- lab_students |>
  group_by(caffeine_mg) |>
  summarise(
    n = n(),
    median_anxiety = median(anxiety),
    iqr_anxiety = IQR(anxiety),
    proportion_hearing_colors = mean(hears_colors == "Yes"),
    .groups = "drop"
  )

ggplot(lab_students, aes(x = sleep_hours, y = anxiety, color = hears_colors)) +
  geom_point(alpha = .70) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal()

most_concerning_students <- lab_students |>
  arrange(desc(caffeine_per_hour_sleep)) |>
  select(student_id, caffeine_mg, sleep_hours, caffeine_per_hour_sleep) |>
  head(10)

practice_summary
most_concerning_students
```
:::

## Final Advice

Write code as a sequence of small, named decisions. Inspect the result after each important transformation. Make graphs before interpreting models. When something fails, read the error message from the bottom up and identify the first object or argument R says it cannot find.

Most importantly, do not confuse code that runs with an analysis that makes sense. R will cheerfully calculate the mean of participant ID numbers, regress anxiety on alphabetical order, and draw a confidence interval around almost any mistake you can encode. The software is obedient. You still have to be the adult in the room, which seems unfair but is apparently how science works.
