---
title: "Pearson's Chi-Square"
---
```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(fig.width=4.25)
knitr::opts_chunk$set(fig.height=4.0)
knitr::opts_chunk$set(fig.align='center')
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(results='hold')
```

```{r}
#| label: stored-values
#| include: false
# Scaffolding, not material. Every number quoted in the prose below is pulled
# from these objects so the text can never drift away from the output. The
# reader builds all of them by hand, and then in code, later in the chapter.
MCQ.Study <- as.table(rbind(c(27, 195)))
dimnames(MCQ.Study) <- list(Freq = c("Observed"),
                            Action = c("RtoW", "WtoR"))
MCQ.chi <- chisq.test(MCQ.Study, p = c(.5, .5))

MCQ.Study.2 <- as.table(rbind(c(97, 411), c(251, 620)))
dimnames(MCQ.Study.2) <- list(Difficulty = c("Easy", "Diff"),
                              Change = c("RtoW", "WtoR"))
MCQ.chi.2 <- chisq.test(MCQ.Study.2, correct = TRUE)
MCQ.chi.2.plain <- chisq.test(MCQ.Study.2, correct = FALSE)
```

<!-- ---Alex--- New cold open, which you approved on 2026-08-17. The device is the
     indict-the-tool move from Ch10/Ch15: hand a t-test the Best counts and watch R
     produce a completely respectable-looking p = .41 out of a table it does not
     understand. The payoff is an identity, not a callback: t.test() reports a mean of
     111, and 111 is exactly the expected frequency f_e that the hand calculation uses
     three sections later. That is called out at the hand calculation. The old opening
     (the parametric/nonparametric taxonomy) now sits directly underneath, where the
     reader has a reason to want it. Delete this comment if you keep the prose. -->

Here is a real result, and it is two numbers.

@best1979 watched 261 introductory psychology students take a multiple-choice exam and counted what happened every time somebody changed an answer. 27 changes went from right to wrong. 195 went from wrong to right.

You have spent this entire book learning how to test things. So test it.

```{r}
#| label: cold-open-ttest
#| echo: true
t.test(c(27, 195))
```

R did not stop me. It did not even hesitate. It reported a mean of 111, a *t*-value, a *p*-value of .41, and a confidence interval running from about $-956$ to about $1178$. Read that output at face value and you would conclude there is nothing here, then go tell every student in the country to stay with their first answer.

That would be one of the more wrong things in this book. Changing an answer helped roughly seven times as often as it hurt.

The arithmetic is not the problem. The arithmetic is flawless. The problem is that 111 is not anything. No student made 111 answer changes. There is no person, no score, and no variable that 111 is the average of. I handed a *t*-test a two-cell frequency table, and a *t*-test does not know what a frequency table is. It knows how to average numbers, so it averaged them.

*R will average anything you hand it. That is not the same as the average meaning something.*

That is why this chapter exists. Every test you have met so far takes a column of measurements and asks how far apart the means are. This study never measured anybody on anything: it counted events and sorted them into bins. Counts need a test of their own, and that test asks a different question.

> Do the observed counts differ from the counts we would expect by chance?

Categorical outcomes are everywhere in psychology. Did the participant comply or not, did the patient relapse or not, which of four options did they choose, did the answer change help or hurt. Whenever your outcome is a category instead of a quantity, this is the chapter you need. It is also the chapter where you get to watch a chance distribution get built from scratch, ten thousand fake students at a time, instead of being handed one and told to trust it.

## Parametric vs. Non-Parametric

Statistical tests are often divided into parametric and nonparametric families. The difference concerns what the model assumes and what part of the data it compares.

### Parametric Tests (everything up to this point)

Parametric tests describe populations with parameters such as means and variances. Particular tests make particular distributional and error assumptions; they do not all demand that the raw scores themselves be perfectly normal. They are powerful when their assumptions are reasonable.

You already know the ingredients, and by now you know the tests: the mean and SD from Part 1 are what every *t*-test, *F*-test, and regression you have run since then is built out of. That is the family we just watched fail on two counts.

### Nonparametric Tests (this chapter and the next)

Nonparametric tests make fewer or different distributional assumptions and often ask questions about counts, signs, or ranks:

- They usually do not require normally distributed raw scores
- Work with ordinal or nominal data
- Also useful when parametric assumptions are violated
- Instead of means, they compare counts/frequencies or rank observations

They are not assumption-free. Independence, sampling, measurement, and the design still matter. The next chapter examines that entire nonparametric family in more detail.

### Chi-Square

Pearson's chi-square is often placed under the “nonparametric” umbrella because it does not model means with a normal error distribution. More precisely, it is a test for **categorical counts**. It is not the emergency version of a *t*-test to use whenever normality looks annoyed.

Rather than asking: “How far is this group mean from another group mean?”

Chi-square asks: “Do the observed counts differ from what we would expect by chance?”

In technical terms, chi-square hypothesis testing asks whether an observed pattern is consistent with a chance-based null hypothesis.

- We compare observed data to expected values under null hypothesis ($H_0$), which means we expect no difference from chance.
- Large discrepancies lead us to reject $H_0$ and provide support for the alternative hypothesis ($H_1$). 

*Note:* You already know this logic from the *t*-test chapters. Chi-square runs the same play with counts instead of means.

Chi-square is especially useful when:

- Outcomes are categories (yes/no, group A/B/C, correct/incorrect)
- Means and standard deviations either don’t exist or don’t make sense

There are two types of chi-square:

- One-Way Pearson's Chi-Square [Goodness of Fit]
- Two-Way Pearson's Chi-Square [Test of Independence]

## One-Way Pearson's Chi-Square [Goodness of Fit]

Back to the two numbers from the start of the chapter, this time with the right tool. You have been told: "Stay with your first answer on a multiple-choice test." *So is changing answers more likely to be harmful?*

> Best studied the responses of 261 students in an introductory psychology course [@best1979]. He recorded the number of right-to-wrong (27), wrong-to-right (195), and wrong-to-wrong (39) answer changes for each student. Note: We will ignore wrong-to-wrong as he did in his first analysis.

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  27            | 195

Total of 222 observations

We will use the **One-Way Chi-Squared** [Goodness of fit]. It asks whether the relative frequencies observed in the categories of a sample frequency distribution are in agreement with the relative frequencies hypothesized to be true in the population.

### Hypothesis for Goodness of fit

There are two versions (but it does not change the math, just how you talk about it): 

- No preference: there will be no preference between different categories (e.g., do people prefer Coke or Pepsi)
-	No difference from a known population: This type hypothesis compares two populations (or representative samples)

If people were making chance guesses from a population of people who don't have any knowledge about material on the test (version 2 of the hypothesis):

- $H_0: P_{right-to-wrong} = .5, P_{wrong-to-right} = .5$ 
- $H_1: H_0$ is False 

*Note:* The probability you set can be whatever you want, but across all cells, they must add to 1. When you do not know what probability should be from a theoretical standpoint, you can simply do 1/Number of cells. 

### Find Expected Frequencies

If we expect by chance people would have gotten p = .50 on wrong-to-right and right to wrong, $f_e = pn$  

$F_{e_{right-to-wrong}}= .5*222=111$
$F_{e_{wrong-to-right}}= .5*222=111$

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  27            | 195
Expected  |  111           | 111

*Note:* These tables are often shown in papers (sometimes as proportions).

### Goodness of Fit Calculation by Hand

$$\chi^2 = \sum \frac{(f_o-f_e)^2}{f_e}$$

$\chi^2 = \frac{(27-111)^2}{111}+\frac{(195-111)^2}{111}$ = `r round(MCQ.chi$statistic, 2)`

*Note:* Never enter proportions into the chi-square. You must work with frequencies.

One thing to notice before we go on. The expected frequency here is 111, which is exactly the mean the *t*-test printed at the top of the chapter. The *t*-test found the right number. It just had no idea what the number was for: it treated 111 as the center of the data, when 111 is the thing the data are supposed to be compared *against*.

### Test Against Chance Distribution

$\chi^2$ values close to zero are probably “chance,” and values that are larger are farther from chance.
Wait. How do I know that?

Look at our $f_o$ values. If the $f_o$ values were close to the $f_e$ values, then the total value we calculate would be very close to zero.

For example, if we actually found this pattern instead:

Frequency | Right-to-Wrong | Wrong-to-Right
:---------|:--------------:|:-------------: 
Observed  |  111           | 111
Expected  |  111           | 111

$\chi^2 = \frac{(111-111)^2}{111}+\frac{(111-111)^2}{111} = 0$

While it’s very rare for the observed frequencies to be exactly what we expect by chance, it’s often really close to zero.

To get a sense of how often it would be close to zero, imagine we created a new test and the multiple-choice question said:

> “Please select an answer, then erase it and select something else. There is one right answer.”

That’s all you get. No actual knowledge is being tested. It’s a pure game of chance! In other words, people are totally guessing and are forced to change their answer.

Sometimes, purely by chance, they will go from Right-to-Wrong or Wrong-to-Right (and we will ignore all the Wrong-to-Wrong cases). What would that look like?

I can run 10,000 subjects to get a sense of what that might look like.

```{r}
#| label: chance-hist-full
#| echo: false
#| fig-alt: "Histogram of 10,000 chi-square values simulated under pure chance with one degree of freedom. The x axis runs from 0 to 20, the y axis counts how many of the 10,000 landed in each narrow bin. Almost the entire distribution is packed into a spike against the left edge at zero, and the bars shrink to a thin trickle that is barely visible past about 5. A few stray values reach as far as 17."
set.seed(42)
x <- rchisq(1e4, 1)
hist(x, prob=FALSE, main="Chance Histogram",
  xlab="Chi-square value", ylab="Number", xlim=c(0,20),breaks =5000)
#lines(density(x), col='red')
```

It looks like we get zero and values close to zero a lot, but also, purely by chance, we can get values all the way out to about 17.25. So even by chance, we can get a value that’s “far” from zero.

But does that happen often?

Before we answer that question, let’s zoom in closer to zero so we can see what’s going on more clearly.

```{r}
#| label: chance-hist-zoom
#| echo: false
#| fig-alt: "The same 10,000 simulated chi-square values, now with the x axis zoomed to run only from 0 to 0.25. What looked like a single spike is a smooth downhill slide: the tallest bar sits against zero at roughly 570 values, drops to about 250 by 0.01, and keeps easing down to roughly 30 per bin by 0.25."
hist(x, prob=FALSE, main="Chance Histogram",
  xlab="Chi-square value", ylab="Number", xlim=c(0,.25), breaks =5000)
#lines(density(x), col='red')
```

So how far from zero is still “chance”?

To answer that question, we would need more than 10,000 people. In fact, we would need the entire population. That would be impossible, but luckily, someone has done the math-magic to figure out what that would look like: 

```{r}
#| label: chisq-density-df1
#| echo: false
#| fig-alt: "The theoretical chi-square density curve with one degree of freedom, x axis 0 to 20 and y axis 0 to about 0.8. The curve starts high against the left edge, falls very steeply through the first two units, and has flattened onto the horizontal axis by about 8, where it stays for the rest of the plot."
curve(
  dchisq(x, df = 1),
  from = 0,
  to = 20,
  main = "Chi-Square Chance Distribution (df = 1)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)
```

Earlier, we looked at frequencies: how often different chi-square values occurred in a simulation.

Here, we switch to a probability density, which describes how likely different chi-square values are under the null hypothesis.

- The x-axis shows possible chi-square values.
- The y-axis shows relative likelihood (how often it will occur), not counts.

The important idea is that:

- Values near zero are very likely under chance
- Larger values are increasingly unlikely under chance

The total area under the curve equals 1, representing all possible outcomes under the null hypothesis.

### How do we set chance? 

How extreme does a chi-square value need to be before we stop believing it’s just chance?

One very common rule is the 95% / 5% rule.

>We say: “If the null hypothesis is true, then 95% of the time our chi-square values should fall below some cutoff.”

That cutoff defines the critical value. Any value beyond that point is rare enough that we start to doubt the chance explanation.

This does not mean:

- That chance is impossible past that value
- Or that we are 100% certain the result is “real”

It means:

- About 5% of the time, even when the null hypothesis is true, we will still see values this extreme by chance alone
- We accept that risk in exchange for having a decision rule

So when we say a result is “unlikely due to chance,” what we really mean is:

> “If this were just chance, we would expect to see something this extreme only about 5% of the time.”

<!-- ---Alex--- Both critical-region plots had a clipped label in the rendered PNG:
     at df = 1 "(critical region)" ran off the bottom edge, and at df = 5 the first
     line was cut to "5% of chance va" on the right. Invisible in the code, obvious in
     the image. The label is now parked in the empty upper right with an arrow pointing
     at the shaded wedge, and ylim is set explicitly so the position is deterministic.
     Nothing about the statistics changed. -->
```{r}
#| label: crit-region-df1
#| echo: false
#| fig-alt: "The chi-square density curve for one degree of freedom again, now with a dashed vertical line at 3.84 and the sliver of area to its right shaded red. The shaded region is a thin wedge hugging the axis, which is the visual point: 5 percent of the total area is squeezed into that long flat tail. An arrow labeled '5 percent of chance values, critical region' points down to it."
df <- 1
crit <- qchisq(0.95, df)

curve(
  dchisq(x, df),
  from = 0,
  to = 20,
  ylim = c(0, 0.85),
  main = "Chi-Square Chance Distribution (df = 1)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)

# Shade critical region (top 5%)
x_vals <- seq(crit, 20, length.out = 1000)
y_vals <- dchisq(x_vals, df)
polygon(
  c(x_vals, rev(x_vals)),
  c(y_vals, rep(0, length(y_vals))),
  col = rgb(1, 0, 0, 0.3),
  border = NA
)

# Add critical value line
abline(v = crit, lty = 2, lwd = 2)

# Label. It is parked in the empty upper right and points down at the wedge,
# because the wedge itself is too thin to write inside.
text(
  20,
  0.62,
  labels = "5% of chance values\n(critical region)",
  adj = 1,
  cex = 0.75
)
arrows(x0 = 16.5, y0 = 0.62 * 0.70, x1 = crit + 1.2, y1 = 0.85 * 0.045,
       length = 0.07, lwd = 1)

```

The distribution changes as a function of the degrees of freedom [$df$ is the number of cells - 1]. Degrees of freedom tell us how much independent information is available to estimate variability in the data. If we have a different number of conditions, we have a different distribution to compare it to.

```{r}
#| label: crit-value-1
Crit.MCQ=qchisq(.05, df=1, lower.tail = FALSE)
```

We get the crit = `r round(Crit.MCQ, 2)`. Any $\chi^2$ we calculate above that value means it has only a 5% probability of being a chance result.

### Our results? 

We ask where is our chi-square value relative to this chance distribution given our $df = 1$.  We calculated `r round(MCQ.chi$statistic, 2)`, and that is in the tail (the small part, well above chance value). We can use R or a table from a book to get the critical values for alpha (p) =.05. We don't need to make the graph every time as someone did this already using calculus and has given us all the values with a simple function in R called `qchisq`.  

This function answers the question:

- “What chi-square value cuts off a certain percentage of chance outcomes?”
  - p = .05
- We are defining the 5% region, the part of the distribution that happens only about 5% of the time under chance.

- df = 1
  - This tells R which chi-square distribution to use.
    - With two categories, the degrees of freedom is 1.

- lower.tail = FALSE
  - This is important:
    - FALSE means we want the upper tail of the distribution

```{r}
#| label: crit-value-2
Crit.MCQ<-qchisq(p = .05, df=1, lower.tail = FALSE)
```

We get the crit = `r round(Crit.MCQ, 2)` < `r round(MCQ.chi$statistic, 2)` (experimental value we got), so we reject the null hypothesis ($H_0$) and provide evidence for the alternative ($H_1$).

### Report in APA

$\chi^2(df, N) = X.XX$, $p = .XXX$

Since our p-value was below .05, we rejected the null. Report the exact p-value to two or three decimals, and only fall back on $p < .001$ when it is smaller than that. Ours is far smaller than that.

Students were significantly more likely to change an answer from wrong to right (`r round(195 / sum(MCQ.Study) * 100, 2)`%) than from right to wrong (`r round(27 / sum(MCQ.Study) * 100, 2)`%), $\chi^2(1, N = 222) =$ `r round(MCQ.chi$statistic, 2)`, $p < .001$.

### Impact of higher DF

Note: shape of chance distribution changes as we have more df (say 6 cells to compare)

```{r}
#| label: crit-region-df5
#| echo: false
#| fig-alt: "The chi-square density curve for five degrees of freedom, which is now a humped, right-skewed shape peaking near 3 rather than a slide off the left edge. A dashed vertical line marks the critical value at 11.07, much farther right than the 3.84 of the previous plot, and the tail area beyond it is shaded red. An arrow labeled '5 percent of chance values, critical region' points down at the shaded tail."
df <- 5
crit <- qchisq(0.95, df)

curve(
  dchisq(x, df),
  from = 0,
  to = 20,
  ylim = c(0, 0.175),
  main = "Chi-Square Chance Distribution (df = 5)",
  xlab = "Chi-square value",
  ylab = "Probability density"
)

# Shade critical region (top 5%)
x_vals <- seq(crit, 20, length.out = 1000)
y_vals <- dchisq(x_vals, df)
polygon(
  c(x_vals, rev(x_vals)),
  c(y_vals, rep(0, length(y_vals))),
  col = rgb(1, 0, 0, 0.3),
  border = NA
)

# Add critical value line
abline(v = crit, lty = 2, lwd = 2)

# Label, parked and pointing, same as the df = 1 plot
text(
  20,
  0.158,
  labels = "5% of chance values\n(critical region)",
  adj = 1,
  cex = 0.75
)
arrows(x0 = 16.5, y0 = 0.158 * 0.70, x1 = crit + 1.2, y1 = 0.175 * 0.045,
       length = 0.07, lwd = 1)

```

As you can see the distribution gets more spread out. So with larger DF our critical value increases

```{r}
#| label: crit-by-df
#| echo: true
#| fig-alt: "Scatterplot of the .05 critical chi-square value against degrees of freedom from 1 to 8. The eight points climb steadily from about 3.8 at one degree of freedom to about 15.5 at eight, in a gentle curve that bends slightly downward as it rises."
ChiCrit<-qchisq(.05, 1:8, lower.tail = FALSE)
plot(ChiCrit, main="Chi-Square Critical Values: alpha = .05",
  xlab="df", ylab="Critical Value", xlim=c(1,8), ylim=c(0,25))
```

### Goodness of Fit Calculation by Code

The way above is the hard way, and it shows you understand what the computer is doing. Doing one-way chi-squares in R is really easy!

### Step 1: Build your Table

```{r}
#| label: gof-table
#| echo: true
MCQ.Study <- as.table(rbind(c(27, 195)))
dimnames(MCQ.Study) <- list(Freq = c("Observed"),
                           Action = c("RtoW", "WtoR"))
MCQ.Study
```

### Step 2: Calculate Chi-Square

```{r}
#| label: gof-test
#| echo: true
MCQ.chi <- chisq.test(MCQ.Study, p=c(.5, .5))
MCQ.chi
```

R prints the same `r round(MCQ.chi$statistic, 2)` we ground out by hand, which is the point of doing it by hand once.

Again, in APA format, we would say,

Students were significantly more likely to change an answer from wrong to right (`r round(195 / sum(MCQ.Study) * 100, 2)`%) than from right to wrong (`r round(27 / sum(MCQ.Study) * 100, 2)`%), $\chi^2(1, N = 222) =$ `r round(MCQ.chi$statistic, 2)`, $p < .001$.



## Two-Way Pearson's Chi-Square [Test of Independence]

"Maybe we should only change our answers on easy items, but on hard items, we should trust ourselves."

> Best (1979) also examined the 1670 total number of responses across the 261 students and divided the items into easy/difficult in an introductory psychology course.  He recorded the number of right-to-wrong, wrong-to-right, and wrong-to-wrong (39) answer changes for each student. Again we will ignore wrong-to-wrong (17% of the responses) for simplicity.

Frequency | Right to Wrong | Wrong to Right 
---------|----------------|----------------
Easy     |  97            | 411
Difficult|  251           | 620

We need to do a Two-Way Chi-Square [Test of Independence]. It asks whether observed frequencies reflect the independence of two qualitative variables. Compares the actual observed frequencies of some phenomenon (in our sample) with the frequencies we would expect if there were no relationship at all between the two variables in the larger (sampled) population. We ask if the two variables are independent: knowledge of the value of one variable provides no information about the value of another variable.

### Hypothesis for Test of Independence

The null hypothesis is that for each, the value obtained for one variable is not related to or influenced by the second variable.  This idea can be expressed in two versions:

- Version 1: A single sample is measured on two separate variables. For example, knowing your personality will help predict your color preference (alternative). Knowing your personality will NOT help you predict your color preference (null) [Personality and color preference are measured per person]
    -	Null: The variables are not related. 
    -	Alternative: The variables are related. 

- Version 2: Two (or more) samples are measured and compared to see if there are differences. For example, the same proportions of extroverts (group 1) and introverts (group 2) prefer the same color (null).  If the proportions are different for the two groups, then you have the alternative hypothesis
    -	Null: The samples are independent 
    -	Alternative: The samples are not independent

Here we are working with Version 1 (one group taking two types of items)

### Null Hypothesis ($H_0$)

- Answer-changing behavior is independent of item difficulty.
- The proportion of right-to-wrong and wrong-to-right changes is the same for easy and difficult items

Knowing whether an item is easy or difficult provides no information about how answers are changed

### Alternative Hypothesis ($H_1$)

Answer-changing behavior depends on item difficulty.

- Students are more likely to benefit from changing answers on easy items

- On difficult items, answer changes are less likely to improve performance

Knowing whether an item is easy or difficult does provide information about how answers are changed

Simply put: 

- $H_0:$ The item difficulty is not related to how a person changes their answer 
- $H_1:$ The item difficulty is related to how a person changes their answer 

### Find Expected Frequencies

This is a little more complex than the goodness of fit test

### Find row/col sums

Frequency | Right to Wrong | Wrong to Right | Sum
---------|----------------|----------------|----
Easy     |  97            | 411            | **508**
Difficult|  251           | 620            | **871**
Sum      |  **348**       | **1031**       | **1379** 

### Step 2: Expected frequencies per cell

$$f_e = \frac{col\,total*row\,total}{overall\,total}$$

Frequency | Right to Wrong | Wrong to Right | Sum
---------|----------------|----------------|----
Easy     |  97 (128.1972) | 411 (379.8028) | **508**
Difficult|  251 (219.8028)| 620 (651.1972) | **871**
Sum      |  **348**       | **1031**       | **1379** 


### Test of Independence Calculation by Hand

$$\chi^2 = \sum \frac{(f_o-f_e)^2}{f_e}$$

$\chi^2 = \frac{(97-128.1972)^2}{128.1972} + \frac{(411-379.8028)^2}{379.8028} +\frac{(251-219.8028)^2}{219.8028}+\frac{(620-651.1972)^2}{651.1972}$

$\chi^2 =$ `r round(MCQ.chi.2.plain$statistic, 2)`

### Test Against Distribution

$df = (Row-1)(Col-1)$
$df = (2-1)(2-1) = 1$

Given our $df = 1$.  We calculated `r round(MCQ.chi.2.plain$statistic, 2)` we can use R or a table to get the critical values for alpha = .05

```{r}
#| label: crit-value-3
Crit.MCQ=qchisq(.05, df=1, lower.tail = FALSE)
```

We get the crit = `r round(Crit.MCQ, 2)` < `r round(MCQ.chi.2.plain$statistic, 2)`, so we reject the null.

### Report in APA

$\chi^2(df, N) = X.XX$, $p = .XXX$, $V = .XX$

In APA format:

Item difficulty was related to how students changed their answers, $\chi^2(1, N = 1379) =$ `r round(MCQ.chi.2.plain$statistic, 2)`, $p < .001$.

One flag before you meet the code. That hand calculation used no continuity correction, and R's default for a $2 \times 2$ table does. So the output a few sections down will print a slightly smaller number than this one. Nothing is broken; we sort it out when it happens.

[That is all we can say statistically.  Normally people present a table in percentages to unpack this result, **but** You have to decide if you want to do row or column percentages. People do not present both]. I think the row percent make more sense in this case we want to compare across item difficulty. 

% Row    | Right to Wrong | Wrong to Right | %
---------|----------------|----------------|----
Easy     |  19.1%         |  80.9%         | 100%
Difficult|  28.8%         |  71.2%         | 100%
Col Total|  25.2%         |  74.8%         | 100% 


### Test of Independence Calculation by Code

Again the code for R is very easy, so I will show you how it works: 

### Step 1: Build Your table

```{r}
#| label: independence-table
#| echo: true
MCQ.Study.2 <- as.table(rbind(c(97, 411), c(251, 620)))
dimnames(MCQ.Study.2) <- list(Difficulty = c("Easy","Diff"),
                              Change = c("RtoW", "WtoR"))

MCQ.Study.2
```

### Step 2: Calculate Chi-Square

In tests of independence, R calculates the expected frequencies for you. For a $2 \times 2$ table, `chisq.test()` applies **Yates's continuity correction** by default when `correct = TRUE`. The correction tries to improve a continuous chi-square approximation to discrete counts, but it can be overly conservative. It is not a ritual that must always be performed. Use it deliberately, report whether you used it, and consider Fisher's exact test when expected counts are sparse.

$$\chi^2_{Yates} = \sum \frac{(|f_o-f_e| - .5)^2}{f_e}$$

*Note*: You can view the observed and expected frequencies. For larger sparse tables, `chisq.test(..., simulate.p.value = TRUE)` can estimate a p-value by simulation.

```{r}
#| label: independence-test
#| eval: true
#| echo: true
MCQ.chi.2 <- chisq.test(MCQ.Study.2,correct = TRUE)
MCQ.chi.2
```

There is the mismatch we promised. R printed `r round(MCQ.chi.2$statistic, 2)`; we calculated `r round(MCQ.chi.2.plain$statistic, 2)` by hand. That gap is Yates, and it is the entire gap. Turn the correction off and R lands right back on the hand calculation:

```{r}
#| label: independence-test-plain
#| echo: true
chisq.test(MCQ.Study.2, correct = FALSE)
```

Neither number is the wrong one. They answer the same question with slightly different arithmetic, and both reject the null here. What matters is that you say which one you ran, because a reader cannot tell `r round(MCQ.chi.2$statistic, 2)` from `r round(MCQ.chi.2.plain$statistic, 2)` by staring at it.

### Row Proportion Code
```{r}
#| label: row-proportions
#| eval: true
#| echo: true
prop.table(MCQ.Study.2,1)
```

### Col Proportion Code
```{r}
#| label: col-proportions
#| eval: true
#| echo: true
prop.table(MCQ.Study.2,2)
```


### Effect Size: Because an Asterisk Is Not a Personality

The chi-square test tells us whether the observed table is difficult to explain under the null hypothesis. It does **not** tell us whether the association is large enough to matter. With enough observations, a microscopic association can earn an asterisk and begin acting important at faculty meetings.

For a $2 \times 2$ table, **phi** ($\phi$) is:

$$
\phi = \sqrt{\frac{\chi^2}{N}}
$$

For a table of any size, use **Cramer's V**:

$$
V = \sqrt{\frac{\chi^2}{N\min(r-1,c-1)}}
$$

Here, $r$ and $c$ are the numbers of rows and columns. Cramer's V ranges from 0, meaning no association, to 1, meaning perfect association. Do not automatically translate `.10`, `.30`, and `.50` into small, medium, and large without thinking about the design and research area. Those labels are rough landmarks, not sacred commandments delivered on stone tablets by Cohen.

```{r}
#| label: effect-size
#| echo: true
chi_uncorrected <- chisq.test(MCQ.Study.2, correct = FALSE)
n_total <- sum(MCQ.Study.2)

phi <- sqrt(as.numeric(chi_uncorrected$statistic) / n_total)
cramers_v <- sqrt(
  as.numeric(chi_uncorrected$statistic) /
    (n_total * min(dim(MCQ.Study.2) - 1))
)

c(phi = phi, cramers_v = cramers_v)
```

For these data, $V = .11$: there is an association, but it is not a gigantic one. The asterisk has not transformed into a flaming sword.

### Odds Ratios for a 2-by-2 Table

An **odds ratio** compares the odds of an outcome between two groups. In our table, the odds of changing an easy item from right to wrong rather than wrong to right are $97/411$. For difficult items, those odds are $251/620$.

$$
OR = \frac{97/411}{251/620} = 0.58
$$

An odds ratio of 1 would mean equal odds. Here, the odds of a right-to-wrong rather than wrong-to-right change were about 42% lower for easy items than for difficult items. If you reverse the comparison, the odds for difficult items are about $1/0.58 = 1.71$ times those for easy items. Odds ratios are direction-sensitive little creatures, so always say which group and outcome are in the numerator.

Base R gives us the odds ratio and its confidence interval with Fisher's exact test:

```{r}
#| label: fisher-or
#| echo: true
fisher_result <- fisher.test(MCQ.Study.2)
fisher_result$estimate
fisher_result$conf.int
```

The estimated odds ratio is 0.58, 95% CI [0.44, 0.77].

### Power for Chi-Square

Power asks how often a study would detect an association of a specified size if that association really exists. Small samples give fickle chance more room to cause trouble. A weak association plus twelve people and one undergraduate who clicked every response while watching TikTok is not a sturdy research program.

For chi-square tests, Cohen's $w$ is commonly used in power calculations. In a $2 \times 2$ table it is equivalent to phi and Cramer's V. The function below uses only base R. It finds the total sample size needed for 80% power when $w = .1062$, approximately the association observed here.

```{r}
#| label: chisq-power
#| echo: true
chisq_power <- function(n, w, df, alpha = .05) {
  critical_value <- qchisq(1 - alpha, df = df)
  1 - pchisq(critical_value, df = df, ncp = n * w^2)
}

required_n <- uniroot(
  function(n) chisq_power(n, w = .1062, df = 1) - .80,
  interval = c(2, 10000)
)$root

ceiling(required_n)
```

The answer is about 696 total observations. That number is not permission to choose an effect size after seeing the results. Plan power using the smallest effect worth detecting, prior evidence, or a defensible range of possible effects.

### Reporting the Result

A compact report can include the test, effect size, and a table of interpretable percentages:

> Item difficulty was associated with answer-change direction, $\chi^2(1, N = 1379) =$ `r round(MCQ.chi.2$statistic, 2)`, $p < .001$, $V = .11$. Compared with difficult items, easy items had lower odds of being changed from right to wrong rather than wrong to right, $OR = 0.58$, 95% CI [0.44, 0.77].

The chi-square value above uses Yates's correction, matching the test already run in this chapter. The effect-size calculation uses the ordinary Pearson chi-square. State your choices so readers do not have to conduct forensic accounting on your results section.

### Which Cells Are Doing the Work?

A significant chi-square says the table as a whole is hard to explain under the null. It does not say which part of the table caused the trouble. The fitted object already knows, and you do not have to compute anything: `chisq.test()` stores the standardized residuals.

```{r}
#| label: standardized-residuals
#| echo: true
MCQ.chi.2$stdres
```

Read those like z-scores, one per cell. Each is the gap between observed and expected divided by its own standard error, so anything past about $\pm 2$ is a cell pulling real weight. The sign tells you the direction: positive means more cases landed there than chance predicted, negative means fewer. Easy items came up short on right-to-wrong changes and long on wrong-to-right ones, which is the association the omnibus test detected, now with an address.

All four values here are the same size because a $2 \times 2$ table has one degree of freedom: the four cells are one fact told four times. Residuals earn their keep on bigger tables, where a significant omnibus test can be hiding behind one rogue cell out of twenty. That is also where you should start counting, because twenty cells inspected is twenty comparisons, and twenty comparisons is a family. See [Multiple Comparisons](Chapter_Multiple_Comparisons.qmd).

### Chi-Square Issues

- Chi-square works with **counts**, not percentages, means, or the emotional intensity with which you typed the numbers.
- The observations must be independent. One person should not quietly appear in several cells unless the design and model explicitly handle repeated observations.
- The approximation becomes unreliable when expected counts are very small. A common guideline is that no expected count should be below 1 and no more than 20% should be below 5, but this is a guideline rather than a commandment. For a small $2 \times 2$ table, use Fisher's exact test. For larger sparse tables, a simulated p-value may be useful.
- Chi-square cannot be negative because every discrepancy is squared. It can equal zero when every observed count equals its expected count, an event normally witnessed immediately before the Probability Gods demand another sacrifice.
- A cell contributes more when its observed and expected counts differ by a large amount **relative to the expected count**.
- A significant omnibus test says that some association exists. It does not identify every cell responsible for it. Standardized residuals are the first place to look, and follow-up comparisons then create a multiple-testing problem, which is what [Multiple Comparisons](Chapter_Multiple_Comparisons.qmd) was for, before the asterisks multiply and form a union.

::: {.callout-important title="Counts First; Interpretation Second"}
A chi-square test asks whether the observed **counts** differ from the counts expected under the null model. After that test, inspect the percentages, residuals, and effect size to learn where the association lives and whether anybody should care.
:::
