---
title: "Calculating the Two-Way Between-Subjects ANOVA"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 4.25,
  fig.height = 4
)
```


## Design Example
> Kardas and O'Brien (2018) reported that, with the proliferation of YouTube videos, people seem to think they can learn by *seeing* rather than *doing*. They "hypothesized that the more people merely watch others, the more they believe they can perform the skill themselves." They compared repeatedly watching the "tablecloth trick" with repeatedly reading or thinking about it. Participants assessed their own ability to perform the trick on a scale from 1 (I feel there is no chance at all I would succeed on this attempt) to 7 (I feel I would definitely succeed on this attempt). The study used a 3 (type of exposure: watch, read, think) x 2 (amount of exposure: low, high) between-subjects design. They collected $N = 1{,}003$ and found small effect sizes ($\eta^2 < .06$). To make this analysis manageable by hand, I will reduce the sample to five people per cell and inflate the effect size while preserving the pattern from Experiment 1.
  
 
**Low Exposure Group**  |Watching  | Reading  | Thinking | **Row Means**
------------------------|---------|----------|----------|----------
                 $\,$   |1        | 3        | 3        |
                 $\,$   |3        | 2        | 3        |
                 $\,$   |3        | 3        | 2        |
                 $\,$   |4        | 1        | 3        |
                 $\,$   |2        | 4        | 1        |
------------------------|---------|----------|----------|----------                 
            Cell Means  |*2.6*    | *2.6*    |*2.4*     | **2.53**
            Cell SS     |*5.2*    | *5.2*    | *3.2*    |


**High Exposure Group** |Watching | Reading  | Thinking | **Row Means**
------------------------|---------|----------|----------|----------
                 $\,$   |6        | 5        | 3        |
                 $\,$   |4        | 3        | 2        |
                 $\,$   |7        | 2        | 1        |
                 $\,$   |5        | 3        | 4        |
                 $\,$   |5        | 2        | 4        |
------------------------|---------|----------|----------|----------      
            Cell Means  |*5.4*    | *3.0*    |*2.8*     | **3.73** 
            Cell SS     |*5.2*    | *6.0*    | *6.8*    | 
**Column Means**        |**4.0**  |**2.8**   |**2.6**   | G = 3.13


## Logic of Two-Way ANOVA

We now have two factors, and we will test the main effects and interaction. 

- Main effect (Factor A: Rows): The independent effect of Length of Exposure on Perceived Ability
- Main effect (Factor B: Columns): The independent effect of Type of Exposure on Perceived Ability
- Interaction: The joint effect of Length and Type of Exposure on Perceived Ability


### Hypotheses

#### Main Effects

$H_0$: All groups have equal means.

$H_1$: At least one group differs from the others.

#### Interaction

$H_0$: There is no interaction between Factors A and B. All mean differences between treatment conditions are explained by the main effects.

$H_1$: There is an interaction between Factors A and B. The mean differences between treatment conditions are not what we would predict from the overall main effects of the two factors.


### Logic of Variance Parsing
A one-way ANOVA has only one factor to test:
$SS_T = SS_{W}+SS_{B}$

A two-way ANOVA has multiple effects to test:
$SS_T = SS_{W}+ (SS_{FactorA}+SS_{FactorB}+SS_{AxB})$

In a two-way ANOVA, $SS_B = SS_{FactorA}+SS_{FactorB}+SS_{AxB}$.


#### *F*-Ratio Logic
The omnibus *F* is the same as in a one-way ANOVA:

$$F= \frac{MS_{B}}{MS_{W}}$$

This still tells us whether at least one cell differs from another, but we must also calculate *F*-tests for Factors A and B and the A x B interaction.

$$F= \frac{MS_{FactorA}}{MS_{W}}$$
$$F= \frac{MS_{FactorB}}{MS_{W}}$$
$$F= \frac{MS_{AXB}}{MS_{W}}$$


## Analysis Procedures

### One-Way ANOVA Review (Simplified for Equal Sample Sizes)

With the formulas (formal equations)

Source | SS     | DF     | MS     | F
-------|--------|--------|--------|----
Between| $n\displaystyle\sum_{i=1}^{k}(M_i-G)^2$ | $K-1$ | $\frac{SS_B}{df_B}$ | $\frac{MS_B}{MS_W}$
Within | $\displaystyle \sum_{i=1}^{K}\sum_{j=1}^{n}(X_{i_{j}}-M_i)^2$| $N-K$ | $\frac{SS_W}{df_W}$ | 
Total  | $\displaystyle\sum_{j=1}^{N}(X_j-G)^2$  | $N-1$ 

With the formulas (conceptually simplified)

Source | SS     | DF     | MS     | F
-------|--------|--------|--------|----
Between| $nSS_{treatment}$ | $K-1$ | $\frac{SS_B}{df_B}$ | $\frac{MS_B}{MS_W}$
Within | $\displaystyle \sum SS_{within}$| $N-K$ | $\frac{SS_W}{df_W}$ | 
Total  | $SS_{scores}$   | $N-1$ 

Note: $SS_B + SS_W = SS_T$ & $df_B + df_W = df_T$


### Two-Way ANOVA Review (Simplified for Equal Sample Sizes)

Source                | SS         | DF         | MS     | F
----------------------|------------|------------|--------|----
Between               | $SS_B$     | $df_B$     | $MS_B$ | F
$Factor\,A_{Row}$            | $SS_r$     | $df_r$     | $MS_r$ | F
$Factor\,B_{Col}$            | $SS_c$     | $df_c$     | $MS_c$ | F
$Factor\,AxB$          | $SS_{rxc}$ | $df_{rxc}$ | $MS_{rxc}$ | F
Within                | $SS_W$     | $df_W$     | $MS_W$ | 
Total                 | $SS_T$     | $df_T$ 


With the formulas (formal equations)


Source  | SS                            | DF      | MS  | F
--------|-------------------------------|---------|-----|---
Between               | $\tiny n\displaystyle\biggl(\sum_{w=1}^{r}(M_r-G)^2+\sum_{u=1}^{c}(M_c-G)^2\biggr)$ | $rc-1$| $\frac{SS_B}{df_B}$ | $\frac{MS_B}{MS_W}$
$\,A_{(Rows)}$      | $\tiny n_r\displaystyle\sum_{w=1}^{r}(M_w-G)^2$     | $r-1$     | $\frac{SS_r}{df_r}$ | $\frac{MS_r}{MS_W}$
$\,B_{(Col)}$  | $\tiny n_c\displaystyle\sum_{u=1}^{c}(M_u-G)^2$    | $c-1$     | $\frac{SS_c}{df_c}$ | $\frac{MS_c}{MS_W}$
$\,AxB$ | $\tiny SS_{B}-SS_r-SS_c$ | $(r-1)(c-1)$ | $\frac{SS_{rxc}}{df_{rxc}}$ | $\frac{MS_{rxc}}{MS_W}$
Within                | $\tiny \displaystyle \sum_{w=1}^{r}\sum_{j=1}^{n}(X_{w_{j}}-M_w)^2+\displaystyle \sum_{u=1}^{c}\sum_{j=1}^{n}(X_{u_{j}}-M_u)^2$     | $nrc-rc$     | $\frac{SS_w}{df_w}$ | 
Total                 | $\tiny \displaystyle\sum_{j=1}^{N}(X_j-G)^2$      | $nrc-1$ 


With the formulas (conceptually simplified)

Source  | SS                            | DF      | MS  | F
--------|-------------------------------|---------|-----|---
Between | $nSS_{all\,cell\,means}$ | $rc-1$| $\frac{SS_B}{df_B}$ | $\frac{MS_B}{MS_W}$
$\,A_{(Rows)}$ | $n_rSS_{row\,means}$     | $r-1$     | $\frac{SS_r}{df_r}$ | $\frac{MS_r}{MS_W}$
$\,B_{(Cols)}$  | $n_cSS_{col\,means}$    | $c-1$     | $\frac{SS_c}{df_c}$ | $\frac{MS_c}{MS_W}$
$\,AxB$ | $\tiny SS_{B}-SS_r-SS_c$ | $(r-1)(c-1)$ | $\frac{SS_{rxc}}{df_{rxc}}$ | $\frac{MS_{rxc}}{MS_W}$
Within  | $\displaystyle \sum SS_{within}$     | $nrc-rc$     | $\frac{SS_w}{df_w}$ | 
Total   | $SS_{scores}$      | $nrc-1$ 


Note for SS: $SS_B + SS_W = SS_T$ and $SS_B = SS_r + SS_c+ SS_{rxc}$.

Note for df: $df_B + df_W = df_T$ and $df_B = df_r + df_c+ df_{rxc}$.




### Effect Size
#### Eta-Squared
As with a one-way ANOVA, we will use eta-squared, the proportion of variance accounted for by the treatment. Remember that $SS_B$ is the variation caused by the treatment, while $SS_T$ is the variation due to treatment plus noise.

$$\eta_B^2 = \frac{SS_B}{SS_B+SS_W} = \frac{SS_B}{SS_T}$$  

$$\eta_r^2 = \frac{SS_r}{SS_T}$$  

$$\eta_c^2 = \frac{SS_c}{SS_T}$$  


$$\eta_{rxc}^2 = \frac{SS_{rxc}}{SS_T}$$  

#### Partial Eta-Squared

Eta-squared is the proportion of the total variance accounted for by an effect. The nice thing is that eta-squared values can add to 100%! However, each value depends on the number and magnitude of the other effects.

People often prefer to report partial eta-squared, $\eta_p^2$, which is the ratio of effect variance to effect plus error variance. This "controls" for the other effects and reports the effect relative to its noise rather than the total variance. Partial eta-squared is a ratio, while eta-squared is a proportion of the total. Thus, partial eta-squared values do not add to 100%.

$$\eta_{p_{r}}^2 = \frac{SS_r}{SS_{r}+SS_w}$$  
  
$$\eta_{p_{c}}^2 = \frac{SS_c}{SS_{c}+SS_w}$$  

$$\eta_{p_{rxc}}^2 = \frac{SS_{rxc}}{SS_{rxc}+SS_w}$$  

For a two-way ANOVA with manipulated between-subjects factors, this partial eta-squared formula equals the more modern generalized eta-squared, $\eta^2_g$, that R automatically generates. When a factor is measured rather than manipulated, such as gender, the formulas change slightly. We will discuss that next week along with $\omega^2_g$ for bias correction.


## Analysis in R

### Build Our Data Frame (or Read a CSV)

I build the data frame by hand in long format.


```{r}
n=5; r=2; c=3;
Example.Data<-data.frame(ID=seq(1:(n*r*c)),
          Level=c(rep("Low",(n*c)),rep("High",(n*c))),
          Type=rep(c(rep("Watching",n),rep("Reading",n),rep("Thinking",n)),2),
          Perceived.Ability = c(1,3,3,4,2,3,2,3,1,4,3,3,2,3,1,
                                6,4,7,5,5,5,3,2,3,2,3,2,1,4,4))

# Also, I want to re-set the order using the `factor` command and set the order I want. 
Example.Data$Level<-factor(Example.Data$Level, 
                              levels = c("Low","High"),
                              labels = c("Low","High"))

Example.Data$Type<-factor(Example.Data$Type, 
                              levels = c("Watching","Reading","Thinking"),
                              labels = c("Watching","Reading","Thinking"))

```


### Check the Means and SS Calculations with dplyr

```{r}
library(dplyr)
Means.Table<-Example.Data %>%
  group_by(Level,Type) %>%
  summarise(N=n(),
            Means=mean(Perceived.Ability),
            SS=sum((Perceived.Ability-Means)^2),
            SD=sd(Perceived.Ability),
            SEM=SD/N^.5)
knitr::kable(Means.Table, digits = 2) 
# Note remember knitr::kable makes it pretty, but you can just call `Means.Table`
```

### Plot Study

Normally, we plot bar graphs with one SEM unit. I use the `Means.Table` object, which contains the means. We will apply some fancy code to make our plot more APA!

```{r, fig.height=4, fig.width=5}
library(ggplot2)

Plot.1<-ggplot(Means.Table, aes(x = Type, y = Means, group=Level))+
  geom_col(aes(group=Level, fill=Level), position=position_dodge(width=0.9))+
  scale_y_continuous(expand = c(0, 0))+ # Forces plot to start at zero
  geom_errorbar(aes(ymax = Means + SEM, ymin= Means - SEM), 
                position=position_dodge(width=0.9), width=0.25)+
  scale_fill_manual(values=c("#f44141","#4286f4"))+
  xlab('')+
  ylab('Perceived Ability')+
  theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        legend.title=element_blank())
Plot.1
```

Which effects do you predict will be significant? 

### Run ANOVA
Again, we will use the `afex` package, which was developed to run SPSS-like Type III ANOVAs in R.

#### ANOVA-Style Output
- This style is close to our ANOVA source table, but it is formatted like a regression *F*-source table.
- You can ignore the intercept line (we will cover that next semester). 
- Level = $SS_r$: Main effect of Level of Exposure (note R will give the order you called them. I put rows first in the formula below). 
- Type  = $SS_c$: Main effect of Type of Exposure 
- Level:Type  = $SS_{rxc}$: Interaction between Level and Type of Exposure 
- Residual = $SS_{w}$: Residual in regression means leftover. This is the error term: chance.
- There is no total (you can do that with addition)

```{r}
library(afex)

ANOVA.JA.Table<-aov_car(Perceived.Ability~Level*Type + Error(ID), 
                  data=Example.Data, return='Anova')
ANOVA.JA.Table
```

#### Nice-Style Output
- This style reports everything you need in APA format. Leave the return argument off or write `nice`.

```{r}
library(afex)

Nice.JA.Table<-aov_car(Perceived.Ability~Level*Type + Error(ID), 
                  data=Example.Data)
Nice.JA.Table
```

### Understanding $\eta^2$ 
$\eta^2$ is useful for understanding how your ANOVA "parsed" the variance, although it is not always the most useful effect size to report in papers.

Here we see that **Type**, **Level**, and **Type x Level** are "explained variance." Together, they explain about 50% of the variance. The error term is "unexplained variance," or the residual. Do not confuse $\eta^2$, shown in the chart below, with $\eta_p^2$.

```{r, echo = FALSE}
SS_level=ANOVA.JA.Table$`Sum Sq`[2]
SS_Type=ANOVA.JA.Table$`Sum Sq`[3]
SS_inter=ANOVA.JA.Table$`Sum Sq`[4]
SS_w=ANOVA.JA.Table$`Sum Sq`[5]
SS_T=SS_level+SS_Type+SS_inter+SS_w

n2_a=SS_level/(SS_T)
n2_b=SS_Type/(SS_T)
n2_axb=SS_inter/(SS_T)
n2_w=SS_w/SS_T

slices <- c(n2_a,n2_b,n2_axb,n2_w) 
lbls <- c("Level", "Type","Level X Type","Error")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percent to labels 
lbls <- paste(lbls,"%",sep="") # ad % to labels 
pie(slices,labels = lbls, col=rainbow(length(lbls)),
  	main="Pie Chart of Parsed Variance: \nEta-Sq")
```

## Assumptions 
We assume HOV and normality. We can test HOV using Levene's test, which is similar to the Brown-Forsythe test used for a one-way ANOVA but can be applied across multiple factors.

Using the `car` package, we can test for violations. As with Brown-Forsythe, you do NOT want this test to be significant.

```{r, fig.height=4, fig.width=7.5}
library(car)
leveneTest(Perceived.Ability ~ Level*Type, data = Example.Data) 
```

We often examine Q-Q plots of the DV. If the data points deviate substantially from the line, that suggests a violation of normality.

```{r, fig.height=3, fig.width=3}
qqnorm(Example.Data$Perceived.Ability)
qqline(Example.Data$Perceived.Ability)
```

The Shapiro-Wilk test evaluates normality statistically, but it can be very sensitive, especially in small samples. You do NOT want this test to be significant. In our case it is, but our sample is very small.

```{r}
shapiro.test(Example.Data$Perceived.Ability)
```


## References
Kardas, M., & O'Brien, E. (2018). Easier seen than done: Merely watching others perform can foster an illusion of skill acquisition. *Psychological Science*, 29(4), 521-536.
