---
title: "Mixed Regression Over Time: More Than Two Occasions"
bibliography: references.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5.5,
  fig.height = 3.6,
  fig.align = "center",
  fig.pos = "H",
  results = "hold"
)
```

## What a Third Measurement Buys

The [mixed regression chapter](../Ch_18_Mixed_Regression.qmd) stopped, deliberately, at two occasions. Baseline and Week 6, one within-subjects factor, one between-subjects factor, four cells, a random intercept. It ended by admitting what that design could not do, and this chapter is the part it could not do.

Three things were out of reach with two measurements:

- **There was no such thing as *the* change.** With baseline and Week 6 only, one subtraction says everything. Add Weeks 2 and 4 and you have to decide what shape change has.
- **A random slope was not estimable.** Two points determine a line exactly, so `lmer` refused to separate "this person improved quickly" from "this measurement was noisy."
- **The residual plot showed the damage.** Individual differences in rate of change were in the simulated data, could not be modelled, and turned up as a trend in the residuals instead.

All three loosen at three occasions and disappear at four. We will use four.

## The Design

Same study, more visits. Researchers compare cognitive behavioral therapy (**CBT**) with a control condition for reducing depression symptoms. Participants are randomly assigned, then complete a depression inventory at baseline and after 2, 4, and 6 weeks.

- **Group** is between subjects: each person is in CBT *or* control.
- **Week** is within subjects: each person is measured four times.
- **Depression score** is the quantitative outcome.

The question is no longer whether one group changed more than the other between two visits. It is about the shape of the whole trajectory:

> Does the slope of depression over time depend on treatment group?

That is still a Week × Group interaction, but the interaction now compares two **slopes** rather than two changes.

This is simulated data. Real treatment studies require clinical expertise, ethical oversight, careful monitoring, informed consent, and substantially more planning than "Alex found `rnorm()`."

```{r simulate-mixed-data}
set.seed(3436)

n_per_group <- 50
people <- data.frame(
  participant = factor(seq_len(2 * n_per_group)),
  group = factor(
    rep(c("Control", "CBT"), each = n_per_group),
    levels = c("Control", "CBT")
  ),
  person_intercept = rnorm(2 * n_per_group, 0, 5.0),
  person_slope = rnorm(2 * n_per_group, 0, 0.45)
)

# four visits per person instead of two
mixed_data <- merge(
  people,
  data.frame(week = c(0, 2, 4, 6)),
  by = NULL
)

mixed_data <- mixed_data[order(mixed_data$participant,
                               mixed_data$week), ]

mixed_data$depression <-
  35 +
  mixed_data$person_intercept +
  (-0.20 * mixed_data$week) +
  ifelse(mixed_data$group == "CBT",
         -1.20 * mixed_data$week, 0) +
  mixed_data$person_slope * mixed_data$week +
  rnorm(nrow(mixed_data), 0, 2.0)

mixed_data$occasion <- factor(
  mixed_data$week,
  levels = c(0, 2, 4, 6),
  labels = c("Baseline", "Week 2", "Week 4", "Week 6")
)
```

The data are **long**: one row per person per occasion, 400 rows from 100 people. The participant identifier repeats because the person repeats. If your participant column disappears, stop. Do not proceed directly to the ritual sacrifice.

That `person_slope` column is the one the two-occasion chapter could see the effects of but could not estimate. This time we can.

## First Look: Means and People

```{r mixed-design-plot}
#| fig-cap: "Every faint line is one participant measured four times. The two heavy lines are the group means at each week. The whole chapter is about whether those two heavy lines have different slopes."
#| fig-alt: "A line plot with Week on the x axis, at 0, 2, 4 and 6, and depression score on the y axis running from about 8 to about 54. One hundred faint lines, one per participant, fan out across that whole range, colored salmon for Control and teal for CBT. Two heavy lines with round markers at each week show the group means. The Control mean drifts down gently from about 36 to about 34. The CBT mean falls steeply from about 34 to about 26. The two heavy lines start close together and end far apart."
library(ggplot2)

ggplot(
  mixed_data,
  aes(x = week, y = depression, group = participant, color = group)
) +
  geom_line(alpha = 0.13) +
  stat_summary(
    aes(group = group), fun = mean, geom = "line",
    linewidth = 1.4
  ) +
  stat_summary(
    aes(group = group), fun = mean, geom = "point",
    size = 2.8
  ) +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  labs(
    x = "Week",
    y = "Depression score",
    color = "Group",
    title = "Group averages sitting on top of actual humans"
  ) +
  theme_minimal(base_size = 11)
```

The bold lines show group means. The faint lines show individual trajectories. The CBT mean declines faster, but people do not all begin at the same score or change at the same rate. Some improve quickly, some slowly, and some appear to have received therapy from a motivational poster in a dentist's office.

Never decide whether an effect is statistically significant by staring at whether confidence intervals overlap. CI overlap is not the direct test of a difference, and interaction tests are about **differences between differences**. Fit the model that asks the question.

## Why Ordinary Regression Is Not Enough

An ordinary regression such as this is incomplete:

```{r ordinary-model-warning}
ordinary_model <- lm(depression ~ week * group, data = mixed_data)
```

It treats all 400 rows as independent. They are not. Four rows belong to each participant, and observations from the same person tend to resemble one another.

This structure is called **nesting**:

$$
\text{occasions nested within participants}
$$

The levels are:

- Level 1: repeated occasions;
- Level 2: participants.

Group belongs to Level 2 because it does not change within a person. Week belongs to Level 1 because it does.

## The Fixed Effects

We begin with the part that resembles ordinary regression:

$$
Y_{ij}=b_0+b_1(\text{Week}_{ij})+b_2(\text{CBT}_i)
+b_3(\text{Week}_{ij}\times\text{CBT}_i)+\text{random part}+e_{ij}
$$

Because Week 0 is baseline and Control is the reference group:

- $b_0$: estimated baseline score for Control;
- $b_1$: expected change per week for Control;
- $b_2$: CBT-minus-Control difference at baseline;
- $b_3$: difference between the CBT and Control slopes.

The CBT slope is $b_1+b_3$. The interaction $b_3$ is the main research target. If it is negative, CBT scores decline faster than Control scores.

Note what changed from the two-occasion version. There, $b_1$ was a *change*, one number for the whole six weeks. Here it is a *rate*, points per week, and the interaction is a difference between rates. The research question did not change. Its units did.

Notice also why coding matters. A "main effect of Group" is not a universal average group difference here. It is the group difference when Week equals zero. We chose zero to mean baseline, so the coefficient has a useful interpretation instead of describing the difference during Week Zero Hundred and Twelve, a period known only to the Probability Gods.

## The Random Effects

The repeated observations require us to describe how people differ.

### Random Intercepts

A random-intercept model allows each participant to have their own starting level:

$$
Y_{ij}=b_0+b_1\text{Week}_{ij}+b_2\text{CBT}_i
+b_3(\text{Week}_{ij}\times\text{CBT}_i)+u_{0i}+e_{ij}
$$

```{r random-intercept-model}
library(lme4)      # fits the mixed model
library(lmerTest)  # adds degrees of freedom and p-values to the fixed effects

random_intercept_model <- lmer(
  depression ~ week * group + (1 | participant),
  data = mixed_data
)

summary(random_intercept_model)
```

This is the same `(1 | participant)` that reproduced the paired *t*-test in the previous chapter, doing the same job on a larger design: give participants different intercepts, by estimating the variance of those intercepts rather than 100 unrelated participant coefficients.

`lmer()` fits with REML by default, which is what we want while we are looking at the random effects. The [previous chapter](../Ch_18_Mixed_Regression.qmd) has the deliberately oversimplified version of what REML is doing and why. The practical rule is unchanged: REML while we are estimating the random-effects structure, REML switched off when we compare models that differ in their fixed effects.

<!-- ---Alex--- Pointer rather than a copy, so your REML callout lives in exactly one place.
If this chapter ends up NOT following Mixed_Regression in the running order, paste the callout
in here instead, because the forward promise it makes is paid off in this chapter too. -->


### Random Slopes

Now the thing two occasions could not do. People may change at different rates, and a random-slope model gives each person their own Week slope:

```{r random-slope-model}
random_slope_model <- lmer(
  depression ~ week * group + (1 + week | participant),
  data = mixed_data
)

summary(random_slope_model)
```

The expression `(1 + week | participant)` estimates:

- variance in participants' intercepts;
- variance in participants' Week slopes;
- correlation between participant intercepts and slopes.

That third quantity is worth a moment. It asks whether the people who start worst are also the people who improve fastest, which is a substantive question that the model answers as a side effect of being specified honestly.

Random slopes are often substantively reasonable in longitudinal psychology. People do not all change identically merely because the spreadsheet has aligned their rows. They are humans, not synchronized microwave clocks.

The random-effects structure should reflect the design and theory, while remaining supported by the data. If a model is singular or refuses to converge, do not repeatedly click Run while shouting encouragement. Inspect the data, scaling, model specification, number of clusters, and whether the random structure is estimable.

## Read the Model Without Panic

```{r fixed-effects-table}
fixed_effects <- summary(random_slope_model)$coefficients
round(fixed_effects, 3)
```

Read the rows in order:

1. `(Intercept)` estimates the Control group's baseline mean.
2. `week` estimates weekly change in Control.
3. `groupCBT` estimates the CBT-minus-Control baseline difference.
4. `week:groupCBT` estimates how much the CBT slope differs from the Control slope.

The interaction estimate is `r round(fixed_effects["week:groupCBT", "Estimate"], 2)`. That means the CBT group's depression score declines about `r abs(round(fixed_effects["week:groupCBT", "Estimate"], 2))` points **more per week** than the Control group's score, on average. It does not mean CBT lowers every participant's score by exactly that much each week. Fixed effects describe an average pattern while the random effects document the inconvenient existence of individuals.

::: {.callout-warning title="The p-Values Are Approximations, and This Is Not Your Engine's Fault"}
Every column in that table except the estimate itself rests on an approximation. There is no exact sampling distribution waiting for a model like this one, so the numbers in the `df` column were not counted, they were **estimated**, using Satterthwaite's method. Here they land near `r round(fixed_effects["week:groupCBT", "df"], 0)`, and in the categorical-time model below they come out frankly fractional.

This is a property of mixed models, not of `lmer`. Change engines and the approximation changes costume rather than disappearing: `glmmTMB` fits the same model and reports Wald *z* statistics with no degrees of freedom at all. Same model, different plumbing, and the same honest uncertainty underneath.

So say which method you used, look at the confidence intervals, and do not treat a *p*-value from a mixed model as a measurement taken with calipers. "R produced a number and I felt hopeful" is not an inferential method.
:::

## Test the Interaction Directly

To test whether the interaction improves the model, compare two models with the same random-effects structure:

- a reduced model without Week × Group;
- a full model with Week × Group.

When comparing **fixed effects**, refit both models with maximum likelihood. This is the moment `lmer()`'s REML default has to be switched off by hand, in both models, or the comparison is not a legal one.

```{r compare-fixed-effects}
reduced_ml <- lmer(
  depression ~ week + group + (1 + week | participant),
  data = mixed_data,
  REML = FALSE
)

full_ml <- lmer(
  depression ~ week * group + (1 + week | participant),
  data = mixed_data,
  REML = FALSE
)

anova(reduced_ml, full_ml)
```

This likelihood-ratio test asks whether allowing the groups to have different slopes improves model fit. It is not the only defensible test; coefficient tests and other approximations are also used. The important point is to match the test to the question and identify the method, not to collect p values like cursed trading cards.

After model comparison, report estimates from the final model fit with REML when appropriate.

## Follow Up the Interaction

The interaction says the slopes differ. We still want the slope within each group.

```{r follow-up-slopes}
library(emmeans)

group_slopes <- emtrends(
  random_slope_model,
  specs = ~ group,
  var = "week"
)

group_slopes
pairs(group_slopes)
```

`emtrends()` estimates the Week slope separately for Control and CBT. `pairs()` tests their difference, which corresponds to the interaction. This is the three-or-more-occasions replacement for the simple effects the two-occasion chapter used: there, each group's change was a single contrast between two cells, and here it is a slope.

We can also estimate model-implied group means at the observed weeks:

```{r estimated-means}
estimated_means <- emmeans(
  random_slope_model,
  specs = ~ group | week,
  at = list(week = c(0, 2, 4, 6))
)

estimated_means
```

Do not automatically test every possible pair at every possible week because the software makes it easy. That is how a sensible research question becomes 28 tests and a dead [salmon](../Ch_20_Multiple_Comparisons.qmd) [@bennett2010]. Choose comparisons that answer the theory, and adjust for multiplicity when you conduct a family of tests.

## Reporting the Result

A compact results paragraph could read:

> Depression scores were analyzed with a linear mixed-effects model containing fixed effects of Week, Group, and their interaction, plus correlated participant random intercepts and Week slopes. The model was fit by restricted maximum likelihood using `lmer()` from the lme4 package, and *p*-values were obtained from lmerTest using Satterthwaite's approximation to the denominator degrees of freedom. The Week × Group interaction was negative, $b =$ `r round(fixed_effects["week:groupCBT", "Estimate"], 2)`, $SE =$ `r round(fixed_effects["week:groupCBT", "Std. Error"], 2)`, $t(`r round(fixed_effects["week:groupCBT", "df"], 1)`) =$ `r round(fixed_effects["week:groupCBT", "t value"], 2)`, $p$ `r ifelse(fixed_effects["week:groupCBT", "Pr(>|t|)"] < .001, "< .001", paste("=", round(fixed_effects["week:groupCBT", "Pr(>|t|)"], 3)))`, indicating that depression declined more rapidly in CBT than in the control condition.

A real report should also provide confidence intervals, descriptive/model-implied means, the exact random-effects structure, software, missing-data handling, and theoretically planned follow-ups. The paragraph above is the skeleton. Please give it organs before submitting it to Reviewer 2.

# Advanced Topics {.unnumbered}

## Should Time Be Continuous or Categorical?

Treating Week as numeric assumes a straight-line average trajectory. That is a strong but efficient summary: one slope per group.

If change may be nonlinear, perhaps because treatment helps rapidly and then levels off, treat occasion as categorical:

```{r categorical-time-model}
categorical_time_model <- lmer(
  depression ~ occasion * group + (1 + week | participant),
  data = mixed_data,
  control = lmerControl(optimizer = "bobyqa")
)

round(summary(categorical_time_model)$coefficients, 3)
```

::: {.callout-warning title="Why That `control =` Argument Is There"}
It is not decoration. Fit this model with `lmer()`'s default optimizer and it reports that it failed to converge, by a margin small enough to be irritating rather than informative.

This is the situation the random-effects section warned about, and `bobyqa` is the ordinary first thing to try: a different optimizer hunting for the same maximum. It finds it, and the estimates agree with the default fit to three decimals, which is how you know the complaint was about the search and not about the model.

What you may not do is make the warning go away by not looking at it.
:::

Categorical time estimates separate deviations at Weeks 2, 4, and 6. It is flexible but uses more parameters. Numeric time is compact but imposes shape. Neither coding is morally superior. The choice depends on the research question, plausible change process, timing, and amount of data.

Other options include polynomial terms, splines, or nonlinear models. Do not add a seventh-degree polynomial merely because R accepts it. R will also accept many bad ideas with the serene indifference of a vending machine.

## Missing Repeated Observations

Mixed models can use participants who are missing some occasions; they do not require deleting every person with one missing row. This is a major advantage over older complete-case repeated-measures procedures, and it grows more valuable as occasions accumulate, because with four visits there are four chances for somebody to miss one.

However, "the model ran" does not mean missing data ceased to matter. Standard likelihood-based mixed-model inference commonly relies on a **missing at random** assumption conditional on variables in the model. If people with worsening symptoms leave the study for reasons the model does not capture, the missingness can bias results. Investigate dropout, include relevant predictors when justified, and conduct sensitivity analyses in serious work.

## Check the Model

At minimum, examine:

- whether the mean trajectory is represented sensibly;
- residual-versus-fitted patterns;
- severe residual outliers;
- the distribution and influence of clusters;
- convergence and singular-fit warnings;
- whether variance changes across time or groups;
- whether enough participants, not merely enough rows, support the model.

```{r mixed-diagnostics, fig.width=6, fig.height=3}
#| fig-cap: "Residuals from the random-slope model plotted against its fitted values. We want a shapeless cloud sitting on the red zero line. Through the middle, that is what we get. The black smoother bends at both ends, where it has almost no data to average."
#| fig-alt: "A scatterplot of residuals against fitted values, with fitted values running from about 10 to 51 and residuals from about -6 to +4. The points form a wide, even band centered on zero, densest between fitted values of 25 and 40, with no funnel shape. A dashed red horizontal line marks zero. A black smoothed line lies on zero through the dense middle, but dips to about -1.3 at the far left and rises to about +0.6 at the far right, where only a handful of points exist."
diagnostic_data <- data.frame(
  fitted = fitted(random_slope_model),
  residual = resid(random_slope_model)
)

ggplot(diagnostic_data, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.35, color = "steelblue4") +
  geom_hline(yintercept = 0, linetype = 2, color = "firebrick") +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  labs(x = "Fitted value", y = "Residual") +
  theme_minimal(base_size = 11)
```

The band is even across the range and centered on zero, with no funnel and no arc, which is what we were hoping for. The smoother's droop at the far left is worth naming rather than worrying about: below a fitted value of 15 there are exactly two observations, and the black line is drawing a conclusion from them. A loess line will always find something to say at the edges of your data, whether or not there is anything there.

Compare that with the same plot in the two-occasion chapter, which had a real trend in it. The trend was the individual differences in rate of change, with nowhere to go. Here they have somewhere to go, so they left the residuals.

Traditional repeated-measures ANOVA emphasizes **sphericity**. With four occasions there are several pairwise differences, so the assumption has teeth again in a way it did not with two. A mixed model does not require that single sphericity assumption, but it does require a defensible covariance and random-effects structure. We did not escape assumptions. We traded one rigid assumption for a model that must be specified thoughtfully.

## The Short Story

- With three or more occasions there is no single change score, so change gets described by a slope or by one estimate per occasion.
- Continuous time turns the Week × Group interaction into a difference between group **slopes**. The research question is the same; the units are not.
- Random intercepts let people start at different levels. Random slopes let them change at different rates, and they need at least three occasions to be estimable at all.
- The intercept-slope correlation is a free substantive result: it says whether the people who start worst improve fastest.
- Compare fixed-effects structures using models fit with maximum likelihood, not REML.
- Continuous time tests a compact trajectory; categorical time permits separate occasion patterns and spends more parameters doing it.
- A convergence warning is about the search, not automatically about the model. Try another optimizer before you try another model, and never fix it by not looking.
- Mixed models can retain partially observed participants, but missingness can still bias the answer.
- Never count rows and forget to count people. Four hundred observations from 100 people are not 400 independent humans, no matter how urgently the grant deadline approaches.

The model is "mixed" because it combines fixed and random effects, not because we placed several statistical procedures into a blender and hoped the lid stayed on.
